---
title: "Ecoplots_Stem_Height_vs_Stem_Diameter"
author: "Elisa"
date: "2024-08-19"
output: word_document
---

# Set your working directory
```{r}
rm(list=ls()) # cleans the workspace
setwd("C:/Users/uqegirol") # Note that your path will be different
```

# Load libraries
```{r}
library(tidyverse)
library(httr)
library(ggpubr)
library(gridExtra)
```

# Load and explore the data
```{r echo = FALSE}
library(httr)

headers = c(
  'X-Api-Key' = 'd3h3UDhZZjJqZ3dUN3dQOS5pcUgxJj1FWU4nMH04JgtBCU54KChkC1ZcWCJYNzFfOmZWMThSdAsuY1NfMEhoXkhZKW10bSh+bwk0MyJac1Rz',
  'Content-Type' = 'application/json'
)

body = '{
  "query": {
    "feature_type": [
      "http://linked.data.gov.au/def/tern-cv/60d7edf8-98c6-43e9-841c-e176c334d270"
    ],
    "observed_property": [
      "http://linked.data.gov.au/def/tern-cv/70646576-6dc7-4bc5-a9d8-c4c366850df0",
      "http://linked.data.gov.au/def/tern-cv/c3111898-8404-4306-9599-e7866a749de7",
      "http://linked.data.gov.au/def/tern-cv/5701d307-bbea-415a-8857-fbf27d5b93ed"
    ]
  }
}';

res <- VERB("POST", url = "https://ecoplots.tern.org.au/api/v1.0/data?dformat=csv", body = body, add_headers(headers))

cat(content(res, 'text'))

df <- read.table(text = content(res, 'text'), sep =",", header = TRUE, stringsAsFactors = FALSE)
```

## Restructure the data
```{r}
str(df)
df$datasetTitle <- as.factor(df$datasetTitle)
df$scientificName <- as.factor(df$scientificName)
df$stemHeight_metre <- as.numeric(df$stemHeight_metre)
df$stemDiameter_centimetre <- as.numeric(df$stemDiameter_centimetre)
```

## Plot the data
```{r}
# Plot all the data
plot(df$stemHeight_metre, df$stemDiameter_centimetre) 
```

# Fit the linear regression model
model1 <- glm(stemDiameter_centimetre ~ stemHeight_metre, data = df)
summary(model1)
```

# Select Eucalyptus trees
THIS NEEDS WORK AFTER WE SORT OUT THE ISSUES RE SCIENTIFIC NAME AND ACCEPTED NAME. ISSUES ARE:
1) WHAT WE NOW CALL SCIENTIFICNAME IS ACTUALLY THE NAME PROVIDED BY THE DATA PROVIDER. THAT IS USEFUL INFORMATION, BUT NOT WHAT WE WANT FOR THE ANALYSES AS THESE ARE NOT STANDARDISED AND THERE ARE REPETITIONS. 
2) WE WANT UNIQUE ACCEPTED NAMES, PLUS THE FAMILY NAME. IN ECOPLOTS THE FIELD TAXON CONTAINS THE FULL TAXONOMY FOR THAT SPECIES. HOWEVER, WHEN WE EXPORT THE DATA, INSTEAD OF THE TAXON AS A CHARACTER STRING, WE GET AN URL.
3) FOR THIS SPECIFIC EXAMPLE WE NEED TO REMOVE THE TREES THAT WERE NOT IDENTIFIED AT A SPECIES LEVEL. THIS WILL BE EASIER ONCE WE RESOLVE ISSUES 1 AND 2. 

Issues
```{r}
#levels(df_Eucalyptus$scientificName)
#levels(df_Eucalyptus$taxon)
```

```{r}
df_Eucalyptus<-subset(df, grepl("Eucalyptus", scientificName))
df_Eucalyptus$scientificName <- droplevels(df_Eucalyptus$scientificName)
plot(df_Eucalyptus$stemHeight_metre, df_Eucalyptus$stemDiameter_centimetre) 

model2 <- glm(stemDiameter_centimetre ~ stemHeight_metre, data = df_Eucalyptus)
summary(model2)

level_counts <- table(df_Eucalyptus$scientificName) # Count the occurrences of each level
print(level_counts)

# Drop levels with less than 1000 entries
levels_to_keep <- names(level_counts[level_counts >= 1000]) # Identify levels with a sample size smaller than the threshold
df_Eucalyptus_filtered <- df_Eucalyptus[df_Eucalyptus$scientificName %in% levels_to_keep, ] # Subset the data to keep only rows with the desired levels
df_Eucalyptus_filtered$scientificName <- droplevels(df_Eucalyptus_filtered$scientificName) # Drop unused levels
levels(df_Eucalyptus_filtered$scientificName) # Check the resulting levels

model3 <- lm(stemDiameter_centimetre ~ stemHeight_metre + scientificName, df_Eucalyptus_filtered)
summary(model3)

```
# Plotting each species in different colors
```{r}
ggplot(df_Eucalyptus_filtered, aes(x = stemHeight_metre, y = stemDiameter_centimetre, color = scientificName)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +  # Adds regression lines
  theme_minimal() +
  labs(title = "Regression Lines for Different Species")

```






