---
title: "Ecoplots_Stem_Height_vs_Stem_Diameter"
author: "Elisa"
date: "2024-08-19"
output: word_document
---

# Set your working directory
```{r}
rm(list=ls()) # cleans the workspace
setwd("C:/Users/uqegirol") # Note that your path will be different
```

# Load libraries
```{r}
library(tidyverse)
library(httr)
library(ggpubr)
library(gridExtra)
```

# Load and explore the data

## WILL NEED TO CHANGE THIS TO SELECT ONLY TREES WITHIN QLD

```{r echo = FALSE}
library(httr)

headers = c(
  'X-Api-Key' = 'd3h3UDhZZjJqZ3dUN3dQOS5pcUgxJj1FWU4nMH04JgtBCU54KChkC1ZcWCJYNzFfOmZWMThSdAsuY1NfMEhoXkhZKW10bSh+bwk0MyJac1Rz',
  'Content-Type' = 'application/json'
)

body = '{
  "query": {
    "feature_type": [
      "http://linked.data.gov.au/def/tern-cv/60d7edf8-98c6-43e9-841c-e176c334d270"
    ],
    "observed_property": [
      "http://linked.data.gov.au/def/tern-cv/70646576-6dc7-4bc5-a9d8-c4c366850df0",
      "http://linked.data.gov.au/def/tern-cv/c3111898-8404-4306-9599-e7866a749de7",
      "http://linked.data.gov.au/def/tern-cv/5701d307-bbea-415a-8857-fbf27d5b93ed"
    ]
  }
}';

res <- VERB("POST", url = "https://ecoplots.tern.org.au/api/v1.0/data?dformat=csv", body = body, add_headers(headers))

cat(content(res, 'text'))

df <- read.table(text = content(res, 'text'), sep =",", header = TRUE, stringsAsFactors = FALSE)
```

## Restructure the data
```{r}
str(df)
df$datasetTitle <- as.factor(df$datasetTitle)
df$scientificName <- as.factor(df$scientificName)
df$stemHeight_metre <- as.numeric(df$stemHeight_metre)
df$stemDiameter_centimetre <- as.numeric(df$stemDiameter_centimetre)
```

## DEAL WITH REPEATED MEASUREMENTS

## Plot the data
```{r}
# Plot all the data
plot(df$stemHeight_metre, df$stemDiameter_centimetre) 
```

## Fit the linear regression model
```{r}
model1 <- glm(stemDiameter_centimetre ~ stemHeight_metre, data = df)
summary(model1)
```

# Select Eucalyptus trees
THIS NEEDS WORK AFTER WE SORT OUT THE ISSUES RE SCIENTIFIC NAME AND ACCEPTED NAME. ISSUES ARE:
1) WHAT WE NOW CALL SCIENTIFICNAME IS ACTUALLY THE NAME PROVIDED BY THE DATA PROVIDER. THAT IS USEFUL INFORMATION, BUT NOT WHAT WE WANT FOR THE ANALYSES AS THESE ARE NOT STANDARDISED AND THERE ARE REPETITIONS. 
2) WE WANT UNIQUE ACCEPTED NAMES, PLUS THE FAMILY NAME. IN ECOPLOTS THE FIELD TAXON CONTAINS THE FULL TAXONOMY FOR THAT SPECIES. HOWEVER, WHEN WE EXPORT THE DATA, INSTEAD OF THE TAXON AS A CHARACTER STRING, WE GET AN URL.
3) FOR THIS SPECIFIC EXAMPLE WE NEED TO REMOVE THE TREES THAT WERE NOT IDENTIFIED AT A SPECIES LEVEL. THIS WILL BE EASIER ONCE WE RESOLVE ISSUES 1 AND 2. 

```{r}
df_Eucalyptus<-subset(df, grepl("Eucalyptus", scientificName))
df_Eucalyptus$scientificName <- droplevels(df_Eucalyptus$scientificName)
plot(df_Eucalyptus$stemHeight_metre, df_Eucalyptus$stemDiameter_centimetre) 
```

## Fit the model only for Eucalyptus trees
```{r}
model2 <- glm(stemDiameter_centimetre ~ stemHeight_metre, data = df_Eucalyptus)
summary(model2)
```

## Issues
```{r}
#levels(df_Eucalyptus$scientificName)
#levels(df_Eucalyptus$taxon)
```

# Drop species with a small sample size
```{r}
level_counts <- table(df_Eucalyptus$scientificName) # Count the occurrences of each level
print(level_counts)
levels_to_keep <- names(level_counts[level_counts >= 1000]) # Identify levels with a sample size above the threshold
df_Eucalyptus_filtered <- df_Eucalyptus[df_Eucalyptus$scientificName %in% levels_to_keep, ] # Subset the data to keep only rows with the desired levels
df_Eucalyptus_filtered$scientificName <- droplevels(df_Eucalyptus_filtered$scientificName) # Drop unused levels
levels(df_Eucalyptus_filtered$scientificName) # Check the resulting levels
```

## Re-fit the model using only species with a large sample size
```{r}
model3 <- lm(stemDiameter_centimetre ~ stemHeight_metre + scientificName, df_Eucalyptus_filtered)
summary(model3)
```

## Plotting each species in different colors
```{r}
ggplot(df_Eucalyptus_filtered, aes(x = stemHeight_metre, y = stemDiameter_centimetre, color = scientificName)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +  # Adds regression lines
  theme_minimal() +
  labs(title = "Regression Lines for Different Species")
```

# Select only large trees
```{r}
df_LargeTrees<-df[df$stemDiameter_centimetre > 100, ]
```

## Identify the most common species
```{r}
df_LargeTrees$scientificName <- droplevels(df_LargeTrees$scientificName) # Drop unused levels
counts <- table(df_LargeTrees$scientificName) # Create a frequency table of the categorical variable

level_counts <- table(df_LargeTrees$scientificName) # Count the occurrences of each level
levels_to_keep <- names(level_counts[level_counts >= 10]) # Identify levels with a sample size above the threshold
df_LargeTrees <- df_LargeTrees[df_LargeTrees$scientificName %in% levels_to_keep, ] # Subset the data to keep only rows with the


ggplot(df_LargeTrees, aes(x = scientificName)) +
  geom_bar(fill = "skyblue") +
  labs(title = "Bar Plot", x = "scientificName", y = "Frequency")
```

## Plot the location of the trees
NOTE: WE HAVE THE LAT LON OF THE PLOT AND THE XY LOCATION OF THE TREE WITHIN THE PLOT. MAY NEED TO INVESTIGATE BEST WAY TO HANDLE THIS.





# Looking at tree growth (only for trees with repeated measurement)
## Select trees that have repeated measurements
## Calculate the growth rate of the tree
## Model teh growth rate by species and/or ecoregion





