---
title: "Integrating Multi-Source Environmental Data to Monitor Plant Biodiversity Across Australian Rangelands"
author: "Elisa Girola and Eleanor Velasquez"
date: "2025-09-22"
output: word_document
editor_options: 
  chunk_output_type: inline
---

In this tutorial we will bring together different data products to investigate the relationship between plant species richness and meteorological and soil parameters in Australian rangelands. The original data products were collected or generated independently. They have different spatial and temporal covers and resolutions, and are saved in different file formats. In this tutorial we will demonstrate how to process the original data to create a harmonised dataset. This will be used in statistical models to investigate the main drivers of vegetation species richness in Australian rangelands.  


# SET UP
## Clean the workspace
```{r, include=FALSE}
rm(list=ls()) # Clean the workspace

```


## Set up global options
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) # Show the R code in all other chunks of this R-markdown into the knitted document.
```


# ACCESSING THE DATA
For this tutorial you will need vegetation data from field surveys, modelled meteorological data, and soil maps. 

There are two ways of getting the data into R:

1) Download each data file to your local machine. With this method the code will run faster. 

2) Access the data directly from R. This method ensures that you will always access the latest version of the data. However, it will require longer to run the code. 

For this tutorial we will use option 1. You will need to download about 50 files. We recommend downloading all the files to your local machine before starting the tutorial. You will need about 7 GB of disk space for this. For the code to run smoothly it is important that all the files are saved in the same folder, without any subfolders. No other files should be saved in this folder.  


## Vegetation data
Usually you would download vegetation survey data through the EcoPlots dashboard https://ecoplots.tern.org.au/ However, for this tutorial we have pre-selected the data and saved them to a csv file that can be accessed through the following link. This ensures that the file contains only the data that we need for the tutorial keeping the file size small and the results reproducible. 

Note: We have a separate video demonstrating how to navigate to this data & download it using EcoPlots on the workshop web page 

https://tern.short.gy/file-plant-occurrence


## Meteorological data
Meteorological data are in netCDF files and can be downloaded through the following links. Since we are using monthly mean data, for each parameters there are 12 files, i.e. one for each month of the year. 

### Precipitation

https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_rainfall_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_rainfall_monthly-mean_0-01deg_1976-2005_00000000_01.nc 

https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_rainfall_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_rainfall_monthly-mean_0-01deg_1976-2005_00000000_02.nc

https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_rainfall_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_rainfall_monthly-mean_0-01deg_1976-2005_00000000_03.nc

https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_rainfall_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_rainfall_monthly-mean_0-01deg_1976-2005_00000000_04.nc

https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_rainfall_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_rainfall_monthly-mean_0-01deg_1976-2005_00000000_05.nc

https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_rainfall_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_rainfall_monthly-mean_0-01deg_1976-2005_00000000_06.nc

https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_rainfall_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_rainfall_monthly-mean_0-01deg_1976-2005_00000000_07.nc

https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_rainfall_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_rainfall_monthly-mean_0-01deg_1976-2005_00000000_08.nc

https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_rainfall_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_rainfall_monthly-mean_0-01deg_1976-2005_00000000_09.nc

https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_rainfall_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_rainfall_monthly-mean_0-01deg_1976-2005_00000000_10.nc

https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_rainfall_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_rainfall_monthly-mean_0-01deg_1976-2005_00000000_11.nc

https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_rainfall_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_rainfall_monthly-mean_0-01deg_1976-2005_00000000_12.nc


### Monthly mean minimum temperature
https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_temperature-min_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_temperature-min_monthly-mean_0-01deg_1976-2005_00000000_01.nc

https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_temperature-min_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_temperature-min_monthly-mean_0-01deg_1976-2005_00000000_02.nc

https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_temperature-min_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_temperature-min_monthly-mean_0-01deg_1976-2005_00000000_03.nc

https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_temperature-min_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_temperature-min_monthly-mean_0-01deg_1976-2005_00000000_04.nc

https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_temperature-min_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_temperature-min_monthly-mean_0-01deg_1976-2005_00000000_05.nc

https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_temperature-min_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_temperature-min_monthly-mean_0-01deg_1976-2005_00000000_06.nc

https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_temperature-min_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_temperature-min_monthly-mean_0-01deg_1976-2005_00000000_07.nc

https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_temperature-min_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_temperature-min_monthly-mean_0-01deg_1976-2005_00000000_08.nc

https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_temperature-min_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_temperature-min_monthly-mean_0-01deg_1976-2005_00000000_09.nc

https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_temperature-min_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_temperature-min_monthly-mean_0-01deg_1976-2005_00000000_10.nc

https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_temperature-min_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_temperature-min_monthly-mean_0-01deg_1976-2005_00000000_11.nc

https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_temperature-min_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_temperature-min_monthly-mean_0-01deg_1976-2005_00000000_12.nc


### Monthly mean maximum temperature
https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_temperature-max_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_temperature-max_monthly-mean_0-01deg_1976-2005_00000000_01.nc

https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_temperature-max_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_temperature-max_monthly-mean_0-01deg_1976-2005_00000000_02.nc

https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_temperature-max_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_temperature-max_monthly-mean_0-01deg_1976-2005_00000000_03.nc

https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_temperature-max_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_temperature-max_monthly-mean_0-01deg_1976-2005_00000000_04.nc

https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_temperature-max_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_temperature-max_monthly-mean_0-01deg_1976-2005_00000000_05.nc

https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_temperature-max_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_temperature-max_monthly-mean_0-01deg_1976-2005_00000000_06.nc

https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_temperature-max_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_temperature-max_monthly-mean_0-01deg_1976-2005_00000000_07.nc

https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_temperature-max_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_temperature-max_monthly-mean_0-01deg_1976-2005_00000000_08.nc

https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_temperature-max_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_temperature-max_monthly-mean_0-01deg_1976-2005_00000000_09.nc

https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_temperature-max_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_temperature-max_monthly-mean_0-01deg_1976-2005_00000000_10.nc

https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_temperature-max_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_temperature-max_monthly-mean_0-01deg_1976-2005_00000000_11.nc

https://thredds.nci.org.au/thredds/fileServer/rr9/ANUClimate/ANUClimate_v1-0_temperature-max_monthly-mean_0-01deg_1976-2005/00000000/ANUClimate_v1-0_temperature-max_monthly-mean_0-01deg_1976-2005_00000000_12.nc


## Soil maps
Soil maps are .tif files and can be downloaded through the following links. Please note that these are large files and may require 10 to 20 minutes to download per file.
#Note you may need to create a TERN account if not done so already in order to download these: Instructions here: https://youtu.be/HTlc0xk4zf8 

### Available Water Capacity 
https://data.tern.org.au/model-derived/slga/NationalMaps/SoilAndLandscapeGrid/AWC/v2/AWC_000_005_EV_N_P_AU_TRN_N_20210614.tif   

### Clay
https://data.tern.org.au/model-derived/slga/NationalMaps/SoilAndLandscapeGrid/CLY/v2/CLY_000_005_EV_N_P_AU_TRN_N_20210902.tif

### Nitrogen 
https://data.tern.org.au/model-derived/slga/NationalMaps/SoilAndLandscapeGrid/NTO/v2/NTO_000_005_EV_N_P_AU_NAT_C_20231101.tif

### pH 
https://data.tern.org.au/model-derived/slga/NationalMaps/SoilAndLandscapeGrid/pHc/v2/PHC_000_005_EV_N_P_AU_NAT_C_20210913.tif

### Phosphorus
https://data.tern.org.au/model-derived/slga/NationalMaps/SoilAndLandscapeGrid/PTO/v2/PTO_000_005_EV_N_P_AU_NAT_C_20231101.tif

### Silt 
https://data.tern.org.au/model-derived/slga/NationalMaps/SoilAndLandscapeGrid/SLT/v2/SLT_000_005_EV_N_P_AU_TRN_N_20210902.tif       

### Sand 
https://data.tern.org.au/model-derived/slga/NationalMaps/SoilAndLandscapeGrid/SND/v2/SND_000_005_EV_N_P_AU_TRN_N_20210902.tif

### Organic Carbon 
https://data.tern.org.au/model-derived/slga/NationalMaps/SoilAndLandscapeGrid/SOC/90m/v2/SOC_000_005_EV_N_P_AU_TRN_N_20220727.tif

Please note that in this tutorial we use only surface maps (0 to 5cm). However, soil maps are available for another five layers, up to a depth of 2m. 
    

## Australian rangeland boundaries
We will need to define the boundaries for Australian rangelands. To do this we will use shapefiles created by the Department of Climate Change, Energy, the Environment and Water (DCCEEW). To download these files:

- Click on this link https://fed.dcceew.gov.au/datasets/erin::australian-rangeland-boundaries/about
- Click on download 
- Select Shapefile
- Extract the shapefile once you have it locally


# VEGETATION
The dataset used for this tutorial is: 

Sparrow, B., Tokmakoff, A., Leitch, E., Guerin, G., O'Neill, S., Macdonald, C., Lowe, A., Flitton, R., Saleeba, T., Coish, C. Starkey, M., Irvine, K., Francis, N., Finlayson, T., Pink, L., Finn, L., Martin-Fores, I. & TERN Ecosystem Surveillance (2021). TERN Ecosystem Surveillance. Version 2.0. Terrestrial Ecosystem Research Network. Dataset. https://portal.tern.org.au/metadata/TERN/9210c331-ba96-4e0c-9554-f029d61534be

Data were collected by TERN Surveillance according to the AusPlots Rangelands Survey Protocols https://swift.rc.nectar.org.au/v1/AUTH_05bca33fce34447ba7033b9305947f11/static-web-assets/ausplots-rangelands-manual/v1.2.9/manual.pdf

Related metadata are available here https://portal.tern.org.au/metadata/TERN/9210c331-ba96-4e0c-9554-f029d61534be

We will also need the shapefiles for the Australian rangeland boundaries: 

Commonwealth of Australia (Department of Climate Change, Energy, the Environment and Water) (2014). Australian Rangeland Boundaries. https://www.arcgis.com/home/item.html?id=bf09a05d02854dcd98caca1cc17b98f6


## Import the data
This chunk imports the csv file with the data or vegetation plot surveys. 
The first line of code in this chunk allows you to interactively select the folder where the data are saved. Hopefully you have saved all the data in this folder and will not need to select the folder again. 

```{r}
# Select data folder
data_folder <- rstudioapi::selectDirectory()

# Specify file name
file_name <- "Plant_occurrence.csv"

# Combine folder + file name
file_path <- file.path(data_folder, file_name)

# Read the CSV
vegetation <- read.csv(file_path, header = TRUE, fill = TRUE)

```


## Create a table of plot level data
We now need to create a table that will contain one row for each plot. All data summarised at the plot level (e.g. species richness, meteorological parameters, soil parameters) will be saved here. 
```{r}
#load library
library(dplyr)

# Create a table of plot-level data with selected columns
plot_level_data <- vegetation %>%
  dplyr::select(datasetTitle, siteName, latitude_Degree, longitude_Degree,
         plotArea_SquareMetre, plotWidth_Metre, plotLength_Metre, plotShape,
         pointType, siteVisitName, siteVisitStartDate, siteVisitEndDate) %>%
  dplyr::group_by(siteName) %>%
  dplyr::slice(1) %>%   # pick the first row per site (all share the same plot details)
  dplyr::ungroup()

```


## Filter out sites outside of the rangelands boundaries
The dataset contains both rangeland and non-rangeland plots. Rangelands boundaries are defined by the Dept of Climate Change, Energy, the Environment & Water (DCCEEW) and available as a shapefile. This chunk reads the shapefile and uses it to filter the plots based on their geographic coordinates (lat and lon). Plots outside of the rangeland boundaries are removed from dataset. 
#Note: ensure you have extracted the rangeland boundary file prior to running this chunk
```{r}
#library(dplyr) - remove if loaded in previous chunk
library(ggplot2)
library(maps)
library(sf)

#-------------------------------------------------------------------------------
# Show the locations of each plot in the dataset
#-------------------------------------------------------------------------------

australia_map <- map_data("world", region = "Australia") # Load the map of Australia

ggplot() +
  # Base map of Australia
  geom_polygon(data = australia_map, aes(x = long, y = lat, group = group),
               fill = "gray90", color = "gray50") +
  # All plots in the dataset
  geom_point(data = plot_level_data,
             aes(x = longitude_Degree, y = latitude_Degree),
             color = "darkgreen", size = 2) +
  coord_quickmap(xlim = c(110, 155), ylim = c(-45, -10)) +  # focus on Australia
  theme_minimal() +
  labs(title = "All vegetation survey plots in Australia",
       x = "Longitude", y = "Latitude")


#-------------------------------------------------------------------------------
# Read the rangeland boundaries shapefile
#-------------------------------------------------------------------------------

shapefile_name <- "Australian_Rangeland_Boundaries.shp"
rangeland_sf <- st_read(file.path(data_folder, shapefile_name))

# Keep only polygons marked as RANGELANDS = 1
rangeland_only_sf <- rangeland_sf %>%
  filter(RANGELANDS == 1)

# Convert plot table to simple features object
plots_sf <- st_as_sf(plot_level_data,
                     coords = c("longitude_Degree", "latitude_Degree"),
                     crs = 4326)  # WGS84

# Force both layers to EPSG:4326
plots_sf <- st_transform(plots_sf, crs = 4326)
rangeland_only_sf <- st_transform(rangeland_only_sf, crs = 4326)

# Create a map to show the location of the rangeland boundaries
ggplot() +
  # Base map of Australia
  geom_polygon(data = australia_map,
               aes(x = long, y = lat, group = group),
               fill = "gray90", color = "gray50") +
  # Rangeland polygons
  geom_sf(data = rangeland_only_sf, fill = "#FFF2B2", color = "gray60", size = 0.8, alpha = 0.6) +
  coord_sf(xlim = c(110, 155), ylim = c(-45, -10)) +
  theme_minimal() +
  labs(title = "Australian rangelands boundaries",
       x = "Longitude", y = "Latitude")


#-------------------------------------------------------------------------------
# Filter plots inside the rangeland boundaries
#-------------------------------------------------------------------------------

# Filter points inside the rangeland polygons
intersect_matrix <- st_intersects(plots_sf, rangeland_only_sf, sparse = FALSE)
points_inside <- apply(intersect_matrix, 1, any)
plots_rangeland_sf <- plots_sf[points_inside, ]

# Convert back to dataframe
coords <- st_coordinates(plots_rangeland_sf) # Extract coordinates from the geometry

plot_level_data <- plots_rangeland_sf %>%
  dplyr::mutate(
    longitude_Degree = coords[, "X"], # Add lon as separate column
    latitude_Degree = coords[, "Y"] # Add lat as separate column
  ) %>%
  st_drop_geometry()  # drop the geometry

# Create a map to show the locations of the rangeland plots
ggplot() +
  # Base map of Australia
  geom_polygon(data = australia_map,
               aes(x = long, y = lat, group = group),
               fill = "gray90", color = "gray50") +
  # Rangeland polygons 
  geom_sf(data = rangeland_only_sf, fill = "#FFF2B2", color = "gray60", size = 0.8, alpha = 0.6) +
  # Rangeland plots
  geom_point(data = plot_level_data,
             aes(x = longitude_Degree, y = latitude_Degree),
             color = "darkgreen", size = 2) +
  coord_sf(xlim = c(110, 155), ylim = c(-45, -10)) +
  theme_minimal() +
  labs(title = "Vegetation survey plots within Australian rangelands boundaries",
       x = "Longitude", y = "Latitude")

# Remove temporary tables
rm(plots_sf, plots_rangeland_sf)
```

## Calculate species richness
This code calculates the number of plant species in each plot, i.e. species richness. 
But first we need to clean the dataset: 
1) Some sites were visited multiple times. For these sites we will keep only the data from the latest visit, so that each site contributes only one survey. 
2) Remove NAs
3) Remove plants that were not identified at least at the species level
4) Check if there are any hybrids 
5) For plants that are classified below the species level (e.g. subspecies, variety, etc...) convert the taxon into species

```{r}
library(dplyr)
library(stringr)

# For sites that have multiple site visits keep only the latest one
vegetation_clean <- vegetation %>%
  group_by(siteName) %>% # group rows for each site
  filter(siteVisitName == max(siteVisitName)) %>% # keep rows only for the latest visit name within that site
  ungroup() # remove grouping after filtering

# Remove rows with NA in the 'taxon' column  
vegetation_clean <- vegetation_clean %>%
  filter(!is.na(taxon))

# Remove rows where the plant was not identified at least at the species level
vegetation_clean <- vegetation_clean %>%
  filter(str_count(taxon, "\\S+") >= 3) # Keep only rows where 'taxon' has 3 or more words (e.g. it is identified at least at the species level)

# Detect hybrids using either '×' or the word "hybrid"
hybrids_count <- vegetation_clean %>%
  filter(str_detect(taxon, "×|[Hh]ybrid")) %>%   # keep only hybrids
  group_by(taxon) %>%                             # group by taxon
  summarise(count = n(), .groups = "drop")       # count occurrences

# Keep only the first two words of the column taxon, i.e. genus and species. These are the same whether the plant is identified at the species level or below (e.g. subspecies, variety, etc...)
vegetation_clean <- vegetation_clean %>%
    mutate(taxon = word(taxon, 1, 2))  # extract first two words

# Calculate species richness
richness_table <- vegetation_clean %>%
  group_by(siteName) %>%
  summarise(
    species_richness = n_distinct(taxon),
    .groups = "drop"
  )

# Join data into main dataframe, i.e. the column species_richness is added to the plot_level_data dataframe
plot_level_data <- plot_level_data %>%
  left_join(richness_table, by = "siteName")

# Remove temporary tables
rm(richness_table) 

# Show the location of the plots, color coded by species richness
ggplot() +
  # Base map of Australia
  geom_polygon(data = australia_map,
               aes(x = long, y = lat, group = group),
               fill = "gray90", color = "gray50") +
  # Plot points, colored by variable values
  geom_point(data = plot_level_data,
             aes(x = longitude_Degree, y = latitude_Degree, color = species_richness),
             size = 2) +
  # Blue–Red continuous color scale
  scale_color_gradient(low = "white", high = "darkgreen",
                       name = "Species richness") +
  # Map extent and theme
  coord_quickmap(xlim = c(110, 155), ylim = c(-45, -10)) +
  theme_minimal() +
  labs(title = "Species richness at plot locations",
       x = "Longitude", y = "Latitude")

```
These steps reduced the dataset by about 13,000 rows.
Please note that our dataset did not contain NAs - we had already excluded them during the data selection. This step was included to show the complete workflow. 
No hybrids were found, so we didn't need to decide how to handle them. 


# TEMPERATURE
The datasets used for this tutorial are: 

Min temperature > Hutchinson, M., Kesteven, J. & Xu, T.. Monthly 1976-2005 mean daily minimum temperature: ANUClimate 1.0, 0.01 degree, Australian Coverage. Version 1. NCI Australia. Dataset. https://portal.tern.org.au/metadata/aabf33e8-8f3c-4f59-96c1-d1556348aed6 NCI Project: rr9 

Max temperature > Hutchinson, M., Kesteven, J. & Xu, T.. Monthly 1976-2005 mean daily maximum temperature: ANUClimate 1.0, 0.01 degree, Australian Coverage. Version 1. NCI Australia. Dataset. https://portal.tern.org.au/metadata/4f724e1e-c799-43cf-882b-c32c7a98419d NCI Project: rr9 

The datasets contain monthly-mean daily-minimum and -maximum temperature for the Australian continent between 1976-2005. Data were modelled by fitting trivariate thin plate smoothing spline functions of longitude, latitude and vertically exaggerated elevation to observed and estimated monthly 1976-2005 daily minimum and maximum temperature means at 1541 Bureau of Meteorology stations.

Related metadata are available here:
Min temperature > https://portal.tern.org.au/metadata/NCI/aabf33e8-8f3c-4f59-96c1-d1556348aed6

Max temperature > https://portal.tern.org.au/metadata/NCI/4f724e1e-c799-43cf-882b-c32c7a98419d

Temperature data are saved in netCDF files. 

## Import min temperature data and extract values for the rangeland plots
Individual netCDF files contain monthly-mean minimum-temperatures for Australia. The code extracts monthly values for each plot and calculates the min value.   
```{r}
library(terra)
library(dplyr)
library(ggplot2)

# List all local NetCDF files containing "temperature-min" in the filename
nc_files <- list.files(data_folder, pattern = "temperature-min.*\\.nc$", full.names = TRUE)

# Sort files to ensure month order
nc_files <- sort(nc_files)

# Prepare coordinates from plot_level_data
coords <- plot_level_data %>%
  dplyr::select(longitude_Degree, latitude_Degree)

# Convert to SpatVector for extraction
plots_sf <- vect(coords, geom = c("longitude_Degree", "latitude_Degree"), crs = "EPSG:4326")

# Extract values from each NetCDF file
temp_matrix <- sapply(nc_files, function(f) {
  r <- rast(f)
  vals <- terra::extract(r, plots_sf)[, 2]  # second column = extracted value
  return(vals)
})

# Convert to dataframe
temp_df <- as.data.frame(temp_matrix)
colnames(temp_df) <- paste0("month_", sprintf("%02d", 1:ncol(temp_df)))

# Find the minimum temperature per site across all months
plot_level_data$minimumTemperature_degreeCelsius <- apply(temp_df, 1, min, na.rm = TRUE)

# Remove temporary tables
rm(temp_df, temp_matrix)

# Show the location of the plots, color coded by min temperature
ggplot() +
  # Base map of Australia
  geom_polygon(data = australia_map,
               aes(x = long, y = lat, group = group),
               fill = "gray90", color = "gray50") +
  # Plot points, colored by variable values
  geom_point(data = plot_level_data,
             aes(x = longitude_Degree, y = latitude_Degree, color = minimumTemperature_degreeCelsius),
             size = 2) +
  # Blue–Red continuous color scale
  scale_color_gradient(low = "blue", high = "white",
                       name = "Min Temperature (°C)") +
  # Map extent and theme
  coord_quickmap(xlim = c(110, 155), ylim = c(-45, -10)) +
  theme_minimal() +
  labs(title = "Minimum temperature at plot locations",
       x = "Longitude", y = "Latitude")


```

## Import max temperature data and extract values for the rangeland plots
Individual netCDF files contain monthly-mean maximum-temperatures for Australia. The code extracts monthly values for each plot and calculates the max value.   
```{r}
library(terra)
library(dplyr)
library(ggplot2)

# List all local NetCDF files containing "temperature-max" in the filename
nc_files <- list.files(data_folder, pattern = "temperature-max.*\\.nc$", full.names = TRUE)

# Sort files to ensure month order
nc_files <- sort(nc_files)

# Extract values from each NetCDF file
temp_matrix <- sapply(nc_files, function(f) {
  r <- rast(f)
  vals <- terra::extract(r, plots_sf)[, 2]  # second column = extracted value
  return(vals)
})

# Convert to dataframe
temp_df <- as.data.frame(temp_matrix)
colnames(temp_df) <- paste0("month_", sprintf("%02d", 1:ncol(temp_df)))

# Find the maximum temperature per site across all months
plot_level_data$maximumTemperature_degreeCelsius <- apply(temp_df, 1, max, na.rm = TRUE)

# Remove temporary tables
rm(temp_df, temp_matrix)

# Show the location of the plots, color coded by max temperature
ggplot() +
  # Base map of Australia
  geom_polygon(data = australia_map,
               aes(x = long, y = lat, group = group),
               fill = "gray90", color = "gray50") +
  # Plot points, colored by variable values
  geom_point(data = plot_level_data,
             aes(x = longitude_Degree, y = latitude_Degree, color = maximumTemperature_degreeCelsius),
             size = 2) +
  # Blue–Red continuous color scale
  scale_color_gradient(low = "white", high = "red",
                       name = "Max Temperature (°C)") +
  # Map extent and theme
  coord_quickmap(xlim = c(110, 155), ylim = c(-45, -10)) +
  theme_minimal() +
  labs(title = "Maximum temperature at plot locations",
       x = "Longitude", y = "Latitude")

```


## Plot species richness vs temperature data
```{r}
library(ggplot2)

# Species richness vs min temperature
ggplot(plot_level_data, aes(x = minimumTemperature_degreeCelsius, y = species_richness)) +
  geom_point(color = "blue", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "darkblue") +
  labs(
    x = "Minimum Temperature (°C)",
    y = "Species Richness") +
  theme_minimal()

# Species richness vs monthly max temperature
ggplot(plot_level_data, aes(x = maximumTemperature_degreeCelsius, y = species_richness)) +
  geom_point(color = "red", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "darkred") +
  labs(
    x = "Maximum Temperature (°C)",
    y = "Species Richness") +
  theme_minimal()

```
The plots suggest that:
1) Species richness increases as the minimum temperature increases.
2) There is no strong trend between species richness and maximum temperature. 


# PRECIPITATION
The dataset used for this tutorial is: 

Hutchinson, M., Kesteven, J. & Xu, T.. Monthly 1976-2005 mean precipitation: ANUClimate 1.0, 0.01 degree, Australian Coverage. Version 1. NCI Australia. Dataset. https://portal.tern.org.au/metadata/3dc8c462-4605-45b4-8a76-7f96933f1610 NCI Project: rr9

The dataset contain monthly-mean total precipitation for the Australian continent between 1976-2005. Data were modelled by fitting a trivariate thin plate smoothing spline function of longitude, latitude and vertically exaggerated elevation to the square roots of observed and estimated monthly 1976-2005 rainfall means at 11,697 Bureau of Meteorology stations.

Related metadata are available here:
https://portal.tern.org.au/metadata/NCI/3dc8c462-4605-45b4-8a76-7f96933f1610

## Import precipitation data and extract values for the rangeland plots
Precipitation data are saved in netCDF files. Individual netCDF files contain monthly-mean total precipitation for Australia. The code extracts monthly values for each plot and calculates the annual cumulative value.   
```{r}
library(terra)
library(dplyr)
library(ggplot2)

# List all local NetCDF files containing "rainfall" in the filename
nc_files <- list.files(data_folder, pattern = "rainfall.*\\.nc$", full.names = TRUE)

# Sort files to ensure month order
nc_files <- sort(nc_files)

# Extract values from each NetCDF file
temp_matrix <- sapply(nc_files, function(f) {
  r <- rast(f)
  vals <- terra::extract(r, plots_sf)[, 2]  # second column = extracted value
  return(vals)
})

# Convert to dataframe
temp_df <- as.data.frame(temp_matrix)
colnames(temp_df) <- paste0("month_", sprintf("%02d", 1:ncol(temp_df)))

# Sum mean-monthly precipitation values to obtain an annual value for each site
plot_level_data$precipitationAnnualMean_millimetre <- apply(temp_df, 1, sum, na.rm = TRUE)

# Remove temporary tables
rm(temp_df, temp_matrix)

# Show the location of the plots, color coded by precipitation 
ggplot() +
  # Base map of Australia
  geom_polygon(data = australia_map,
               aes(x = long, y = lat, group = group),
               fill = "gray90", color = "gray50") +
  # Plot points, colored by variable values
  geom_point(data = plot_level_data,
             aes(x = longitude_Degree, y = latitude_Degree, color = precipitationAnnualMean_millimetre),
             size = 2) +
  # Blue–Red continuous color scale
  scale_color_gradient(low = "lightblue", high = "darkblue",
                       name = "Precipitation (mm)") +
  # Map extent and theme
  coord_quickmap(xlim = c(110, 155), ylim = c(-45, -10)) +
  theme_minimal() +
  labs(title = "Annual precipitation at plot locations",
       x = "Longitude", y = "Latitude")

```


## Plot species richness vs precipitation data
```{r}
library(ggplot2)

# Species richness vs precipitation
ggplot(plot_level_data, aes(x = precipitationAnnualMean_millimetre, y = species_richness)) +
  geom_point(color = "lightblue", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "darkblue") +
  labs(
    x = "Annual Precipitation (mm)",
    y = "Species Richness") +
  theme_minimal()

```
The plot suggests that species richness increases as annual precipitation increases. 


# SOIL PARAMETERS
The datasets used for this tutorial are: 

Searle, R., Nimalka Somarathna, P. & Malone, B. (2023). Soil and Landscape Grid National Soil Attribute Maps - Available Volumetric Water Capacity (Percent) (3 arc second resolution) Version 2. Version 2.0. Terrestrial Ecosystem Research Network. Dataset. https://dx.doi.org/10.25919/4jwj-na34     

Malone, B. & Searle, R. (2022). Soil and Landscape Grid National Soil Attribute Maps - Clay (3" resolution) - Release 2. Version 2. Terrestrial Ecosystem Research Network. Dataset. https://dx.doi.org/10.25919/hc4s-3130 

Malone, B. & Searle, R. (2024). Soil and Landscape Grid National Soil Attribute Maps - Total Nitrogen (3" resolution) - Release 2. Version 2. Terrestrial Ecosystem Research Network. Dataset. https://dx.doi.org/10.25919/pm2n-ww12 

Malone, B. & Searle, R. (2024). Soil and Landscape Grid National Soil Attribute Maps - pH - Calcium Chloride (3" resolution) - Release 2. Version 2. Terrestrial Ecosystem Research Network. Dataset. https://dx.doi.org/10.25919/7320-hw30

Malone, B. & Searle, R. (2024). Soil and Landscape Grid National Soil Attribute Maps - Total Phosphorus (3" resolution) - Release 2. Version 2. Terrestrial Ecosystem Research Network. Dataset. https://dx.doi.org/10.25919/7j78-md43 

Malone, B. & Searle, R. (2022). Soil and Landscape Grid National Soil Attribute Maps - Silt (3" resolution) - Release 2. Version 2.0. Terrestrial Ecosystem Research Network. Dataset. https://dx.doi.org/10.25919/2ew1-0w57 

Malone, B. & Searle, R. (2022). Soil and Landscape Grid National Soil Attribute Maps - Sand (3" resolution) - Release 2. Version 2.0. Terrestrial Ecosystem Research Network. Dataset. https://dx.doi.org/10.25919/rjmy-pa10 

Wadoux, A., Dobarco, M., Malone, B., Minasny, B., McBratney, A. & Searle, R. (2024). Soil and Landscape Grid National Soil Attribute Maps - Organic Carbon (3" resolution) - Release 2. Version 2.0. Terrestrial Ecosystem Research Network. Dataset. https://dx.doi.org/10.25919/ejhm-c070


These data are part of The Soil and Landscape Grid of Australia. Data were generated by applying standard Digital Soil Modelling (DSM) methods utilising publicly available soil observation data and publicly available environmental covariate. A full description of the methodology can be found here https://aussoilsdsm.esoil.io/slga-version-2-products/soil-hydraulic-properties


Related metadata are available here:

Available Water Capacity > https://portal.tern.org.au/metadata/TERN/482301c2-b9a1-4345-b142-815f9b37890a 

Clay > https://portal.tern.org.au/metadata/TERN/f95dc442-013b-4fad-b31f-91ba86fbe7f5

Nitrogen > https://portal.tern.org.au/metadata/TERN/e9484508-c705-4c23-9195-f26d64b9d4f1

pH > https://portal.tern.org.au/metadata/TERN/258afc98-7407-4781-b241-cb0293b4b8aa

Phosphorus > https://portal.tern.org.au/metadata/TERN/be382e63-5ff6-40a9-930f-c84655a5bd87

Silt > https://portal.tern.org.au/metadata/TERN/11375f04-b5cd-46a7-bcac-0e83fcb58605       

Sand > https://portal.tern.org.au/metadata/TERN/4224ddff-5fb4-4170-b5ea-c0c500599700

Organic Carbon> https://portal.tern.org.au/metadata/TERN/405bdb14-e7d2-4570-93c0-35f75d635f14


File naming convention:

SOC = Organic Carbon, i.e.	mass fraction of carbon by weight in the < 2 mm soil material as determined by dry combustion at 900° C 
CLY =	Clay, i.e.< 2 um mass fraction of the < 2 mm soil material determined using the pipette method 	
SLT = Silt, i.e.	2-20 um mass fraction of the < 2 mm soil material determined using the pipette method 	
SND = Sand, i.e.	20 um - 2 mm mass fraction of the < 2 mm soil material determined using the pipette method 
PHC = pH - CaCl2, i.e.	pH of 1:5 soil/0.01M calcium chloride extract 
AWC = Available Water Capacity
NTO = Total Nitrogen 
PTO = Total Phosphorus

Soil parameters are saved as .tif files. 

## Import soil data and extract values for the rangeland plots
```{r}
library(terra)
library(dplyr)

# List all .tif files in the folder
tif_files <- list.files(data_folder, pattern = "\\.tif$", full.names = TRUE)

# Extract unique parameters from file names (first 3 letters)
parameters <- unique(substr(basename(tif_files), 1, 3))

# Create a SpatVector of plot locations
plot_points <- vect(plot_level_data, geom = c("longitude_Degree", "latitude_Degree"), crs = "EPSG:4326")

# Loop through each parameter
for (param in parameters) {
  
  # Select the single TIFF file for this parameter
  param_file <- tif_files[substr(basename(tif_files), 1, 3) == param]
  
  # Load raster (only one layer)
  r <- rast(param_file)
  
  # Extract values at plot locations (returns a 1-column data frame)
  extracted <- terra::extract(r, plot_points, ID = FALSE)
  
  # Store as a numeric vector
  plot_level_data[[param]] <- as.numeric(extracted[, 1])
}

# Remove temporary data
rm(rasters, plot_points) 

# Rename columns
plot_level_data <- plot_level_data %>%
  rename(
    availableWaterCapacity_percentage = AWC,
    clay_percentage = CLY,
    nitrogen_percentage = NTO,
    pH = PHC,
    phosphorus_percentage = PTO,
    silt_percentage = SLT,
    sand_percentage = SND,
    organicCarbon_percentage = SOC
  )

# Show the location of the plots, color coded by available water capacity 
ggplot() +
  # Base map of Australia
  geom_polygon(data = australia_map,
               aes(x = long, y = lat, group = group),
               fill = "gray90", color = "gray50") +
  # Plot points, colored by variable values
  geom_point(data = plot_level_data,
             aes(x = longitude_Degree, y = latitude_Degree, color = availableWaterCapacity_percentage),
             size = 2) +
  # Blue–Red continuous color scale
  scale_color_gradient(low = "white", high = "brown",
                       name = "Available water capacity (percentage)") +
  # Map extent and theme
  coord_quickmap(xlim = c(110, 155), ylim = c(-45, -10)) +
  theme_minimal() +
  labs(title = "Available water capacity at plot locations",
       x = "Longitude", y = "Latitude")

# Show the location of the plots, color coded by clay 
ggplot() +
  # Base map of Australia
  geom_polygon(data = australia_map,
               aes(x = long, y = lat, group = group),
               fill = "gray90", color = "gray50") +
  # Plot points, colored by variable values
  geom_point(data = plot_level_data,
             aes(x = longitude_Degree, y = latitude_Degree, color = clay_percentage),
             size = 2) +
  # Blue–Red continuous color scale
  scale_color_gradient(low = "white", high = "brown",
                       name = "Clay (percentage)") +
  # Map extent and theme
  coord_quickmap(xlim = c(110, 155), ylim = c(-45, -10)) +
  theme_minimal() +
  labs(title = "Clay at plot locations",
       x = "Longitude", y = "Latitude")

# Show the location of the plots, color coded by nitrogen 
ggplot() +
  # Base map of Australia
  geom_polygon(data = australia_map,
               aes(x = long, y = lat, group = group),
               fill = "gray90", color = "gray50") +
  # Plot points, colored by variable values
  geom_point(data = plot_level_data,
             aes(x = longitude_Degree, y = latitude_Degree, color = nitrogen_percentage),
             size = 2) +
  # Blue–Red continuous color scale
  scale_color_gradient(low = "white", high = "brown",
                       name = "Nitrogen (percentage)") +
  # Map extent and theme
  coord_quickmap(xlim = c(110, 155), ylim = c(-45, -10)) +
  theme_minimal() +
  labs(title = "Nitrogen at plot locations",
       x = "Longitude", y = "Latitude")

# Show the location of the plots, color coded by pH 
ggplot() +
  # Base map of Australia
  geom_polygon(data = australia_map,
               aes(x = long, y = lat, group = group),
               fill = "gray90", color = "gray50") +
  # Plot points, colored by variable values
  geom_point(data = plot_level_data,
             aes(x = longitude_Degree, y = latitude_Degree, color = pH),
             size = 2) +
  # Blue–Red continuous color scale
  scale_color_gradient(low = "white", high = "brown",
                       name = "pH") +
  # Map extent and theme
  coord_quickmap(xlim = c(110, 155), ylim = c(-45, -10)) +
  theme_minimal() +
  labs(title = "pH at plot locations",
       x = "Longitude", y = "Latitude")

# Show the location of the plots, color coded by phosphorus 
ggplot() +
  # Base map of Australia
  geom_polygon(data = australia_map,
               aes(x = long, y = lat, group = group),
               fill = "gray90", color = "gray50") +
  # Plot points, colored by variable values
  geom_point(data = plot_level_data,
             aes(x = longitude_Degree, y = latitude_Degree, color = phosphorus_percentage),
             size = 2) +
  # Blue–Red continuous color scale
  scale_color_gradient(low = "white", high = "brown",
                       name = "Phosphorus (percentage)") +
  # Map extent and theme
  coord_quickmap(xlim = c(110, 155), ylim = c(-45, -10)) +
  theme_minimal() +
  labs(title = "Phosphorus at plot locations",
       x = "Longitude", y = "Latitude")

# Show the location of the plots, color coded by silt 
ggplot() +
  # Base map of Australia
  geom_polygon(data = australia_map,
               aes(x = long, y = lat, group = group),
               fill = "gray90", color = "gray50") +
  # Plot points, colored by variable values
  geom_point(data = plot_level_data,
             aes(x = longitude_Degree, y = latitude_Degree, color = silt_percentage),
             size = 2) +
  # Blue–Red continuous color scale
  scale_color_gradient(low = "white", high = "brown",
                       name = "Silt (percentage)") +
  # Map extent and theme
  coord_quickmap(xlim = c(110, 155), ylim = c(-45, -10)) +
  theme_minimal() +
  labs(title = "Silt at plot locations",
       x = "Longitude", y = "Latitude")

# Show the location of the plots, color coded by sand 
ggplot() +
  # Base map of Australia
  geom_polygon(data = australia_map,
               aes(x = long, y = lat, group = group),
               fill = "gray90", color = "gray50") +
  # Plot points, colored by variable values
  geom_point(data = plot_level_data,
             aes(x = longitude_Degree, y = latitude_Degree, color = sand_percentage),
             size = 2) +
  # Blue–Red continuous color scale
  scale_color_gradient(low = "white", high = "brown",
                       name = "Sand (percentage)") +
  # Map extent and theme
  coord_quickmap(xlim = c(110, 155), ylim = c(-45, -10)) +
  theme_minimal() +
  labs(title = "Sand at plot locations",
       x = "Longitude", y = "Latitude")

# Show the location of the plots, color coded by organicCarbon 
ggplot() +
  # Base map of Australia
  geom_polygon(data = australia_map,
               aes(x = long, y = lat, group = group),
               fill = "gray90", color = "gray50") +
  # Plot points, colored by variable values
  geom_point(data = plot_level_data,
             aes(x = longitude_Degree, y = latitude_Degree, color = organicCarbon_percentage),
             size = 2) +
  # Blue–Red continuous color scale
  scale_color_gradient(low = "white", high = "brown",
                       name = "Organic Carbon (percentage)") +
  # Map extent and theme
  coord_quickmap(xlim = c(110, 155), ylim = c(-45, -10)) +
  theme_minimal() +
  labs(title = "Organic Carbon at plot locations",
       x = "Longitude", y = "Latitude")

```


# Create graphs for each variable
```{r}
library(ggplot2)
library(tidyr)
library(dplyr)

# Variables of interest
vars <- c(
  "availableWaterCapacity_percentage",
  "clay_percentage",
  "nitrogen_percentage",
  "pH",
  "phosphorus_percentage",
  "silt_percentage",
  "sand_percentage",
  "organicCarbon_percentage"
)

# Reshape your data from wide to long format
# You want to plot the same relationship (e.g., species richness vs each soil variable) in a single call to ggplot(), with multiple facets — one facet for each variable. To do that, ggplot2 needs a single column to hold all soil variable values, and another column to indicate which variable they belong to.
plot_long <- plot_level_data %>%
  pivot_longer(
    cols = all_of(vars),
    names_to = "variable",
    values_to = "value"
  )

# Create the plot titles
plot_long$variable <- factor(plot_long$variable,
  levels = vars,
  labels = c(
    "Available Water Capacity (%)",
    "Clay (%)",
    "Nitrogen (%)",
    "pH",
    "Phosphorus (%)",
    "Silt (%)",
    "Sand (%)",
    "Organic Carbon (%)"
  )
)

# Plot species_richness vs each variable
ggplot(plot_long, aes(x = value, y = species_richness)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "darkred", linewidth = 0.8) +
  facet_wrap(~ variable, scales = "free_x") +
  theme_bw() +
  labs(
    x = "Soil property value",
    y = "Species richness"
  )

```
The plots suggest that:
1) Species richness increases as available water capacity, nitrogen, sand, and organic carbon increase. 
2) Species richness decreases as clay, pH, and silt increase.
3) There is no strong trend between species richness and phosphorus. 
































#Now to build the GLM models

But why can't we just use the simple linear regressions mentioned above?

It is possible - and best practice to start with the simplest models - but ecological data rarely meets the assumptions required for a linear regression to be accurate, such as:
- normality
- homogeneity 
- independence 
- correct model specification

Species richness is a count — it’s the number of species at a site. Counts don’t behave the way linear regression needs them to. They’re usually skewed, they often have lots of low values, and the spread of the data tends to increase as the counts get higher. 

Because these assumptions are broken, linear regression can give misleading results if we use it as the main model.

So we use Generalised Linear Models instead — like Poisson or Negative Binomial models — because they’re designed for count data and handle the messy, uneven variation you get in ecology. They give us more reliable estimates and better reflect the underlying biology

#Ok - but how do we know linear regression is not good enough? Good question - let's check!


Residuals vs Fitted = funnel shape - spread of the residuals gets wider as fitted values get larger (plain english: The model makes small errors for low richness values and much larger errors for higher richness values. Variance increases with the mean (classic for ecological count data). Linear regression assumption is violated)

QQnorm: is concave and does not follow the line - meaing residuals are skewed / not normal. another assumption violation

Scale-location plot - similar to residuals vs fitted - The model becomes less reliable at higher values.

Residuals vs leverage - clustered to far left - Most of our data points have very low leverage. That simply means none of them are extreme or unusual, and none have an unusually large influence on the fitted line.

#Why are GLMs / GLMMs so cool!?

A GLM is powerful because it lets us look at all our predictors together in one model, instead of running separate relationships one at a time.
This means we can see the unique effect of each soil variable while controlling for the others.

In other words, GLMs let us answer questions like:
‘What is the effect of nitrogen once we’ve accounted for clay, pH, moisture, and temperature?’

This is something simple linear regressions cannot do — they only show one predictor at a time and can’t separate overlapping effects. GLMs allow us to compare predictors directly, see which ones matter most, and build a more realistic picture of what’s driving species richness.

```{r}
plot(lm(species_richness ~ clay_percentage, data = plot_level_data))

#graphical validation of linear model
```


First we install library packages for the build
```{r}
# load + install the packages as needed
library(dplyr)
#install.packages("performance")
library(performance)

```


#Hold on! There is a very important step we need to do first. Because our predictor variables are on different scales e.g., percentages for soil covariates, soil pH, degrees celsius (max, min), precipitation (mm). This means we need to scale or 'centre' these variables so that they can be compared easily between models.
IMPORTANT : we are NOT centering site/visit IDs or dates.
If we do a GLMM random effects this is handled by partitioning by site / date.

#centre the predictors!

```{r}

#let's centre the predictors
#I like to see the headers so I can know which columns I want to work with for my key predictions (this is done for you below - but if you were working with fresh data - it's a handy command).

#first inspect the column headers
head(plot_level_data)

```
#create a correlation matrix to understand the level of collinearity between our predictive variables

We want to know how correlated our different predictive variables are before modelling. We already expect some correlation (e.g., soils tend to move together; min/max temps hug each other), and we also know that data collected at the same sites on the same day will be correlated. 

It's a good idea to check correlations between predictors so we can drop/merge highly redundant ones and reduce the risk of multicollinearity. Multicollinearity = predictors telling the same story so strongly that coefficients become wobbly, standard errors inflate, and the model can look unstable.

We will first create 'centred' or “_z” versions (z-centred predictors): for each predictor X, we subtract its mean and divide by its standard deviation: X_z = (X − mean(X)) / sd(X). This doesn’t change the relationships in the data; it simply puts everything on a common, unitless scale. Benefits: coefficients become comparable (“effect per 1 SD change”) across % vs mm vs °C, models tend to converge more easily (especially GLMs/GLMMs), and any interactions we add later are easier to interpret. 

We did not centre IDs like site/visit or dates as we only have one visit per site. If we had multiple site visits we could use site as a random effect within a GLMM 

We’ll use the corresponding *_z columns for the correlations and for the model formulas so everything is on the same footing.

#Here is the list of headers we want to use for this step (note - of course some will be autocorrelated):

maximumTemperature_degreeCelsius 
minimumTemperature_degreeCelsius 
temperatureAnnualMean_degreeCelsius 
precipitationAnnualMean_millimetre 
availableWaterCapacity_percentage 
clay_percentage nitrogen_percentage 
pH 
phosphorus_percentage 
silt_percentage 
sand_percentage 
organicCarbon_percentage

#

```{r}
# now that we have the columns / headers we need - let's centre the predictors
# centre the predictors (add *_z columns directly to plot_level_data)
# use the headers determined above:

pred_vars <- c(
"maximumTemperature_degreeCelsius",
"minimumTemperature_degreeCelsius",
"precipitationAnnualMean_millimetre",
"availableWaterCapacity_percentage",
"clay_percentage",
"nitrogen_percentage",
"pH",
"phosphorus_percentage",
"silt_percentage",
"sand_percentage",
"organicCarbon_percentage"
)

# 1) coerce those columns to numeric (only if they exist)

plot_level_data <- plot_level_data |>
dplyr::mutate(
dplyr::across(dplyr::any_of(pred_vars), ~ suppressWarnings(as.numeric(.x)))
)

# 2) create centred (z-scored) versions in-place (again, only for columns that exist)

plot_level_data <- plot_level_data |>
dplyr::mutate(
dplyr::across(dplyr::any_of(pred_vars), ~ as.numeric(scale(.x)), .names = "{.col}_z")
)

# quick peek (optional): which *_z columns did we just create?

created_z <- paste0(pred_vars, "_z")
intersect(created_z, names(plot_level_data))

```

#Now we're ready to run the collinearity check

```{r}
# Correlation matrix on our centred predictors (the *_z columns)

# 1) List the predictor headers we care about (exactly as in your sheet)

pred_vars <- c(
"maximumTemperature_degreeCelsius",
"minimumTemperature_degreeCelsius",
"precipitationAnnualMean_millimetre",
"availableWaterCapacity_percentage",
"clay_percentage",
"nitrogen_percentage",
"pH",
"phosphorus_percentage",
"silt_percentage",
"sand_percentage",
"organicCarbon_percentage"
)

# 2) Use the centred versions for correlation

preds_z <- intersect(paste0(pred_vars, "_z"), names(plot_level_data))

if (length(preds_z) >= 2) {

# Spearman correlation is robust for eco predictors

cor_mat <- suppressWarnings(
cor(
plot_level_data[, preds_z, drop = FALSE],
use = "pairwise.complete.obs",
method = "spearman"
)
)
print(round(cor_mat, 2))
} else {
message("Not enough centred predictors found to compute a correlation matrix (need at least 2).")
}


```


#fantastic - we can see there is some collinearity between predictors we expected e.g., temperatures / soil sand / clay. As a rule of thumb many use a cut-off of ~0.75 for eliminating predictors - but this needs to be based on expert knowledge as well. E.g., if we were really interested in a plant that had low nutrient soil requirements we may want to run a model examining the differences between these predictors in relation to model fit.

#let's visualise the correlations between predictors so it's easier to visually determine which to remove

```{r}
pred_vars <- c(
"maximumTemperature_degreeCelsius",
"minimumTemperature_degreeCelsius",
"precipitationAnnualMean_millimetre",
"availableWaterCapacity_percentage",
"clay_percentage",
"nitrogen_percentage",
"pH",
"phosphorus_percentage",
"silt_percentage",
"sand_percentage",
"organicCarbon_percentage"
)

#use the centred versions for correlation

preds_z <- paste0(pred_vars, "_z")

#compute Spearman correlations

cm <- suppressWarnings(cor(
plot_level_data[, preds_z, drop = FALSE],
use = "pairwise.complete.obs",
method = "spearman"
))

#print numeric matrix (rounded)

print(round(cm, 2))

#visual heatmap (corrplot)
#install.packages("corrplot") # run once if you don't have it

library(corrplot)

pred_vars <- c(
"maximumTemperature_degreeCelsius","minimumTemperature_degreeCelsius","precipitationAnnualMean_millimetre",
"availableWaterCapacity_percentage","clay_percentage","nitrogen_percentage","pH",
"phosphorus_percentage","silt_percentage","sand_percentage","organicCarbon_percentage"
)
preds_z <- paste0(pred_vars, "_z")

cm <- suppressWarnings(cor(
plot_level_data[, preds_z, drop = FALSE],
use = "pairwise.complete.obs",
method = "spearman"
))

#draw the heatmap inline (NO windows()/quartz()/x11())

corrplot(
cm,
method = "color",
type = "upper",
diag = FALSE,
order = "hclust",
addCoef.col = "black",
number.cex = 0.55, # smaller numbers
tl.cex = 0.65, # smaller labels
tl.srt = 45, # angled labels
mar = c(0,0,1,0)
)

```

#Which predictors to remove? Here is where our knowledge of plant ecosystems comes into play
As clay and silt are highly positively correlated (0.94), while sand and silt are highly negatively correlated (-0.99)
Let's remove silt and run separate models with sand and clay as they may have interesting effects on plant richness.

Nitrogen is also highlighted as correlated with organic Carbon (again logically) as plant available nitrogen is a key plant limiting soil nutrient - let's cut organic Carbon.

Removing correlated predictors reduces our risk of an overfitted model + multicollinearity! 

#What does this mean?

With too many, highly redundant predictors, the model starts “fitting noise.” Coefficients become unstable, standard errors inflate, and signs can flip with small data changes. You might get impressive in-sample fit, but poor generalisation out-of-sample. Pruning correlated predictors (what you’re doing) reduces this risk and makes predictor effects more interpretable.

#As sand and clay are highly negatively correlated we might want to check the VIF (variance inflation factor for these before we start modelling)

```{r}
#Option A — keep clay + sand (drop silt, mean temp, organic C)

preds_z <- c(
"minimumTemperature_degreeCelsius_z",
"maximumTemperature_degreeCelsius_z",
"precipitationAnnualMean_millimetre_z",
"availableWaterCapacity_percentage_z",
"clay_percentage_z",
"sand_percentage_z",
"nitrogen_percentage_z",
"pH_z",
"phosphorus_percentage_z"
)

preds_z <- intersect(preds_z, names(plot_level_data))

# build X matrix and keep complete rows

X <- as.matrix(plot_level_data[, preds_z, drop = FALSE])
X <- X[stats::complete.cases(X), , drop = FALSE]

# correlation matrix among predictors


R <- suppressWarnings(cor(X, use = "pairwise.complete.obs", method = "pearson"))

# VIF for standardised predictors is diag(solve(R)); fall back to pseudo-inverse if needed

invR <- tryCatch(solve(R), error = function(e) {
if (!requireNamespace("MASS", quietly = TRUE)) {
stop("Matrix is singular; install.packages('MASS') to use a pseudo-inverse.")
}
MASS::ginv(R)
})

vif_vals <- diag(invR)

# tidy table with flags

out <- data.frame(term = colnames(X), VIF = as.numeric(vif_vals))
out <- out[order(-out$VIF), ]
out$flag <- ifelse(out$VIF > 10, ">10 (problematic)",
ifelse(out$VIF > 5, ">5 (concerning)", ""))
print(out, row.names = FALSE)

```

# sand and clay are highly negatively correlated. So as noted above let's run separate models for these
 
# Fantastic - We are ready to create our models

#We only have 1 visit per site across 621 sites so we do not need to use a GLMM and partition the site visits using random effect e.g., (1 | siteName)

Therefore, we use a standard GLM (Generalized Linear Model) with appropriate
distribution families to handle our response variables.

# Note: If we had multiple visits per site, a GLMM would be appropriate to 
 account for site-level correlation and unmeasured heterogeneity (e.g., soil 
 history, management, microclimate).
 
 # --- Distribution Family Choices ---

# 1. SPECIES RICHNESS (count data: 0, 1, 2, 3, ...)
    - Natural families: Poisson or Negative Binomial
    - Ecological count data is typically OVERDISPERSED (variance > mean)
    - Poisson assumes variance = mean (too restrictive for ecological data)
    - NEGATIVE BINOMIAL allows variance > mean → better fit for species counts
    - We use: glm.nb() with family = Negative Binomial

# Summary: Use Negative Binomial GLM for species richness counts

#First let's check for incomplete rows - as the models will not converge if there are missing data

```{r}
#check for missing data:
# Check how many complete cases we have
cat("Rows in plot_level_data:", nrow(plot_level_data), "\n")

# ============================================================================
# CHECK FOR INCOMPLETE CASES IN plot_level_data (NO REMOVAL)
# ============================================================================

# Check which columns we need for models
model_cols <- c(
  "species_richness",
  "minimumTemperature_degreeCelsius_z",
  "maximumTemperature_degreeCelsius_z",
  "precipitationAnnualMean_millimetre_z",
  "availableWaterCapacity_percentage_z",
  "clay_percentage_z",
  "sand_percentage_z",
  "nitrogen_percentage_z",
  "pH_z",
  "phosphorus_percentage_z"
)

# Count NAs in each column
cat("=== CHECKING FOR MISSING DATA ===\n")
for (col in model_cols) {
  n_missing <- sum(is.na(plot_level_data[[col]]))
  cat(sprintf("%-45s: %d NAs\n", col, n_missing))
}

# Count complete cases
complete_rows <- complete.cases(plot_level_data[, model_cols])
cat("\nTotal rows:", nrow(plot_level_data), "\n")
cat("Complete cases:", sum(complete_rows), "\n")
cat("Incomplete cases:", sum(!complete_rows), "\n")

  
```

#we do have 11 incomplete rows - let's create a data subset to use in our models mod_df

```{r}
#create mod_df

needed <- c(
"species_richness",
"siteName",
"minimumTemperature_degreeCelsius_z",
"maximumTemperature_degreeCelsius_z",
"precipitationAnnualMean_millimetre_z",
"availableWaterCapacity_percentage_z",
"clay_percentage_z",
"sand_percentage_z",
"nitrogen_percentage_z",
"pH_z",
"phosphorus_percentage_z"
)

#subset to those columns

mod_df <- plot_level_data[, needed, drop = FALSE]

#count rows before/after removing incomplete cases (this is where the ~14 rows go)

n_before <- nrow(mod_df)
mod_df <- mod_df[stats::complete.cases(mod_df), , drop = FALSE]
n_after <- nrow(mod_df)

#make sure site is a clean factor

mod_df$siteName <- droplevels(factor(mod_df$siteName))

#quick audit to the console

cat("Model DF rebuilt: ", n_before, "rows ->", n_after, "rows (dropped", n_before - n_after, "incomplete rows)\n")

#sanity checks

stopifnot(all(mod_df$species_richness >= 0, na.rm = TRUE))

```
```{r}

##let's start with species richness response and clay as a predictor

library(MASS)  # for negative binomial GLM

# Since each site has only 1 observation, use GLM (no random effects)

# Model with Negative Binomial family (handles overdispersion)
m_rich_clay <- glm.nb(
  species_richness ~
    minimumTemperature_degreeCelsius_z +
    maximumTemperature_degreeCelsius_z +
    precipitationAnnualMean_millimetre_z +
    availableWaterCapacity_percentage_z +
    clay_percentage_z +
    nitrogen_percentage_z +
    pH_z +
    phosphorus_percentage_z,
  data = mod_df
)

# Check the model
summary(m_rich_clay)

```
```{r}

# Check for overdispersion and model fit
cat("Dispersion parameter (theta):", m_rich_clay$theta, "\n")

# Model diagnostics
par(mfrow = c(2, 2))
plot(m_rich_clay)

# Pearson residuals
pr_clay <- residuals(m_rich_clay, type = "pearson")
qqnorm(pr_clay)
qqline(pr_clay, col = "red")

# DHARMa diagnostics (works with GLMs too)
library(DHARMa)
sim_clay <- simulateResiduals(m_rich_clay, n = 1000)
plot(sim_clay)
testDispersion(sim_clay)
testZeroInflation(sim_clay)

```
#model fit is fairly good

residuals vs fitted / scale location and QQ residuals all look as we would expect.

And residuals vs leverage is showing the majority of the data vs outliers seems fine.

#wow some interesting results for the clay model - what is your interpretation? We'd love to hear it.

#Key Ecological Story:
Rangeland plant richness is primarily driven by temperature extremes (the min/max temp effects are opposite and strongest) and soil chemistry (pH, clay). 

The results suggest that richness is highest under moderate conditions — where minimum temperatures are warm enough to avoid cold stress, maximum temperatures are cool enough to avoid heat stress, and soils are moderately acidic with lower clay content.

Interesting that, richer plant communities are found in slightly more acidic soils in this dataset.

The fact that precipitation wasn’t significant is interesting. It suggests that, in this dataset, temperature stress may be a stronger filter on richness than water availability. This could make sense if most sites are already in water-limited rangelands where species are adapted to low rainfall, but temperature extremes still shape which species can persist.

#Ecological Summary – Clay Model

In clay-dominated rangeland soils, species richness is highest at sites that are warmer at night, cooler during the day, more acidic, and lower in clay and AWC. These conditions reduce cold stress, heat stress, and soil structural constraints, allowing more species to coexist. In contrast, high-clay, high-AWC, alkaline soils favour a smaller set of stress-tolerant dominants, reducing overall richness.

#🔍 Significant Predictors and Ecological Interpretation
#1. Minimum Temperature (positive effect)

Estimate: +0.1403

exp(Estimate): 1.1505

Effect: +15.0% richness per 1 SD warmer minimum temperature

#Ecological meaning:
Warmer nights/winters reduce cold stress → more species can persist.
Common in drylands where frost limits many annuals and forbs.

#2. Maximum Temperature (negative effect)

Estimate: −0.1028

exp(Estimate): 0.902

Effect: −9.8% richness per 1 SD hotter maximum temperature

#Ecological meaning:
High daytime heat increases evaporative demand, seedling mortality, and water stress → richness declines as Tmax rises.

3. Available Water Capacity (AWC) — negative

Estimate: −0.0525

exp(Estimate): 0.949

Effect: −5.1% richness per 1 SD higher AWC

#Ecological meaning:
High AWC = finer particles = more clay/silt.
In Australian rangelands, these soils:

hold water too tightly for plants (not plant-available)

swell/crack

reduce aeration

favour a few clay specialists

So richness is higher on low-AWC, coarse, sandy soils.

#4. Clay Percentage — negative

Estimate: −0.0651

exp(Estimate): 0.937

Effect: −6.3% richness per 1 SD more clay

#Ecological meaning:
High-clay soils impose strong physical/chemical constraints:

difficult root penetration

structural instability

slow infiltration / poor drainage

They tend to support fewer species → lower richness as clay increases.

#5. Soil pH — negative

Estimate: −0.1468

exp(Estimate): 0.864

Effect: −13.6% richness per 1 SD more alkaline

#Ecological meaning:
More alkaline soils (often sodic/saline in rangelands) strongly filter species.
Acidic to slightly acidic soils support higher richness.

🚫 Non-Significant Predictors
Precipitation (p = 0.706)

No effect once temperature and soil texture are included.
Interpretation: Water limitation is widespread; rainfall differences don’t add predictive value over soil and thermal stress.

Nitrogen (p = 0.864)
Phosphorus (p = 0.935)

No detectable influence on richness.
Interpretation: Nutrients shape productivity more than richness in these systems.

#📊 Model Fit & Diagnostics

Null deviance: 837.3

Residual deviance: 636.6 → substantial deviance explained

Theta: 5.75 → moderate overdispersion; NB is appropriate

AIC: 4403.7

Fisher iterations: 1 → clean convergence

DHARMa: no dispersion or zero-inflation issues

#let's check species richness against the same glm but with sand substiuted for sand

```{r}

# ============================================================================
# SPECIES RICHNESS ~ SAND MODEL
# ============================================================================

library(MASS)  # for glm.nb()
library(DHARMa)  # for diagnostics

# --- Fit Negative Binomial GLM with SAND as predictor ---
m_rich_sand <- glm.nb(
  species_richness ~
    minimumTemperature_degreeCelsius_z +
    maximumTemperature_degreeCelsius_z +
    precipitationAnnualMean_millimetre_z +
    availableWaterCapacity_percentage_z +
    sand_percentage_z +  
    nitrogen_percentage_z +
    pH_z + 
    phosphorus_percentage_z,
  data = mod_df
)

# --- Model summary ---
summary(m_rich_sand)


# 
```
#let's check model fit

```{r}
# --- Check dispersion parameter ---
cat("\nDispersion parameter (theta):", m_rich_sand$theta, "\n\n")

# --- Visual diagnostics ---
par(mfrow = c(2, 2))
plot(m_rich_sand)

# --- Pearson residuals QQ plot ---
pr_sand <- residuals(m_rich_sand, type = "pearson")
qqnorm(pr_sand, main = "QQ Plot: Species Richness ~ Sand Model")
qqline(pr_sand, col = "red")

# --- DHARMa diagnostics ---
sim_sand <- simulateResiduals(m_rich_sand, n = 1000)
plot(sim_sand, main = "DHARMa: Species Richness ~ Sand Model")
testDispersion(sim_sand)
testZeroInflation(sim_sand)

```


#Key insight: Sand and clay are telling the SAME story from opposite directions (they're negatively correlated, remember?). Sandier soils = less clay = better drainage + more available nutrients = higher richness. Both models are essentially equivalent in explanatory power.

#Ecological Story
rangeland species richness is maximized in:

Sandier soils (better drainage, aeration, + likely more available nutrients)
Warmer minimum temps (less cold stress)
Cooler maximum temps (less heat stress)
More acidic pH (lower pH values)
Lower phosphorus (possibly reduced competitive dominance)

Since sand and clay are just mirror images of the same gradient, possibly pick one model (probably sand, since it has slightly lower AIC and phosphorus becomes significant). 

#Pick the best model / model selection for species richness

```{r}

# Choose the sand model (slightly better AIC, phosphorus significant)
# Mention clay in text as supporting evidence

cat("=== MODEL SELECTION ===\n")
cat("Clay model AIC:", AIC(m_rich_clay), "\n")
cat("Sand model AIC:", AIC(m_rich_sand), "\n")
cat("Difference:", abs(AIC(m_rich_clay) - AIC(m_rich_sand)), "\n")
cat("\nAIC difference < 2 indicates models are essentially equivalent.\n")
cat("Choosing SAND model (marginally lower AIC, clearer P effect).\n")
cat("Clay shows inverse pattern as expected (r = -0.XX between sand/clay).\n")

```

#Species Richness:
Species richness peaks on sandy, acidic, low-P soils.
These infertile, freely draining substrates limit dominant competitors and favour a diverse mix of stress-tolerant species. Minimum temperatures boost richness, while extreme maximum temperatures reduce it.
Negative pH effect = species richness increases as pH goes down (more acidic).

#let's create a summary plot of the 2 models

```{r}
# === Clean coefficient plotting (robust; handles betareg + NB; no Log(nu)) ==
library(ggplot2)
library(dplyr)
library(stringr)

.extract_coef_table <- function(model) {
  # Returns a data.frame with columns: term, estimate, se
  if (inherits(model, "betareg")) {
    # 1) Try broom tidy (mean component only)
    td <- try({
      if (requireNamespace("broom", quietly = TRUE)) {
        broom::tidy(model, component = "mean", conf.int = FALSE)
      } else NULL
    }, silent = TRUE)
    if (inherits(td, "data.frame") && all(c("term","estimate","std.error") %in% names(td))) {
      return(td |>
               dplyr::transmute(term, estimate, se = std.error))
    }

    # 2) Fall back to summary() mean block
    sm <- summary(model)
    cf <- try(sm$coefficients$mean, silent = TRUE)
    if (inherits(cf, c("matrix","array")) && ncol(cf) >= 2) {
      df <- as.data.frame(cf)
      # Typical names: "Estimate", "Std. Error", "z value", "Pr(>|z|)"
      est_col <- names(df)[which(grepl("^estimate$", names(df), ignore.case = TRUE))[1]]
      se_col  <- names(df)[which(grepl("^std\\.?\\s*error$", names(df), ignore.case = TRUE))[1]]
      if (!is.na(est_col) && !is.na(se_col)) {
        df$term <- rownames(df); rownames(df) <- NULL
        return(df |>
                 dplyr::transmute(term, estimate = .data[[est_col]], se = .data[[se_col]]))
      }
    }

    # 3) Last resort: use coef() + vcov() for mean submodel
    est_vec <- try(coef(model)[["mean"]], silent = TRUE)
    if (inherits(est_vec, "numeric")) {
      vc  <- try(vcov(model, model = "mean"), silent = TRUE)
      se_vec <- if (inherits(vc, "matrix")) sqrt(diag(vc)) else NA_real_
      df <- data.frame(term = names(est_vec), estimate = as.numeric(est_vec), se = as.numeric(se_vec))
      return(df)
    }

    stop("Could not extract coefficients for betareg model.")
  }

  # MASS::glm.nb and similar
  if ("negbin" %in% class(model)) {
    sm <- summary(model)
    cf <- sm$coefficients
    if (inherits(cf, c("matrix","array")) && ncol(cf) >= 2) {
      df <- as.data.frame(cf)
      # Typical colnames: "Estimate", "Std. Error", ...
      est_col <- names(df)[which(grepl("^estimate$", names(df), ignore.case = TRUE))[1]]
      se_col  <- names(df)[which(grepl("^std\\.?\\s*error$", names(df), ignore.case = TRUE))[1]]
      if (is.na(est_col) || is.na(se_col)) {
        # Fallback to first two numeric columns
        num_idx <- which(vapply(df, is.numeric, logical(1)))
        if (length(num_idx) >= 2) {
          est_col <- names(df)[num_idx[1]]
          se_col  <- names(df)[num_idx[2]]
        }
      }
      df$term <- rownames(df); rownames(df) <- NULL
      return(df |>
               dplyr::transmute(term, estimate = .data[[est_col]], se = .data[[se_col]]))
    }
    stop("Could not extract coefficients for glm.nb model.")
  }

  stop(sprintf("Unsupported model class: %s", paste(class(model), collapse = "/")))
}

one_coef_plot <- function(model, model_name, title = NULL) {
  # --- extract --------------------------------------------------------------
  df <- .extract_coef_table(model)

  dat <- df |>
    dplyr::filter(term != "(Intercept)") |>
    # Belt-and-braces: drop any auxiliary/precision/nu rows
    dplyr::filter(!stringr::str_detect(term, "(?i)log\\((nu|phi)\\)|^nu$|^phi$|precision")) |>
    dplyr::mutate(
      lower = estimate - 1.96 * se,
      upper = estimate + 1.96 * se,
      term  = term |>
        stringr::str_replace("_z$", "") |>
        stringr::str_replace_all("_", " ")
    )

  aic_val <- tryCatch(round(AIC(model), 1), error = function(e) NA_real_)

  ggplot(dat, aes(x = reorder(term, abs(estimate)), y = estimate)) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_pointrange(aes(ymin = lower, ymax = upper)) +
    coord_flip() +
    labs(
      x = NULL, y = "Coefficient (±95% CI)",
      title = if (is.null(title)) model_name else title,
      subtitle = if (!is.na(aic_val)) paste0("AIC: ", aic_val) else NULL
    ) +
    theme_minimal(base_size = 12) +
    theme(plot.title = element_text(face = "bold"),
          plot.margin = margin(10, 20, 10, 10))
}

# --- your models -----------------------------------------------------------
models <- list(
  "SR ~ sand (NB)"    = m_rich_sand,
  "SR ~ clay (NB)"    = m_rich_clay
)

# --- print plots -----------------------------------------------------------
for (nm in names(models)) {
  print(one_coef_plot(models[[nm]], nm))
}

```

#what now - we are at the end of this workshop - well done!

What you can go and do now is see if dropping some of the nonsignificant predictors improves model fit further - in the interests of time - we have left this component up to you.