---
title: "TERN_workshop_ESA_2025"
author: "Elisa Girola and Eleanor Velasquez"
date: "2025-09-22"
output: word_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading the packages
```{r, include=FALSE}
# Loading the packages ----------------------------------------------------
rm(list=ls()) # cleans the workspace

```

# FOLDER PATH
For each dataset we are providing both the code to access the files from a locally saved copy and from a remote copy. Remote access is typically slower, but ensures that you are always accessing the latest version of the data. Due to time constraints, during the workshop we will access locally saved data. The code that accesses the data remotely has been commented out. 

The following code specifies the folder path to access data saved locally. You can simply copy and paste your file path here, without the quotation marks. 
#note put this before the command/chunk to knitR or there is a warning / error message.
```{r}
# Select data folder
data_folder <- rstudioapi::selectDirectory()

```



# VEGETATION

<!-- ## Import the data from Ecoplots -->
<!-- ```{r} -->


<!-- ``` -->

## Import the data from a locally saved file
```{r}
# File name
file_name <- "Plant_occurrence.csv"

# Combine folder + file name
file_path <- file.path(data_folder, file_name)

# Read the CSV
vegetation <- read.csv(file_path, header = TRUE, fill = TRUE)

```

## For sites that have multiple site visits keep only the latest one
#should we comment our justification in here just briefly? yes - let's do that - definitely
```{r}
library(dplyr)
vegetation <- vegetation %>%
  group_by(siteName) %>% # group rows for each site
  filter(siteVisitName == max(siteVisitName)) %>% # keep rows only for the latest visit name within that site
  ungroup() # remove grouping after filtering

```

# Create a table of plot level data
```{r}
# Keep one row per site
plot_level_data <- vegetation %>%
  group_by(siteName) %>%
  slice(1) %>%   # pick the first row per site (they all share the same plot details)
  ungroup()

```

## Create a map to show the locations of each plot for the rangelands dataset
```{r}
library(dplyr)
library(ggplot2)
library(maps)

# Load Australia map
australia_map <- map_data("world", region = "Australia")

# Plot
ggplot() +
  geom_polygon(data = australia_map, aes(x = long, y = lat, group = group),
               fill = "gray90", color = "gray50") +
  geom_point(data = plot_level_data,
             aes(x = longitude_Degree, y = latitude_Degree),
             color = "darkgreen", size = 2) +
  coord_quickmap(xlim = c(110, 155), ylim = c(-45, -10)) +  # focus on Australia
  theme_minimal() +
  labs(title = "Rangeland vegetation plots",
       x = "Longitude", y = "Latitude")

rm(australia_map) # remove temporary tables

```

## Calculate species richness, species diversity and fractional cover
## NOTE: FOR ROWS THAT HAVE A SUBSPECIES, WE MAY BE ABLE TO EXTRACT THE SPECIES NAME FROM TAXON (TO DO). 
```{r}
# Calculate species richness 
library(dplyr)

richness_table <- vegetation %>%
  group_by(siteName) %>%
  summarise(
    species_richness = n_distinct(scientificName),
    .groups = "drop"
  )

# Calculate species diversity (Simpson Diversity Index)
library(dplyr)
library(vegan)

simpson_table <- vegetation %>%
  group_by(siteName) %>%
  summarise(
    SDI = diversity(table(scientificName), index = "simpson"),
    .groups = "drop"
  )

# Calculate "fractional cover", i.e. proportion of points in each plot that had vegetation

# Join data into main dataframe
plot_level_data <- plot_level_data %>%
  left_join(richness_table, by = "siteName") %>%
  left_join(simpson_table, by = "siteName")



# Remove temporary data
rm(richness_table, simpson_table) 
```


# TEMPERATURE
<!-- ## Import the data from Ecoplots -->
<!-- ```{r} -->


<!-- ``` -->

## Import the data from a locally saved file
```{r}
# File name
file_name <- "Temperature.csv"

# Combine folder + file name
file_path <- file.path(data_folder, file_name)

# Read the CSV
temperature <- read.csv(file_path, header = TRUE, fill = TRUE)

```

## Select data that match plot ID and site visit for vegetation
```{r}
library(dplyr)

# Join data into main dataframe
plot_level_data <- plot_level_data %>%
  left_join(
    temperature %>%
      select(siteName, siteVisitName, 
             maximumTemperature_degreeCelsius, 
             minimumTemperature_degreeCelsius, 
             temperatureAnnualMean_degreeCelsius),
    by = c("siteName", "siteVisitName")
  )


# Remove temporary data
rm(temperature) 
```

## Plot the vegetation and temperature data
```{r}
library(ggplot2)

# Species richness vs temperature
ggplot(plot_level_data, aes(x = temperatureAnnualMean_degreeCelsius, y = species_richness)) +
  geom_point(color = "blue", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "darkblue") +
  labs(
    x = "Annual Mean Temperature (°C)",
    y = "Species Richness") +
  theme_minimal()

# Simpson Diversity Index (SDI) vs temperature
ggplot(plot_level_data, aes(x = temperatureAnnualMean_degreeCelsius, y = SDI)) +
  geom_point(color = "green", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "darkgreen") +
  labs(
    x = "Annual Mean Temperature (°C)",
    y = "Simpson Diversity Index (SDI)") +
  theme_minimal()

```

# PRECIPITATION

# NOTE: CHECK THE DEFINITION OF RANGELAND. DOES IT INCLUDE PRECIPITATION???

<!-- ## Import the data from Ecoplots -->
<!-- ```{r} -->


<!-- ``` -->

## Import the data from a locally saved file
```{r}
# File name
file_name <- "Precipitation.csv"

# Combine folder + file name
file_path <- file.path(data_folder, file_name)

# Read the CSV
precipitation <- read.csv(file_path, header = TRUE, fill = TRUE)

```

## Select data that match plot ID and sitevisit for vegetation
```{r}
library(dplyr)

# Join data into main dataframe
plot_level_data <- plot_level_data %>%
  left_join(
    precipitation %>%
      select(siteName, siteVisitName, 
             precipitationAnnualMean_millimetre),
    by = c("siteName", "siteVisitName")
  )

# Remove temporary data
rm(precipitation) 

```

## Plot the vegetation and precipitation data
```{r}
library(ggplot2)

# Species richness vs precipitation
ggplot(plot_level_data, aes(x = precipitationAnnualMean_millimetre, y = species_richness)) +
  geom_point(color = "blue", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "darkblue") +
  labs(
    x = "Annual Mean Precipitation (mm)",
    y = "Species Richness") +
  theme_minimal()

# Simpson Diversity Index (SDI) vs precipitation
ggplot(plot_level_data, aes(x = precipitationAnnualMean_millimetre, y = SDI)) +
  geom_point(color = "green", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "darkgreen") +
  labs(
    x = "Annual Mean Precipitation (mm)",
    y = "Simpson Diversity Index (SDI)") +
  theme_minimal()

```

# SOIL - LANDSCAPE
## Access data directly from WCS
This is an example to access just one file directly. More layers can be added. Please note each layer requires several minutes to run. 
```{r}
# library(terra)
# 
# # WCS endpoint and layer
# base_url <- "https://geoserver.tern.org.au/geoserver/slga/CLY_v2/wcs"
# layer <- "slga__cly_ev_n_p_au_trn_n_20210902"
# 
# # Small buffer around each point (0.01 degrees ~1 km)
# buffer <- 0.01
# 
# # Function to get clay value at a single point
# get_clay_at_point <- function(lon, lat) {
#   xmin <- lon - buffer
#   xmax <- lon + buffer
#   ymin <- lat - buffer
#   ymax <- lat + buffer
#   
#   getcov <- paste0(
#     base_url,
#     "?service=WCS&version=2.0.1&request=GetCoverage",
#     "&coverageId=", layer,
#     "&format=image/geotiff",
#     "&subset=Long(", xmin, ",", xmax, ")",
#     "&subset=Lat(", ymin, ",", ymax, ")"
#   )
#   
#   # Download tiny raster
#   r <- rast(getcov)
#   
#   # Extract value at the exact point
#   extract(r, vect(data.frame(lon=lon, lat=lat), geom=c("lon","lat"), crs="EPSG:4326"))[,2]
# }
# 
# # Apply to all points
# plot_level_data$CLY_0_5cm <- mapply(get_clay_at_point,
#                                lon = plot_level_data$longitude_Degree,
#                                lat = plot_level_data$latitude_Degree)
# 
# head(plot_level_data)

```

## From locally saved files
File naming convention is as follows. Please note we are considering only the 0-5cm layer in this example. Models exists for deeper layers.
SOC = Organic Carbon, i.e.	mass fraction of carbon by weight in the < 2 mm soil material as determined by dry combustion at 900° C 
CLY =	Clay, i.e.< 2 um mass fraction of the < 2 mm soil material determined using the pipette method 	
SLT = Silt, i.e.	2-20 um mass fraction of the < 2 mm soil material determined using the pipette method 	
SND = Sand, i.e.	20 um - 2 mm mass fraction of the < 2 mm soil material determined using the pipette method 
PHC = pH - CaCl2, i.e.	pH of 1:5 soil/0.01M calcium chloride extract 
AWC = Available Water Capacity
NTO = Total Nitrogen 
PTO = Total Phosphorus
```{r}
library(terra)
library(dplyr)

# List all .tif files in the folder
tif_files <- list.files(data_folder, pattern = "\\.tif$", full.names = TRUE)

# Extract unique parameters from file names (first 3 letters)
parameters <- unique(substr(basename(tif_files), 1, 3))

# Create a SpatVector of plot locations
plot_points <- vect(plot_level_data, geom = c("longitude_Degree", "latitude_Degree"), crs = "EPSG:4326")

# Loop through each parameter
for (param in parameters) {
  
  # Select all files for this parameter
  param_files <- tif_files[substr(basename(tif_files), 1, 3) == param]
  
  # Load rasters
  rasters <- rast(param_files)
  
  # Extract values at plot locations
  values <- terra::extract(rasters, plot_points)[,-1]  # remove ID column
  
  # Add as new column to plot_level_data
  plot_level_data[[param]] <- values
}

library(dplyr)

# Rename columns
plot_level_data <- plot_level_data %>%
  rename(
    availableWaterCapacity_percentage = AWC,
    clay_percentage = CLY,
    nitrogen_percentage = NTO,
    pH = PHC,
    phosphorus_percentage = PTO,
    silt_percentage = SLT,
    sand_percentage = SND,
    organicCarbon_percentage = SOC
  )

# Remove temporary data
rm(rasters, plot_points) 
```

# create graphs for each variable
```{r}
library(ggplot2)
library(tidyr)
library(dplyr)

# Variables of interest
vars <- c(
  "availableWaterCapacity_percentage",
  "clay_percentage",
  "nitrogen_percentage",
  "pH",
  "phosphorus_percentage",
  "silt_percentage",
  "sand_percentage",
  "organicCarbon_percentage"
)

# Reshape your data from wide to long format
# You want to plot the same relationship (e.g., species richness vs each soil variable) in a single call to ggplot(), with multiple facets — one facet for each variable. To do that, ggplot2 needs a single column to hold all soil variable values, and another column to indicate which variable they belong to.
plot_long <- plot_level_data %>%
  pivot_longer(
    cols = all_of(vars),
    names_to = "variable",
    values_to = "value"
  )

# Create the plot titles
plot_long$variable <- factor(plot_long$variable,
  levels = vars,
  labels = c(
    "Available Water Capacity (%)",
    "Clay (%)",
    "Nitrogen (%)",
    "pH",
    "Phosphorus (%)",
    "Silt (%)",
    "Sand (%)",
    "Organic Carbon (%)"
  )
)

# Plot species_richness vs each variable
ggplot(plot_long, aes(x = value, y = species_richness)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "darkred", linewidth = 0.8) +
  facet_wrap(~ variable, scales = "free_x") +
  theme_bw() +
  labs(
    x = "Soil property value",
    y = "Species richness"
  )

# Plot SDI vs each variable
ggplot(plot_long, aes(x = value, y = SDI)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "darkred", linewidth = 0.8) +
  facet_wrap(~ variable, scales = "free_x") +
  theme_bw() +
  labs(
    x = "Soil property value",
    y = "Simpson Diversity Index"
  )
```


# SOIL - SURVEILLANCE
```{r}
# Grade (soil texture)

# Soil pH 

# match locations with vegetation data > is any plot missing? 

# compare values with Landscape > choose which one is better

# create histograms for each variable

```

#Now to build the models

First we install library packages for the build
```{r}
# load + install the packages as needed
library(dplyr)
#install.packages("performance")
library(performance)

```


#Hold on! There is a very important step we need to do first. Because our preditor variables are on different scales e.g., percentages for soil covariates, soil pH, degrees celsius (max, min and annual mean), precipitation (mm). This means we need to scale or 'centre' these variables so that they can be compared easily between models.
IMPORTANT : we are NOT centering site/visit IDs or dates.
If we do a GLMM random effects this is handled by partitioning by site / date.

#centre the predictors!

```{r}

#let's centre the predictors
#I like to see the headers so I can know which columns I want to work with for my key predictions (this is done for you below - but if you were working with fresh data - it's a handy command).

#first inspect the column headers
head(plot_level_data)

```
#create a correlation matrix to understand the level of collinearity between our predictive variables

We want to know how correlated our different predictive variables are before modelling. We already expect some correlation (e.g., soils tend to move together; min/max temps hug each other), and we also know that data collected at the same sites on the same day will be correlated. 

It's a good idea to check correlations between predictors so we can drop/merge highly redundant ones and reduce the risk of multicollinearity. Multicollinearity = predictors telling the same story so strongly that coefficients become wobbly, standard errors inflate, and the model can look unstable.

Why we just created “_z” versions (z-centred predictors): for each predictor X, we subtract its mean and divide by its standard deviation: X_z = (X − mean(X)) / sd(X). This doesn’t change the relationships in the data; it simply puts everything on a common, unitless scale. Benefits: coefficients become comparable (“effect per 1 SD change”) across % vs mm vs °C, models tend to converge more easily (especially GLMs/GLMMs), and any interactions we add later are easier to interpret. 

We did not centre IDs like site/visit or dates as we only have one visit per site. If we had multiple site visits we could use site as a random effect within a GLMM 

We’ll use the corresponding *_z columns for the correlations and for the model formulas so everything is on the same footing.

#Here is the list of headers we want to use for this step (note - of course some will be autocorrelated):

maximumTemperature_degreeCelsius 
minimumTemperature_degreeCelsius 
temperatureAnnualMean_degreeCelsius 
precipitationAnnualMean_millimetre 
availableWaterCapacity_percentage 
clay_percentage nitrogen_percentage 
pH 
phosphorus_percentage 
silt_percentage 
sand_percentage 
organicCarbon_percentage

#

```{r}
# now that we have the columns / headers we need - let's centre the predictors
# centre the predictors (add *_z columns directly to plot_level_data)
# use the headers determined above:

pred_vars <- c(
"maximumTemperature_degreeCelsius",
"minimumTemperature_degreeCelsius",
"temperatureAnnualMean_degreeCelsius",
"precipitationAnnualMean_millimetre",
"availableWaterCapacity_percentage",
"clay_percentage",
"nitrogen_percentage",
"pH",
"phosphorus_percentage",
"silt_percentage",
"sand_percentage",
"organicCarbon_percentage"
)

# 1) coerce those columns to numeric (only if they exist)

plot_level_data <- plot_level_data |>
dplyr::mutate(
dplyr::across(dplyr::any_of(pred_vars), ~ suppressWarnings(as.numeric(.x)))
)

# 2) create centred (z-scored) versions in-place (again, only for columns that exist)

plot_level_data <- plot_level_data |>
dplyr::mutate(
dplyr::across(dplyr::any_of(pred_vars), ~ as.numeric(scale(.x)), .names = "{.col}_z")
)

# quick peek (optional): which *_z columns did we just create?

created_z <- paste0(pred_vars, "_z")
intersect(created_z, names(plot_level_data))

```

#Now we're ready to run the collinearity check

```{r}
# Correlation matrix on our centred predictors (the *_z columns)

# 1) List the predictor headers we care about (exactly as in your sheet)

pred_vars <- c(
"maximumTemperature_degreeCelsius",
"minimumTemperature_degreeCelsius",
"temperatureAnnualMean_degreeCelsius",
"precipitationAnnualMean_millimetre",
"availableWaterCapacity_percentage",
"clay_percentage",
"nitrogen_percentage",
"pH",
"phosphorus_percentage",
"silt_percentage",
"sand_percentage",
"organicCarbon_percentage"
)

# 2) Use the centred versions for correlation

preds_z <- intersect(paste0(pred_vars, "_z"), names(plot_level_data))

if (length(preds_z) >= 2) {

# Spearman correlation is robust for eco predictors

cor_mat <- suppressWarnings(
cor(
plot_level_data[, preds_z, drop = FALSE],
use = "pairwise.complete.obs",
method = "spearman"
)
)
print(round(cor_mat, 2))
} else {
message("Not enough centred predictors found to compute a correlation matrix (need at least 2).")
}

# Optional: a quick pairs plot (uncomment if you want a visual)

# pairs(plot_level_data[, preds_z, drop = FALSE], pch = 16)

# Optional: VIF diagnostic using 'car' (uncomment the two lines below if you want VIFs)

# install.packages("car")  # run once

# car::vif(lm(species_richness ~ ., data = cbind(species_richness = plot_level_data$species_richness,

# plot_level_data[, preds_z, drop = FALSE])))



```


#fantastic - we can see there is some colliearity between predictors we expected e.g., temperatures / soil sand / silt clay. As a rule of thumb many use a cut-off of ~0.75 for eliminating predictors - but this needs to be based on expert knowledge as well. E.g., if we were really interested in a plant that had low nutrient soil requirements we may want to run a model examining the differences between these predictors in relation to model fit.

#let's visualise the correlations between predictors so it's easier to visually determine which to remove

```{r}
pred_vars <- c(
"maximumTemperature_degreeCelsius",
"minimumTemperature_degreeCelsius",
"temperatureAnnualMean_degreeCelsius",
"precipitationAnnualMean_millimetre",
"availableWaterCapacity_percentage",
"clay_percentage",
"nitrogen_percentage",
"pH",
"phosphorus_percentage",
"silt_percentage",
"sand_percentage",
"organicCarbon_percentage"
)

#use the centred versions for correlation

preds_z <- paste0(pred_vars, "_z")

#compute Spearman correlations

cm <- suppressWarnings(cor(
plot_level_data[, preds_z, drop = FALSE],
use = "pairwise.complete.obs",
method = "spearman"
))

#print numeric matrix (rounded)

print(round(cm, 2))

#visual heatmap (corrplot)
#install.packages("corrplot") # run once if you don't have it

library(corrplot)

pred_vars <- c(
"maximumTemperature_degreeCelsius","minimumTemperature_degreeCelsius",
"temperatureAnnualMean_degreeCelsius","precipitationAnnualMean_millimetre",
"availableWaterCapacity_percentage","clay_percentage","nitrogen_percentage","pH",
"phosphorus_percentage","silt_percentage","sand_percentage","organicCarbon_percentage"
)
preds_z <- paste0(pred_vars, "_z")

cm <- suppressWarnings(cor(
plot_level_data[, preds_z, drop = FALSE],
use = "pairwise.complete.obs",
method = "spearman"
))

#draw the heatmap inline (NO windows()/quartz()/x11())

corrplot(
cm,
method = "color",
type = "upper",
diag = FALSE,
order = "hclust",
addCoef.col = "black",
number.cex = 0.55, # smaller numbers
tl.cex = 0.65, # smaller labels
tl.srt = 45, # angled labels
mar = c(0,0,1,0)
)

```

#Which predictors to remove? Here is where our knowledge of plant ecosystems comes into play
As clay and silt are highly positively correlated (0.93), while sand and silt are highly negatively correlated (-0.99)
Let's remove silt and run separate models with sand and clay separately as they may have interesting effects on plant richness.

Temperature values are also (logically) correlated. As temp annual mean is correlated with both temp maxiumum and temp minimum - let's remove the average temp.

Nitrogen is also highlighted as correlated with organic Carbon (again logically) as plant available nitrogen is a key plant limiting soil nutrient - let's cut organic Carbon.

Removing correlated predictors reduces our risk of an overfitted model + multicollinearity! 

#What does this mean?

With too many, highly redundant predictors, the model starts “fitting noise.” Coefficients become unstable, standard errors inflate, and signs can flip with small data changes. You might get impressive in-sample fit, but poor generalisation out-of-sample. Pruning correlated predictors (what you’re doing) reduces this risk and makes predictor effects more interpretable.

#As sand and clay are highly negatively correlated we might want to check the VIF (variance inflation factor for these before we start modelling)

```{r}
#Option A — keep clay + sand (drop silt, mean temp, organic C)

preds_z <- c(
"minimumTemperature_degreeCelsius_z",
"maximumTemperature_degreeCelsius_z",
"precipitationAnnualMean_millimetre_z",
"availableWaterCapacity_percentage_z",
"clay_percentage_z",
"sand_percentage_z",
"nitrogen_percentage_z",
"pH_z",
"phosphorus_percentage_z"
)

preds_z <- intersect(preds_z, names(plot_level_data))

# build X matrix and keep complete rows

X <- as.matrix(plot_level_data[, preds_z, drop = FALSE])
X <- X[stats::complete.cases(X), , drop = FALSE]

# correlation matrix among predictors


R <- suppressWarnings(cor(X, use = "pairwise.complete.obs", method = "pearson"))

# VIF for standardised predictors is diag(solve(R)); fall back to pseudo-inverse if needed

invR <- tryCatch(solve(R), error = function(e) {
if (!requireNamespace("MASS", quietly = TRUE)) {
stop("Matrix is singular; install.packages('MASS') to use a pseudo-inverse.")
}
MASS::ginv(R)
})

vif_vals <- diag(invR)

# tidy table with flags

out <- data.frame(term = colnames(X), VIF = as.numeric(vif_vals))
out <- out[order(-out$VIF), ]
out$flag <- ifelse(out$VIF > 10, ">10 (problematic)",
ifelse(out$VIF > 5, ">5 (concerning)", ""))
print(out, row.names = FALSE)

```

# sand and clay are highly negatively correlated. So as noted above let's run separate models for these
 
# Fantastic - We are ready to create our models

#We only have 1 visit per site across 785 sites so we do not need to use a GLMM and partition the site visits using random effect e.g., (1 | siteName)

Therefore, we use a standard GLM (Generalized Linear Model) with appropriate
distribution families to handle our response variables.

# Note: If we had multiple visits per site, a GLMM would be appropriate to 
 account for site-level correlation and unmeasured heterogeneity (e.g., soil 
 history, management, microclimate).

# --- Distribution Family Choices ---

# Our two response variables have different distributions:

# 1. SPECIES RICHNESS (count data: 0, 1, 2, 3, ...)
    - Natural families: Poisson or Negative Binomial
    - Ecological count data is typically OVERDISPERSED (variance > mean)
    - Poisson assumes variance = mean (too restrictive for ecological data)
    - NEGATIVE BINOMIAL allows variance > mean → better fit for species counts
    - We use: glm.nb() with family = Negative Binomial

# 2. SIMPSON DIVERSITY INDEX (SDI) (continuous proportion: 0 to 1)
    - Natural family: Beta distribution (via betareg or mgcv::betar())
    - Beta distribution models continuous (0,1) responses with flexible shapes
    - NOT appropriate: Binomial (for binary 0/1 or count/total outcomes)
    - NOT appropriate: Gaussian (assumes unbounded continuous data)
    - We use: Beta regression for SDI models

# Summary: Use Negative Binomial GLM for species richness counts,
#          and Beta GLM for Simpson Diversity Index proportions.

#First let's check for incomplete rows - as the models will not converge if there are missing data

```{r}
#check for missing data:
# Check how many complete cases we have
cat("Rows in plot_level_data:", nrow(plot_level_data), "\n")

# Check if SDI has NAs
cat("NAs in SDI:", sum(is.na(plot_level_data$SDI)), "\n")

# ============================================================================
# CHECK FOR INCOMPLETE CASES IN plot_level_data (NO REMOVAL)
# ============================================================================

# Check which columns we need for models
model_cols <- c(
  "species_richness",
  "SDI",
  "minimumTemperature_degreeCelsius_z",
  "maximumTemperature_degreeCelsius_z",
  "precipitationAnnualMean_millimetre_z",
  "availableWaterCapacity_percentage_z",
  "clay_percentage_z",
  "sand_percentage_z",
  "nitrogen_percentage_z",
  "pH_z",
  "phosphorus_percentage_z"
)

# Count NAs in each column
cat("=== CHECKING FOR MISSING DATA ===\n")
for (col in model_cols) {
  n_missing <- sum(is.na(plot_level_data[[col]]))
  cat(sprintf("%-45s: %d NAs\n", col, n_missing))
}

# Count complete cases
complete_rows <- complete.cases(plot_level_data[, model_cols])
cat("\nTotal rows:", nrow(plot_level_data), "\n")
cat("Complete cases:", sum(complete_rows), "\n")
cat("Incomplete cases:", sum(!complete_rows), "\n")

  
```

#we do have 14 incomplete rows - let's create a data subset to use in our models mod_df

```{r}
#create mod_df

needed <- c(
"species_richness",
"siteName",
"minimumTemperature_degreeCelsius_z",
"maximumTemperature_degreeCelsius_z",
"precipitationAnnualMean_millimetre_z",
"availableWaterCapacity_percentage_z",
"clay_percentage_z",
"sand_percentage_z",
"nitrogen_percentage_z",
"pH_z",
"phosphorus_percentage_z"
)

#subset to those columns

mod_df <- plot_level_data[, needed, drop = FALSE]

#count rows before/after removing incomplete cases (this is where the ~14 rows go)

n_before <- nrow(mod_df)
mod_df <- mod_df[stats::complete.cases(mod_df), , drop = FALSE]
n_after <- nrow(mod_df)

#make sure site is a clean factor

mod_df$siteName <- droplevels(factor(mod_df$siteName))

#quick audit to the console

cat("Model DF rebuilt: ", n_before, "rows ->", n_after, "rows (dropped", n_before - n_after, "incomplete rows)\n")

#sanity checks

stopifnot(all(mod_df$species_richness >= 0, na.rm = TRUE))

```
```{r}

##let's start with species richness response and clay as a predictor

library(MASS)  # for negative binomial GLM

# Since each site has only 1 observation, use GLM (no random effects)

# Model with Negative Binomial family (handles overdispersion)
m_rich_clay <- glm.nb(
  species_richness ~
    minimumTemperature_degreeCelsius_z +
    maximumTemperature_degreeCelsius_z +
    precipitationAnnualMean_millimetre_z +
    availableWaterCapacity_percentage_z +
    clay_percentage_z +
    nitrogen_percentage_z +
    pH_z +
    phosphorus_percentage_z,
  data = mod_df
)

# Check the model
summary(m_rich_clay)

```
```{r}

# Check for overdispersion and model fit
cat("Dispersion parameter (theta):", m_rich_clay$theta, "\n")

# Model diagnostics
par(mfrow = c(2, 2))
plot(m_rich_clay)

# Pearson residuals
pr_clay <- residuals(m_rich_clay, type = "pearson")
qqnorm(pr_clay)
qqline(pr_clay, col = "red")

# DHARMa diagnostics (works with GLMs too)
library(DHARMa)
sim_clay <- simulateResiduals(m_rich_clay, n = 1000)
plot(sim_clay)
testDispersion(sim_clay)
testZeroInflation(sim_clay)

```
#model fit is fairly good
#wow some interesting results for the clay model - what is your interpretation? We'd love to hear it.

##ECOLOGICAL STORY TO BE DELETED FROM WORKSHOP VERSION##

#Key Ecological Story:
Your rangeland plant diversity is primarily driven by temperature extremes (the min/max temp effects are opposite and strongest) and soil chemistry (pH, clay). 

The results suggest that diversity is highest under moderate conditions — where minimum temperatures are warm enough to avoid cold stress, maximum temperatures are cool enough to avoid heat stress, and soils are moderately acidic with lower clay content.

The fact that precipitation wasn’t significant is interesting. It suggests that, in your dataset, temperature stress may be a stronger filter on diversity than water availability. This could make sense if most sites are already in water-limited rangelands where species are adapted to low rainfall, but temperature extremes still shape which species can persist.

#Model Performance

Good model fit: The reduction from null deviance (1136.99) to residual deviance (796.15) shows your predictors explain a substantial portion of variation in species richness
Theta = 7.38: This indicates moderate overdispersion is present (if it were very high, like >50, you'd be close to Poisson; low values indicate more overdispersion). The negative binomial is definitely the right choice here.
DHARMa diagnostics look great: No dispersion issues (p=0.24) and no zero-inflation problems (p=1.0)

Significant Predictors:
Minimum Temperature (positive effect, p < 0.001)*

Coefficient: +0.121
Interpretation: For each 1 SD increase in minimum temperature, species richness increases by ~12.9% (exp(0.121) = 1.129)
Ecological meaning: Warmer minimum temps (less cold stress) → more species

#Maximum Temperature (negative effect, p < 0.001)*

Coefficient: -0.188
Interpretation: For each 1 SD increase in maximum temperature, species richness decreases by ~17.1% (exp(-0.188) = 0.829)
Ecological meaning: Extreme heat stress → fewer species can tolerate conditions

#pH (negative effect, p < 0.001)*

Coefficient: -0.139
Interpretation: For each 1 SD increase in pH (more alkaline), species richness decreases by ~13.0% (exp(-0.139) = 0.870)
Ecological meaning: More acidic soils support higher diversity in these rangelands

#Clay percentage (negative effect, p = 0.006)

Coefficient: -0.044
Interpretation: For each 1 SD increase in clay, species richness decreases by ~4.3% (exp(-0.044) = 0.957)
Ecological meaning: Heavier clay soils → lower diversity (possibly due to drainage/aeration issues)

#Phosphorus (marginally significant, p = 0.054)

Coefficient: -0.049
Interpretation: Borderline negative effect; higher P → slightly fewer species
Ecological meaning: Could indicate competitive exclusion at high nutrient levels, but interpret cautiously

#Non-significant Predictors:

Precipitation (p = 0.68): Surprisingly not significant - possibly because temperature already captures moisture stress, or your sites don't span a wide enough precipitation gradient
Available Water Capacity (p = 0.11): Marginal but not quite significant
Nitrogen (p = 0.61): No effect
Sand/Silt (not in model - excluded due to collinearity with clay)


#let's check species richness against the same glm but with sand substiuted for sand

```{r}

# ============================================================================
# SPECIES RICHNESS ~ SAND MODEL
# ============================================================================

library(MASS)  # for glm.nb()
library(DHARMa)  # for diagnostics

######################run this extra bit if sand df has dissapeared
# Quick fix: just add the sand column to mod_df
#mod_df$sand_percentage_z <- plot_level_data$sand_percentage_z[
#  match(mod_df$siteName, plot_level_data$siteName)
#]

# Verify it worked
#cat("sand_percentage_z added to mod_df\n")
#cat("NAs:", sum(is.na(mod_df$sand_percentage_z)), "\n")
#cat("Range:", range(mod_df$sand_percentage_z, na.rm = TRUE), "\n")


# --- Fit Negative Binomial GLM with SAND as predictor ---
m_rich_sand <- glm.nb(
  species_richness ~
    minimumTemperature_degreeCelsius_z +
    maximumTemperature_degreeCelsius_z +
    precipitationAnnualMean_millimetre_z +
    availableWaterCapacity_percentage_z +
    sand_percentage_z +  
    nitrogen_percentage_z +
    pH_z +
    phosphorus_percentage_z,
  data = mod_df
)

# --- Model summary ---
summary(m_rich_sand)



```
#let's check model fit

```{r}
# --- Check dispersion parameter ---
cat("\nDispersion parameter (theta):", m_rich_sand$theta, "\n\n")

# --- Visual diagnostics ---
par(mfrow = c(2, 2))
plot(m_rich_sand)

# --- Pearson residuals QQ plot ---
pr_sand <- residuals(m_rich_sand, type = "pearson")
qqnorm(pr_sand, main = "QQ Plot: Species Richness ~ Sand Model")
qqline(pr_sand, col = "red")

# --- DHARMa diagnostics ---
sim_sand <- simulateResiduals(m_rich_sand, n = 1000)
plot(sim_sand, main = "DHARMa: Species Richness ~ Sand Model")
testDispersion(sim_sand)
testZeroInflation(sim_sand)

```


#Key insight: Sand and clay are telling the SAME story from opposite directions (they're negatively correlated, remember?). Sandier soils = less clay = better drainage + more available nutrients = higher diversity. Both models are essentially equivalent in explanatory power.

#Ecological Story
rangeland species richness is maximized in:

Sandier soils (better drainage, aeration, + likely more available nutrients)
Warmer minimum temps (less cold stress)
Cooler maximum temps (less heat stress)
More acidic pH (lower pH values)
Lower phosphorus (possibly reduced competitive dominance)

Since sand and clay are just mirror images of the same gradient, possibly pick one model (probably sand, since it has slightly lower AIC and phosphorus becomes significant). 

#Pick the best model / model selection for species richness

```{r}

# Choose the sand model (slightly better AIC, phosphorus significant)
# Mention clay in text as supporting evidence

cat("=== MODEL SELECTION ===\n")
cat("Clay model AIC:", AIC(m_rich_clay), "\n")
cat("Sand model AIC:", AIC(m_rich_sand), "\n")
cat("Difference:", abs(AIC(m_rich_clay) - AIC(m_rich_sand)), "\n")
cat("\nAIC difference < 2 indicates models are essentially equivalent.\n")
cat("Choosing SAND model (marginally lower AIC, clearer P effect).\n")
cat("Clay shows inverse pattern as expected (r = -0.XX between sand/clay).\n")

```
#summary## - Elisa should we delete - I feel in the interests of time probably delete? 

```{r}
# ============================================================================
# FINAL MODEL SELECTION: SPECIES RICHNESS
# ============================================================================

cat("=== SPECIES RICHNESS MODEL SELECTION ===\n\n")

cat("Clay vs Sand comparison:\n")
cat("  Clay model AIC:", round(AIC(m_rich_clay), 2), "\n")
cat("  Sand model AIC:", round(AIC(m_rich_sand), 2), "\n")
cat("  ΔAIC =", round(abs(AIC(m_rich_clay) - AIC(m_rich_sand)), 2), "\n\n")

cat("Decision: SAND model selected (ΔAIC < 2 = models equivalent)\n")
cat("Sand and clay are negatively correlated and represent\n")
cat("opposite ends of the same soil texture gradient.\n\n")

# Calculate correlation between sand and clay
cor_sand_clay <- cor(mod_df$sand_percentage_z, 
                      plot_level_data$clay_percentage_z[
                        match(mod_df$siteName, plot_level_data$siteName)
                      ], 
                      use = "complete.obs")

cat("Correlation between sand and clay:", round(cor_sand_clay, 3), "\n\n")

cat("Final species richness model uses SAND percentage.\n")
cat("Key findings:\n")
cat("  • Sand (+): β = +0.047, p = 0.005 → sandier soils = higher richness\n")
cat("  • Min temp (+): β = +0.118, p < 0.001 → warmer minimums = higher richness\n")  
cat("  • Max temp (−): β = -0.185, p < 0.001 → cooler maximums = higher richness\n")
cat("  • pH (−): β = -0.138, p < 0.001 → more acidic soils = higher richness\n")
cat("  • Phosphorus (−): β = -0.050, p = 0.048 → lower P = higher richness\n\n")


```
#now the SDI (simpson's diversity index) models


```{r}

#1) SDI and clay

# ============================================================================
# SDI ~ CLAY MODEL
# ============================================================================

library(betareg)
library(statmod)

m_SDI_clay <- betareg(
  SDI ~
    minimumTemperature_degreeCelsius_z +
    maximumTemperature_degreeCelsius_z +
    precipitationAnnualMean_millimetre_z +
    availableWaterCapacity_percentage_z +
    clay_percentage_z +
    nitrogen_percentage_z +
    pH_z +
    phosphorus_percentage_z,
  data = plot_level_data
)

cat("=== SDI ~ CLAY MODEL ===\n")
summary(m_SDI_clay)


```
#check model fit

```{r}
# Diagnostics

# SDI ~ CLAY
par(mfrow = c(1, 2))
plot(m_SDI_clay, which = 1)   # residuals vs fitted  (OK)
plot(m_SDI_clay, which = 4)   # leverage/Cook's      (OK)

# Manual QQ + Pearson residuals (kept)
par(mfrow = c(1, 2))
res_clay <- residuals(m_SDI_clay, type = "pearson")
fit_clay <- fitted(m_SDI_clay)
plot(fit_clay, res_clay, xlab = "Fitted (mu)", ylab = "Pearson residuals"); abline(h=0,lty=2)
qqnorm(res_clay, main = "QQ: Pearson residuals (CLAY)"); qqline(res_clay, col="red")




```
#model fit interpretation
🧩 1. Why the QQ plot might look odd

A slightly curved or scattered QQ line in betareg residuals is normal because:

Residuals are not expected to be normal — the model uses a beta (bounded 0–1) likelihood, not a Gaussian one.

The residuals depend on the mean–variance relationship, which can change with predictors.
So, unlike a linear model, you don’t expect a perfect straight QQ line.

In short: a “wonky” QQ plot isn’t automatically a problem here.

#however! the residuals vs linear does have a funnel shape so we are going to test this with a “precision” (phi) sub-model

# In beta regression, φ (phi) controls the precision or spread of values around the mean (μ).
# By default, φ is constant (equal variability for all observations).
# A "phi model" lets φ vary with predictors when residuals show changing spread (heteroskedasticity).
# This improves fit and reveals where SDI is more or less consistent across environments.
# μ model → predicts mean SDI; φ model → predicts variability of SDI.


```{r}
m_SDI_clay_phi <- betareg(
  SDI ~
    minimumTemperature_degreeCelsius_z +
    maximumTemperature_degreeCelsius_z +
    precipitationAnnualMean_millimetre_z +
    availableWaterCapacity_percentage_z +
    clay_percentage_z +
    nitrogen_percentage_z +
    pH_z +
    phosphorus_percentage_z |
    clay_percentage_z + maximumTemperature_degreeCelsius_z,   # precision model
  data = plot_level_data
)

summary(m_SDI_clay_phi)

AIC(m_SDI_clay, m_SDI_clay_phi)

```
#interesting!
Because residual variance increased with the fitted values, we allowed the precision parameter (φ) to vary with environmental predictors.
The variable-precision model provided a substantially better fit (ΔAIC = 28), indicating that SDI variability changes systematically across environmental gradients.

##let's recheck model fit

```{r}
# --- Re-check fit for phi model ------------------------------------------
# Diagnostics: same 4 plots as before (but for the phi model)
par(mfrow = c(2, 2))
plot(m_SDI_clay_phi, which = 1:4)

# Extra quick looks
par(mfrow = c(1, 2))
# Pearson residuals vs fitted (should look like random noise; funnel should be reduced)
res_phi <- residuals(m_SDI_clay_phi, type = "pearson")
fit_phi <- fitted(m_SDI_clay_phi)
plot(fit_phi, res_phi, xlab = "Fitted (mu)", ylab = "Pearson residuals",
     main = "Residuals vs Fitted (phi model)")
abline(h = 0, lty = 2)

# QQ plot of Pearson residuals (wonky is OK in beta; looking for major deviations)
qqnorm(res_phi, main = "QQ: Pearson residuals (phi model)")
qqline(res_phi, col = "red")

# Compare to constant-phi model via LR test and AIC (confirms improvement)
library(lmtest)
print(lrtest(m_SDI_clay, m_SDI_clay_phi))
AIC(m_SDI_clay, m_SDI_clay_phi)

```

#let's check the model summary table for this better phi model

```{r}
# --- Inspect model summary tables ----------------------------------------
s <- summary(m_SDI_clay_phi)
s  # full console summary (includes pseudo R^2, logLik, etc.)
cat("Pseudo R^2 (McFadden):", round(s$pseudo.r.squared, 3), "\n")
```

#SDI ~ clay predictors model interpretation:

The variable-precision beta regression indicated that diversity–evenness (SDI) is most strongly influenced by temperature extremes.
Minimum temperature had a positive effect, while maximum temperature had a strong negative effect on SDI, suggesting that diversity is highest under moderate temperature conditions — warm enough to avoid cold stress but not so hot that heat stress reduces evenness.
Available water capacity showed a modest positive effect, while clay had a weak negative trend.

The precision (φ) submodel revealed that SDI variability increases under hotter maximum temperatures, indicating that diversity patterns become less consistent as temperature stress intensifies.

Together, these results show that temperature drives both the mean and variability of community evenness, while soil properties play secondary roles.


#let's move onto the model for SDI ~ sand + predictors

```{r}
# ============================================================================
# SDI ~ SAND MODEL
# ============================================================================

m_SDI_sand <- betareg(
  SDI ~
    minimumTemperature_degreeCelsius_z +
    maximumTemperature_degreeCelsius_z +
    precipitationAnnualMean_millimetre_z +
    availableWaterCapacity_percentage_z +
    sand_percentage_z +
    nitrogen_percentage_z +
    pH_z +
    phosphorus_percentage_z,
  data = plot_level_data
)

cat("\n=== SDI ~ SAND MODEL ===\n")
summary(m_SDI_sand)


```

#let's check model fit
```{r}
# SDI ~ SAND
par(mfrow = c(1, 2))
plot(m_SDI_sand, which = 1)   # residuals vs fitted
plot(m_SDI_sand, which = 4)   # leverage/Cook's

# Manual QQ + Pearson residuals (kept)
par(mfrow = c(1, 2))
res_sand <- residuals(m_SDI_sand, type = "pearson")
fit_sand <- fitted(m_SDI_sand)
plot(fit_sand, res_sand, xlab = "Fitted (mu)", ylab = "Pearson residuals"); abline(h=0,lty=2)
qqnorm(res_sand, main = "QQ: Pearson residuals (SAND)"); qqline(res_sand, col="red")


```

#again we have a bit of funnel shape - let's fit the phi model and compare

```{r}
# === Fit variable-precision (phi) model for SAND ===
m_SDI_sand_phi <- betareg(
  SDI ~
    minimumTemperature_degreeCelsius_z +
    maximumTemperature_degreeCelsius_z +
    precipitationAnnualMean_millimetre_z +
    availableWaterCapacity_percentage_z +
    sand_percentage_z +
    nitrogen_percentage_z +
    pH_z +
    phosphorus_percentage_z |
    sand_percentage_z + maximumTemperature_degreeCelsius_z,   # phi submodel
  data = plot_level_data
)

# Compare with LR test and AIC (as done for clay)
library(lmtest)
print(lrtest(m_SDI_sand, m_SDI_sand_phi))
AIC(m_SDI_sand, m_SDI_sand_phi)


```
#which one to keep?
ΔAIC ≈ 29 (–827.4 vs –798.0) → the variable-precision - phi model fits much better.

Likelihood ratio test: χ² = 33.37, df = 2, p < 0.001 → highly significant improvement.

This tells us the precision (φ) varies systematically with sand and maximum temperature—just like it did with clay.
In other words, SDI variability isn’t constant: it changes across sandier and hotter environments.


#check model fit

```{r}
# === MODEL FIT CHECK: SDI ~ SAND (phi model) ==============================

# Supported base plots for betareg (which = 1 and 4 only)
par(mfrow = c(1, 2))
plot(m_SDI_sand_phi, which = 1)  # Residuals vs fitted
plot(m_SDI_sand_phi, which = 4)  # Leverage/Cook's distance

# Additional custom diagnostic plots (same as clay)
par(mfrow = c(1, 2))
res_sand_phi <- residuals(m_SDI_sand_phi, type = "pearson")
fit_sand_phi <- fitted(m_SDI_sand_phi)

# Residuals vs fitted (to check funnel pattern)
plot(fit_sand_phi, res_sand_phi,
     xlab = "Fitted (mu)",
     ylab = "Pearson residuals",
     main = "Residuals vs Fitted (SAND, phi model)")
abline(h = 0, lty = 2, col = "grey40")

# QQ plot (shape can be slightly curved in beta models)
qqnorm(res_sand_phi, main = "QQ: Pearson residuals (SAND, phi model)")
qqline(res_sand_phi, col = "red", lty = 2)

```
#it's a better fit - let's check the phi SDI ~ sand model
```{r}
# === MODEL SUMMARY: SDI ~ SAND (phi model) ===============================

# View full model summary
summary(m_SDI_sand_phi)
```

#sand phi model interpretation:
Temperature sets the tone. SDI (evenness/diversity) is highest where minimum temps are warmer and maximum temps are cooler (μ: min T +, max T −; both significant). That’s a “moderate temperatures = more even communities” story.

Soil water matters. Available water capacity has a negative coefficient (μ), indicating lower AWC → higher evenness (significant). In these rangelands, places with less plant-available water tend to have communities that are more even.

Texture signal from sand. Sand % is positive and significant in the φ model: μ: sand + (significant) → sandier sites show slightly higher evenness after controlling for climate and chemistry.

Rainfall is secondary. Precipitation is only marginal (μ, p≈0.07), consistent with temperature extremes being the stronger filter here.

Variance tells an extra story (φ model). Max temperature strongly reduces φ (φ: max T −, p<0.001), meaning hotter sites have more variable SDI (less predictable evenness). Sand’s effect on φ is small/non-sig, so most variance changes are tied to heat rather than texture.

One-liner: In these sandier rangelands, community evenness is highest under warm nights, cooler days, and lower AWC, with a modest boost from sandiness; heat stress not only lowers evenness on average, it also makes it more variable across sites.


#The Overall Ecological Story:
#Elisa - let's check that this actually makes sense - can you read and see if the interpretation makes sense?#

#Species Richness:
Species richness is highest on coarse-textured, acidic, low-phosphorus soils—conditions typical of freely draining sandy substrates. These soils reduce competitive exclusion and favour the coexistence of many stress-tolerant species. Temperature also influences richness, but soil texture and chemistry act as the primary environmental filters.

#Diversity (SDI):
Once species are established, temperature extremes control community evenness: warmer minimum temperatures enhance evenness, while hotter maximum temperatures reduce it. Lower water-holding capacity (AWC) is associated with higher evenness, consistent with environments where limited resources prevent dominance by a few species.

#Variance (φ model):
Hotter conditions increase the variability of SDI across sites, suggesting that heat stress not only lowers evenness on average but also makes community structure less predictable.


#let's create a summary plot of all 4 models

```{r}
# === Clean coefficient plotting (robust; handles betareg + NB; no Log(nu)) ==
library(ggplot2)
library(dplyr)
library(stringr)

.extract_coef_table <- function(model) {
  # Returns a data.frame with columns: term, estimate, se
  if (inherits(model, "betareg")) {
    # 1) Try broom tidy (mean component only)
    td <- try({
      if (requireNamespace("broom", quietly = TRUE)) {
        broom::tidy(model, component = "mean", conf.int = FALSE)
      } else NULL
    }, silent = TRUE)
    if (inherits(td, "data.frame") && all(c("term","estimate","std.error") %in% names(td))) {
      return(td |>
               dplyr::transmute(term, estimate, se = std.error))
    }

    # 2) Fall back to summary() mean block
    sm <- summary(model)
    cf <- try(sm$coefficients$mean, silent = TRUE)
    if (inherits(cf, c("matrix","array")) && ncol(cf) >= 2) {
      df <- as.data.frame(cf)
      # Typical names: "Estimate", "Std. Error", "z value", "Pr(>|z|)"
      est_col <- names(df)[which(grepl("^estimate$", names(df), ignore.case = TRUE))[1]]
      se_col  <- names(df)[which(grepl("^std\\.?\\s*error$", names(df), ignore.case = TRUE))[1]]
      if (!is.na(est_col) && !is.na(se_col)) {
        df$term <- rownames(df); rownames(df) <- NULL
        return(df |>
                 dplyr::transmute(term, estimate = .data[[est_col]], se = .data[[se_col]]))
      }
    }

    # 3) Last resort: use coef() + vcov() for mean submodel
    est_vec <- try(coef(model)[["mean"]], silent = TRUE)
    if (inherits(est_vec, "numeric")) {
      vc  <- try(vcov(model, model = "mean"), silent = TRUE)
      se_vec <- if (inherits(vc, "matrix")) sqrt(diag(vc)) else NA_real_
      df <- data.frame(term = names(est_vec), estimate = as.numeric(est_vec), se = as.numeric(se_vec))
      return(df)
    }

    stop("Could not extract coefficients for betareg model.")
  }

  # MASS::glm.nb and similar
  if ("negbin" %in% class(model)) {
    sm <- summary(model)
    cf <- sm$coefficients
    if (inherits(cf, c("matrix","array")) && ncol(cf) >= 2) {
      df <- as.data.frame(cf)
      # Typical colnames: "Estimate", "Std. Error", ...
      est_col <- names(df)[which(grepl("^estimate$", names(df), ignore.case = TRUE))[1]]
      se_col  <- names(df)[which(grepl("^std\\.?\\s*error$", names(df), ignore.case = TRUE))[1]]
      if (is.na(est_col) || is.na(se_col)) {
        # Fallback to first two numeric columns
        num_idx <- which(vapply(df, is.numeric, logical(1)))
        if (length(num_idx) >= 2) {
          est_col <- names(df)[num_idx[1]]
          se_col  <- names(df)[num_idx[2]]
        }
      }
      df$term <- rownames(df); rownames(df) <- NULL
      return(df |>
               dplyr::transmute(term, estimate = .data[[est_col]], se = .data[[se_col]]))
    }
    stop("Could not extract coefficients for glm.nb model.")
  }

  stop(sprintf("Unsupported model class: %s", paste(class(model), collapse = "/")))
}

one_coef_plot <- function(model, model_name, title = NULL) {
  # --- extract --------------------------------------------------------------
  df <- .extract_coef_table(model)

  dat <- df |>
    dplyr::filter(term != "(Intercept)") |>
    # Belt-and-braces: drop any auxiliary/precision/nu rows
    dplyr::filter(!stringr::str_detect(term, "(?i)log\\((nu|phi)\\)|^nu$|^phi$|precision")) |>
    dplyr::mutate(
      lower = estimate - 1.96 * se,
      upper = estimate + 1.96 * se,
      term  = term |>
        stringr::str_replace("_z$", "") |>
        stringr::str_replace_all("_", " ")
    )

  aic_val <- tryCatch(round(AIC(model), 1), error = function(e) NA_real_)

  ggplot(dat, aes(x = reorder(term, abs(estimate)), y = estimate)) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_pointrange(aes(ymin = lower, ymax = upper)) +
    coord_flip() +
    labs(
      x = NULL, y = "Coefficient (±95% CI)",
      title = if (is.null(title)) model_name else title,
      subtitle = if (!is.na(aic_val)) paste0("AIC: ", aic_val) else NULL
    ) +
    theme_minimal(base_size = 12) +
    theme(plot.title = element_text(face = "bold"),
          plot.margin = margin(10, 20, 10, 10))
}

# --- your models -----------------------------------------------------------
models <- list(
  "SR ~ sand (NB)"    = m_rich_sand,
  "SR ~ clay (NB)"    = m_rich_clay,
  "SDI ~ sand (beta)" = m_SDI_sand_phi,
  "SDI ~ clay (beta)" = m_SDI_clay_phi
)

# --- print plots -----------------------------------------------------------
for (nm in names(models)) {
  print(one_coef_plot(models[[nm]], nm))
}

```


