---
title: "TERN_workshop_ESA_2025"
author: "Elisa Girola and Eleanor Velasquez"
date: "2025-09-22"
output: word_document
editor_options: 
  chunk_output_type: inline
---
#note for some reason this chunk tries to open python in my Rstudio - chat gpt did suggest some ways to fix it - it's very weird!
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading the packages
```{r, include=FALSE}
# Loading the packages ----------------------------------------------------
rm(list=ls()) # cleans the workspace

```

# FOLDER PATH
For each dataset we are providing both the code to access the files from a locally saved copy and from a remote copy. Remote access is typically slower, but ensures that you are always accessing the latest version of the data. Due to time constraints, during the workshop we will access locally saved data. The code that accesses the data remotely has been commented out. 

The following code specifies the folder path to access data saved locally. You can simply copy and paste your file path here, without the quotation marks. 
```{r}
# Select data folder
data_folder <- rstudioapi::selectDirectory()

```



# VEGETATION

<!-- ## Import the data from Ecoplots -->
<!-- ```{r} -->


<!-- ``` -->

## Import the data from a locally saved file
```{r}
# File name
file_name <- "Plant_occurrence.csv"

# Combine folder + file name
file_path <- file.path(data_folder, file_name)

# Read the CSV
vegetation <- read.csv(file_path, header = TRUE, fill = TRUE)

```

## For sites that have multiple site visits keep only the latest one
```{r}
library(dplyr)
vegetation <- vegetation %>%
  group_by(siteName) %>% # group rows for each site
  filter(siteVisitName == max(siteVisitName)) %>% # keep rows only for the latest visit name within that site
  ungroup() # remove grouping after filtering

```

# Create a table of plot level data
```{r}
# Keep one row per site
plot_level_data <- vegetation %>%
  group_by(siteName) %>%
  slice(1) %>%   # pick the first row per site (they all share the same plot details)
  ungroup()

```

## Create a map to show the locations of each plot
```{r}
library(dplyr)
library(ggplot2)
library(maps)

# Load Australia map
australia_map <- map_data("world", region = "Australia")

# Plot
ggplot() +
  geom_polygon(data = australia_map, aes(x = long, y = lat, group = group),
               fill = "gray90", color = "gray50") +
  geom_point(data = plot_level_data,
             aes(x = longitude_Degree, y = latitude_Degree),
             color = "darkgreen", size = 2) +
  coord_quickmap(xlim = c(110, 155), ylim = c(-45, -10)) +  # focus on Australia
  theme_minimal() +
  labs(title = "Rangeland vegetation plots",
       x = "Longitude", y = "Latitude")

rm(australia_map) # remove temporary tables

```

## Calculate species richness, species diversity and fractional cover
## NOTE: FOR ROWS THAT HAVE A SUBSPECIES, WE MAY BE ABLE TO EXTRACT THE SPECIES NAME FROM TAXON (TO DO). 
```{r}
# Calculate species richness 
library(dplyr)

richness_table <- vegetation %>%
  group_by(siteName) %>%
  summarise(
    species_richness = n_distinct(scientificName),
    .groups = "drop"
  )

# Calculate species diversity (Simpson Diversity Index)
library(dplyr)
library(vegan)

simpson_table <- vegetation %>%
  group_by(siteName) %>%
  summarise(
    SDI = diversity(table(scientificName), index = "simpson"),
    .groups = "drop"
  )

# Calculate "fractional cover", i.e. proportion of points in each plot that had vegetation

# Join data into main dataframe
plot_level_data <- plot_level_data %>%
  left_join(richness_table, by = "siteName") %>%
  left_join(simpson_table, by = "siteName")



# Remove temporary data
rm(richness_table, simpson_table) 
```


# TEMPERATURE
<!-- ## Import the data from Ecoplots -->
<!-- ```{r} -->


<!-- ``` -->

## Import the data from a locally saved file
```{r}
# File name
file_name <- "Temperature.csv"

# Combine folder + file name
file_path <- file.path(data_folder, file_name)

# Read the CSV
temperature <- read.csv(file_path, header = TRUE, fill = TRUE)

```

## Select data that match plot ID and site visit for vegetation
```{r}
library(dplyr)

# Join data into main dataframe
plot_level_data <- plot_level_data %>%
  left_join(
    temperature %>%
      select(siteName, siteVisitName, 
             maximumTemperature_degreeCelsius, 
             minimumTemperature_degreeCelsius, 
             temperatureAnnualMean_degreeCelsius),
    by = c("siteName", "siteVisitName")
  )


# Remove temporary data
rm(temperature) 
```

## Plot the vegetation and temperature data
```{r}
library(ggplot2)

# Species richness vs temperature
ggplot(plot_level_data, aes(x = temperatureAnnualMean_degreeCelsius, y = species_richness)) +
  geom_point(color = "blue", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "darkblue") +
  labs(
    x = "Annual Mean Temperature (°C)",
    y = "Species Richness") +
  theme_minimal()

# Simpson Diversity Index (SDI) vs temperature
ggplot(plot_level_data, aes(x = temperatureAnnualMean_degreeCelsius, y = SDI)) +
  geom_point(color = "green", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "darkgreen") +
  labs(
    x = "Annual Mean Temperature (°C)",
    y = "Simpson Diversity Index (SDI)") +
  theme_minimal()

```

# PRECIPITATION

# NOTE: CHECK THE DEFINITION OF RANGELAND. DOES IT INCLUDE PRECIPITATION???

<!-- ## Import the data from Ecoplots -->
<!-- ```{r} -->


<!-- ``` -->

## Import the data from a locally saved file
```{r}
# File name
file_name <- "Precipitation.csv"

# Combine folder + file name
file_path <- file.path(data_folder, file_name)

# Read the CSV
precipitation <- read.csv(file_path, header = TRUE, fill = TRUE)

```

## Select data that match plot ID and sitevisit for vegetation
```{r}
library(dplyr)

# Join data into main dataframe
plot_level_data <- plot_level_data %>%
  left_join(
    precipitation %>%
      select(siteName, siteVisitName, 
             precipitationAnnualMean_millimetre),
    by = c("siteName", "siteVisitName")
  )

# Remove temporary data
rm(precipitation) 

```

## Plot the vegetation and precipitation data
```{r}
library(ggplot2)

# Species richness vs precipitation
ggplot(plot_level_data, aes(x = precipitationAnnualMean_millimetre, y = species_richness)) +
  geom_point(color = "blue", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "darkblue") +
  labs(
    x = "Annual Mean Precipitation (mm)",
    y = "Species Richness") +
  theme_minimal()

# Simpson Diversity Index (SDI) vs precipitation
ggplot(plot_level_data, aes(x = precipitationAnnualMean_millimetre, y = SDI)) +
  geom_point(color = "green", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "darkgreen") +
  labs(
    x = "Annual Mean Precipitation (mm)",
    y = "Simpson Diversity Index (SDI)") +
  theme_minimal()

```

# SOIL - LANDSCAPE
## Access data directly from WCS
This is an example to access just one file directly. More layers can be added. Please note each layer requires several minutes to run. 
```{r}
# library(terra)
# 
# # WCS endpoint and layer
# base_url <- "https://geoserver.tern.org.au/geoserver/slga/CLY_v2/wcs"
# layer <- "slga__cly_ev_n_p_au_trn_n_20210902"
# 
# # Small buffer around each point (0.01 degrees ~1 km)
# buffer <- 0.01
# 
# # Function to get clay value at a single point
# get_clay_at_point <- function(lon, lat) {
#   xmin <- lon - buffer
#   xmax <- lon + buffer
#   ymin <- lat - buffer
#   ymax <- lat + buffer
#   
#   getcov <- paste0(
#     base_url,
#     "?service=WCS&version=2.0.1&request=GetCoverage",
#     "&coverageId=", layer,
#     "&format=image/geotiff",
#     "&subset=Long(", xmin, ",", xmax, ")",
#     "&subset=Lat(", ymin, ",", ymax, ")"
#   )
#   
#   # Download tiny raster
#   r <- rast(getcov)
#   
#   # Extract value at the exact point
#   extract(r, vect(data.frame(lon=lon, lat=lat), geom=c("lon","lat"), crs="EPSG:4326"))[,2]
# }
# 
# # Apply to all points
# plot_level_data$CLY_0_5cm <- mapply(get_clay_at_point,
#                                lon = plot_level_data$longitude_Degree,
#                                lat = plot_level_data$latitude_Degree)
# 
# head(plot_level_data)

```

## From locally saved files
File naming convention is as follows. Please note we are considering only the 0-5cm layer in this example. Models exists for deeper layers.
SOC = Organic Carbon, i.e.	mass fraction of carbon by weight in the < 2 mm soil material as determined by dry combustion at 900° C 
CLY =	Clay, i.e.< 2 um mass fraction of the < 2 mm soil material determined using the pipette method 	
SLT = Silt, i.e.	2-20 um mass fraction of the < 2 mm soil material determined using the pipette method 	
SND = Sand, i.e.	20 um - 2 mm mass fraction of the < 2 mm soil material determined using the pipette method 
PHC = pH - CaCl2, i.e.	pH of 1:5 soil/0.01M calcium chloride extract 
AWC = Available Water Capacity
NTO = Total Nitrogen 
PTO = Total Phosphorus
```{r}
library(terra)
library(dplyr)

# List all .tif files in the folder
tif_files <- list.files(data_folder, pattern = "\\.tif$", full.names = TRUE)

# Extract unique parameters from file names (first 3 letters)
parameters <- unique(substr(basename(tif_files), 1, 3))

# Create a SpatVector of plot locations
plot_points <- vect(plot_level_data, geom = c("longitude_Degree", "latitude_Degree"), crs = "EPSG:4326")

# Loop through each parameter
for (param in parameters) {
  
  # Select all files for this parameter
  param_files <- tif_files[substr(basename(tif_files), 1, 3) == param]
  
  # Load rasters
  rasters <- rast(param_files)
  
  # Extract values at plot locations
  values <- terra::extract(rasters, plot_points)[,-1]  # remove ID column
  
  # Add as new column to plot_level_data
  plot_level_data[[param]] <- values
}

library(dplyr)

# Rename columns
plot_level_data <- plot_level_data %>%
  rename(
    availableWaterCapacity_percentage = AWC,
    clay_percentage = CLY,
    nitrogen_percentage = NTO,
    pH = PHC,
    phosphorus_percentage = PTO,
    silt_percentage = SLT,
    sand_percentage = SND,
    organicCarbon_percentage = SOC
  )

# Remove temporary data
rm(rasters, plot_points) 
```

# create graphs for each variable
```{r}
library(ggplot2)
library(tidyr)
library(dplyr)

# Variables of interest
vars <- c(
  "availableWaterCapacity_percentage",
  "clay_percentage",
  "nitrogen_percentage",
  "pH",
  "phosphorus_percentage",
  "silt_percentage",
  "sand_percentage",
  "organicCarbon_percentage"
)

# Reshape your data from wide to long format
# You want to plot the same relationship (e.g., species richness vs each soil variable) in a single call to ggplot(), with multiple facets — one facet for each variable. To do that, ggplot2 needs a single column to hold all soil variable values, and another column to indicate which variable they belong to.
plot_long <- plot_level_data %>%
  pivot_longer(
    cols = all_of(vars),
    names_to = "variable",
    values_to = "value"
  )

# Create the plot titles
plot_long$variable <- factor(plot_long$variable,
  levels = vars,
  labels = c(
    "Available Water Capacity (%)",
    "Clay (%)",
    "Nitrogen (%)",
    "pH",
    "Phosphorus (%)",
    "Silt (%)",
    "Sand (%)",
    "Organic Carbon (%)"
  )
)

# Plot species_richness vs each variable
ggplot(plot_long, aes(x = value, y = species_richness)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "darkred", linewidth = 0.8) +
  facet_wrap(~ variable, scales = "free_x") +
  theme_bw() +
  labs(
    x = "Soil property value",
    y = "Species richness"
  )

# Plot SDI vs each variable
ggplot(plot_long, aes(x = value, y = SDI)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "darkred", linewidth = 0.8) +
  facet_wrap(~ variable, scales = "free_x") +
  theme_bw() +
  labs(
    x = "Soil property value",
    y = "Simpson Diversity Index"
  )
```


# SOIL - SURVEILLANCE
```{r}
# Grade (soil texture)

# Soil pH 

# match locations with vegetation data > is any plot missing? 

# compare values with Landscape > choose which one is better

# create histograms for each variable

```

#Now to build the models

First we install library packages for the build
```{r}
# load + install the packages as needed
library(dplyr)
#install.packages("performance")
library(performance)

```


#Hold on! There is a very important step we need to do first. Because our preditor variables are on different scales e.g., percentages for soil covariates, soil pH, degrees celsius (max, min and annual mean), precipitation (mm). This means we need to scale or 'centre' these variables so that they can be compared easily between models.
IMPORTANT : we are NOT centring site/visit IDs or dates.
Those are handled by the GLMM random effects later (partitioning by site / date).

#centre the predictors!

```{r}

#let's centre the predictors
#I like to see the headers so I can know which columns I want to work with for my key predictions (this is done for you below - but if you were working with fresh data - it's a handy command).

#first inspect the column headers
head(plot_level_data)

```
#create a correlation matrix to understand the level of collinearity between our predictive variables

We want to know how correlated our different predictive variables are before modelling. We already expect some correlation (e.g., soils tend to move together; min/max temps hug each other), and we also know that data collected at the same sites on the same day will be correlated. The models we’re going to use (GLMMs) will partition some of that site/day dependence via random effects, but it’s still a good idea to check correlations between predictors so we can drop/merge highly redundant ones and reduce the risk of multicollinearity. Multicollinearity = predictors telling the same story so strongly that coefficients become wobbly, standard errors inflate, and the model can look unstable.

Why we just created “_z” versions (z-centred predictors): for each predictor X, we subtract its mean and divide by its standard deviation: X_z = (X − mean(X)) / sd(X). This doesn’t change the relationships in the data; it simply puts everything on a common, unitless scale. Benefits: coefficients become comparable (“effect per 1 SD change”) across % vs mm vs °C, models tend to converge more easily (especially GLMMs), and any interactions we add later are easier to interpret. We did not centre IDs like site/visit or dates (those are random effects later), and we’re leaving latitude/longitude for the spatial correlogram rather than as fixed effects.
We’ll use the corresponding *_z columns for the correlations and for the model formulas so everything is on the same footing.

#Here is the list of headers we want to use for this step (note - of course some will be autocorrelated):

maximumTemperature_degreeCelsius 
minimumTemperature_degreeCelsius 
temperatureAnnualMean_degreeCelsius 
precipitationAnnualMean_millimetre 
availableWaterCapacity_percentage 
clay_percentage nitrogen_percentage 
pH 
phosphorus_percentage 
silt_percentage 
sand_percentage 
organicCarbon_percentage

#

```{r}
# now that we have the columns / headers we need - let's centre the predictors
# centre the predictors (add *_z columns directly to plot_level_data)
# se the headers determined above:

pred_vars <- c(
"maximumTemperature_degreeCelsius",
"minimumTemperature_degreeCelsius",
"temperatureAnnualMean_degreeCelsius",
"precipitationAnnualMean_millimetre",
"availableWaterCapacity_percentage",
"clay_percentage",
"nitrogen_percentage",
"pH",
"phosphorus_percentage",
"silt_percentage",
"sand_percentage",
"organicCarbon_percentage"
)

# 1) coerce those columns to numeric quietly (only if they exist)

plot_level_data <- plot_level_data |>
dplyr::mutate(
dplyr::across(dplyr::any_of(pred_vars), ~ suppressWarnings(as.numeric(.x)))
)

# 2) create centred (z-scored) versions in-place (again, only for columns that exist)

plot_level_data <- plot_level_data |>
dplyr::mutate(
dplyr::across(dplyr::any_of(pred_vars), ~ as.numeric(scale(.x)), .names = "{.col}_z")
)

# quick peek (optional): which *_z columns did we just create?

created_z <- paste0(pred_vars, "_z")
intersect(created_z, names(plot_level_data))

```

#Now we're ready to run the collinearity check

```{r}
# Correlation matrix on our centred predictors (the *_z columns

# 1) List the predictor headers we care about (exactly as in your sheet)

pred_vars <- c(
"maximumTemperature_degreeCelsius",
"minimumTemperature_degreeCelsius",
"temperatureAnnualMean_degreeCelsius",
"precipitationAnnualMean_millimetre",
"availableWaterCapacity_percentage",
"clay_percentage",
"nitrogen_percentage",
"pH",
"phosphorus_percentage",
"silt_percentage",
"sand_percentage",
"organicCarbon_percentage"
)

# 2) Use the centred versions for correlation

preds_z <- intersect(paste0(pred_vars, "_z"), names(plot_level_data))

if (length(preds_z) >= 2) {

# Spearman correlation is robust for eco predictors

cor_mat <- suppressWarnings(
cor(
plot_level_data[, preds_z, drop = FALSE],
use = "pairwise.complete.obs",
method = "spearman"
)
)
print(round(cor_mat, 2))
} else {
message("Not enough centred predictors found to compute a correlation matrix (need at least 2).")
}

# Optional: a quick pairs plot (uncomment if you want a visual)

# pairs(plot_level_data[, preds_z, drop = FALSE], pch = 16)

# Optional: VIF diagnostic using 'car' (uncomment the two lines below if you want VIFs)

# install.packages("car")  # run once

# car::vif(lm(species_richness ~ ., data = cbind(species_richness = plot_level_data$species_richness,

# plot_level_data[, preds_z, drop = FALSE])))



```


#fantastic - we can see there is some colliearity between predictors we expected e.g., temperatures / soil sand / silt clay. As a rule of thumb many use a cut-off of ~0.75 for eliminating predictors - but this needs to be based on expert knowledge as well. E.g., if we were really interested in a plant that had low nutrient soil requirements we may want to run a model examining the differences between these predictors in relation to model fit.

#let's visualise the correlations between predictors so it's easier to determine which to remove

```{r}
pred_vars <- c(
"maximumTemperature_degreeCelsius",
"minimumTemperature_degreeCelsius",
"temperatureAnnualMean_degreeCelsius",
"precipitationAnnualMean_millimetre",
"availableWaterCapacity_percentage",
"clay_percentage",
"nitrogen_percentage",
"pH",
"phosphorus_percentage",
"silt_percentage",
"sand_percentage",
"organicCarbon_percentage"
)

#use the centred versions for correlation


preds_z <- paste0(pred_vars, "_z")

#compute Spearman correlations

cm <- suppressWarnings(cor(
plot_level_data[, preds_z, drop = FALSE],
use = "pairwise.complete.obs",
method = "spearman"
))

#print numeric matrix (rounded)

print(round(cm, 2))

#visual heatmap (corrplot)
#install.packages("corrplot") # run once if you don't have it

library(corrplot)

pred_vars <- c(
"maximumTemperature_degreeCelsius","minimumTemperature_degreeCelsius",
"temperatureAnnualMean_degreeCelsius","precipitationAnnualMean_millimetre",
"availableWaterCapacity_percentage","clay_percentage","nitrogen_percentage","pH",
"phosphorus_percentage","silt_percentage","sand_percentage","organicCarbon_percentage"
)
preds_z <- paste0(pred_vars, "_z")

cm <- suppressWarnings(cor(
plot_level_data[, preds_z, drop = FALSE],
use = "pairwise.complete.obs",
method = "spearman"
))

#draw the heatmap inline (NO windows()/quartz()/x11())

corrplot(
cm,
method = "color",
type = "upper",
diag = FALSE,
order = "hclust",
addCoef.col = "black",
number.cex = 0.55, # smaller numbers
tl.cex = 0.65, # smaller labels
tl.srt = 45, # angled labels
mar = c(0,0,1,0)
)

```

#Which predictors to remove? Here is where our knowledge of plant ecosystems comes into play
As clay and silt are highly positively correlated (0.93) and many plants are clay soil specialists. 

Temperature values are also (logically) correlated. As temp annual mean is correlated with both temp maxiumum and temp minimum - let's remove the average temp.

Nitrogen is also highlighted as correlated with organic Carbon (again logically) as plant available nitrogen (#need to check if this is how it is measured#) is a key plant limiting soil nutrient - let's cut organic Carbon.

Removing correlated predictors reduces our risk of an overfitted model + multicollinearity! 

#What does this mean?

With too many, highly redundant predictors, the model starts “fitting noise.” Coefficients become unstable, standard errors inflate, and signs can flip with small data changes. You might get impressive in-sample fit, but poor generalisation out-of-sample. Pruning correlated predictors (what you’re doing) reduces this risk and makes predictor effects more interpretable.

#But sand and clay are highly negatively correlated so we might want to check the VIF (variance inflation factor for these before we start modelling)

```{r}
#Option A — keep clay + sand (drop silt, mean temp, organic C)

preds_z <- c(
"minimumTemperature_degreeCelsius_z",
"maximumTemperature_degreeCelsius_z",
"precipitationAnnualMean_millimetre_z",
"availableWaterCapacity_percentage_z",
"clay_percentage_z",
"sand_percentage_z",
"nitrogen_percentage_z",
"pH_z",
"phosphorus_percentage_z"
)

preds_z <- intersect(preds_z, names(plot_level_data))

# build X matrix and keep complete rows

X <- as.matrix(plot_level_data[, preds_z, drop = FALSE])
X <- X[stats::complete.cases(X), , drop = FALSE]

# correlation matrix among predictors


R <- suppressWarnings(cor(X, use = "pairwise.complete.obs", method = "pearson"))

# VIF for standardised predictors is diag(solve(R)); fall back to pseudo-inverse if needed

invR <- tryCatch(solve(R), error = function(e) {
if (!requireNamespace("MASS", quietly = TRUE)) {
stop("Matrix is singular; install.packages('MASS') to use a pseudo-inverse.")
}
MASS::ginv(R)
})

vif_vals <- diag(invR)

# tidy table with flags

out <- data.frame(term = colnames(X), VIF = as.numeric(vif_vals))
out <- out[order(-out$VIF), ]
out$flag <- ifelse(out$VIF > 10, ">10 (problematic)",
ifelse(out$VIF > 5, ">5 (concerning)", ""))
print(out, row.names = FALSE)

```

# sand and clay are highly negatively correlated. So we will have to drop one. We could keep one or we can create a PCA - let's see how it goes.
## Note for Elisa: thinking not to do this - as it's so hard to interpret the model with the PCA included. But it was an interesting idea.Unless you have further thoghts let's drop this part.

The below code compresses the trio (sand, silt, clay)—which are highly redundant—into one axis that captures the main texture gradient.

Loadings tell the story: the numbers printed under pc1_loadings show how PC1 changes with each component.

e.g., if clay has a positive loading and sand a negative loading, higher PC1 means “more clay / less sand” (the sign can flip depending on the PCA; flipping the sign doesn’t change the meaning).

Use in models: plug Texture_PC1_z into your correlations/VIFs/GLMM instead of including sand/silt/clay separately. This removes the multicollinearity while keeping the texture signal.

#Creating a PCA

```{r}
# --- PCA on soil texture (%): create Texture_PC1 and a centred Texture_PC1_z ---

# 1) grab the three composition columns

tex_vars <- c("sand_percentage", "silt_percentage", "clay_percentage")

# silently keep only those that exist

tex_vars <- intersect(tex_vars, names(plot_level_data))

# quick safety: need all three to run a texture PCA

if (length(tex_vars) != 3) {
stop("Texture PCA needs sand_percentage, silt_percentage, and clay_percentage in plot_level_data.")
}

# 2) build a small matrix and keep complete rows

tex_mat <- as.data.frame(plot_level_data[tex_vars])

# coerce to numeric just in case

tex_mat[] <- lapply(tex_mat, function(x) suppressWarnings(as.numeric(x)))

ok <- stats::complete.cases(tex_mat)

# 3) PCA (standardised) on the complete cases

tex_pca <- prcomp(tex_mat[ok, ], scale. = TRUE)  # scale.=TRUE standardises variables

# 4) attach PC1 scores back to the full data (NA where any of the three were missing)

plot_level_data$Texture_PC1 <- NA_real_
plot_level_data$Texture_PC1[ok] <- tex_pca$x[, 1]

# 5) also create a z-centred PC1 for use in models

plot_level_data$Texture_PC1_z <- as.numeric(scale(plot_level_data$Texture_PC1))

# 6) handy readouts 

pc1_var_explained <- summary(tex_pca)$importance["Proportion of Variance", 1] * 100
pc1_loadings      <- tex_pca$rotation[, 1]

cat(sprintf("Texture PC1 explains ~%.1f%% of variance in (sand, silt, clay).\n", pc1_var_explained))
print(round(pc1_loadings, 3))

```

# Fantastic - We are ready to create our models

#We only have 1 visit per site across 785 sites so the GLMM random intercept will partition the data by site name: (1 | siteName)

GLMM (Generalized Linear Mixed Model) requires multiple observations per 
group to estimate random effect variance. Since each site has only one 
observation, we CANNOT use a random intercept like (1 | siteName).

Therefore, we use a standard GLM (Generalized Linear Model) with appropriate
distribution families to handle our response variables.

# Note: If we had multiple visits per site, a GLMM would be appropriate to 
 account for site-level correlation and unmeasured heterogeneity (e.g., soil 
 history, management, microclimate).

# --- Distribution Family Choices ---

# Our two response variables have different distributions:

# 1. SPECIES RICHNESS (count data: 0, 1, 2, 3, ...)
    - Natural families: Poisson or Negative Binomial
    - Ecological count data is typically OVERDISPERSED (variance > mean)
    - Poisson assumes variance = mean (too restrictive for ecological data)
    - NEGATIVE BINOMIAL allows variance > mean → better fit for species counts
    - We use: glm.nb() with family = Negative Binomial

# 2. SIMPSON DIVERSITY INDEX (SDI) (continuous proportion: 0 to 1)
    - Natural family: Beta distribution (via betareg or mgcv::betar())
    - Beta distribution models continuous (0,1) responses with flexible shapes
    - NOT appropriate: Binomial (for binary 0/1 or count/total outcomes)
    - NOT appropriate: Gaussian (assumes unbounded continuous data)
    - We use: Beta regression for SDI models

# Summary: Use Negative Binomial GLM for species richness counts,
#          and Beta GLM for Simpson Diversity Index proportions.

#First let's check for incomplete rows - as the models will not converge if there are missing data

```{r}
#check for missing data:
# Check how many complete cases we have
cat("Rows in plot_level_data:", nrow(plot_level_data), "\n")

# Check if SDI has NAs
cat("NAs in SDI:", sum(is.na(plot_level_data$SDI)), "\n")

# ============================================================================
# CHECK FOR INCOMPLETE CASES IN plot_level_data (NO REMOVAL)
# ============================================================================

# Check which columns we need for models
model_cols <- c(
  "species_richness",
  "SDI",
  "minimumTemperature_degreeCelsius_z",
  "maximumTemperature_degreeCelsius_z",
  "precipitationAnnualMean_millimetre_z",
  "availableWaterCapacity_percentage_z",
  "clay_percentage_z",
  "sand_percentage_z",
  "nitrogen_percentage_z",
  "pH_z",
  "phosphorus_percentage_z"
)

# Count NAs in each column
cat("=== CHECKING FOR MISSING DATA ===\n")
for (col in model_cols) {
  n_missing <- sum(is.na(plot_level_data[[col]]))
  cat(sprintf("%-45s: %d NAs\n", col, n_missing))
}

# Count complete cases
complete_rows <- complete.cases(plot_level_data[, model_cols])
cat("\nTotal rows:", nrow(plot_level_data), "\n")
cat("Complete cases:", sum(complete_rows), "\n")
cat("Incomplete cases:", sum(!complete_rows), "\n")

  
```

#we do have 14 incomplete rows - let's create a data subset to use in our models mod_df

```{r}
#create mod_df

needed <- c(
"species_richness",
"siteName",
"minimumTemperature_degreeCelsius_z",
"maximumTemperature_degreeCelsius_z",
"precipitationAnnualMean_millimetre_z",
"availableWaterCapacity_percentage_z",
"clay_percentage_z",
"nitrogen_percentage_z",
"pH_z",
"phosphorus_percentage_z"
)

#subset to those columns

mod_df <- plot_level_data[, needed, drop = FALSE]

#count rows before/after removing incomplete cases (this is where the ~14 rows go)

n_before <- nrow(mod_df)
mod_df <- mod_df[stats::complete.cases(mod_df), , drop = FALSE]
n_after <- nrow(mod_df)

#make sure site is a clean factor

mod_df$siteName <- droplevels(factor(mod_df$siteName))

#quick audit to the console

cat("Model DF rebuilt: ", n_before, "rows ->", n_after, "rows (dropped", n_before - n_after, "incomplete rows)\n")

#sanity checks

stopifnot(all(mod_df$species_richness >= 0, na.rm = TRUE))

```
```{r}

##let's start with species richness response and clay as a predictor

library(MASS)  # for negative binomial GLM

# Since each site has only 1 observation, use GLM (no random effects)

# Model with Negative Binomial family (handles overdispersion)
m_rich_clay <- glm.nb(
  species_richness ~
    minimumTemperature_degreeCelsius_z +
    maximumTemperature_degreeCelsius_z +
    precipitationAnnualMean_millimetre_z +
    availableWaterCapacity_percentage_z +
    clay_percentage_z +
    nitrogen_percentage_z +
    pH_z +
    phosphorus_percentage_z,
  data = mod_df
)

# Check the model
summary(m_rich_clay)

# Check for overdispersion
cat("Dispersion parameter (theta):", m_rich_clay$theta, "\n")

# Model diagnostics
par(mfrow = c(2, 2))
plot(m_rich_clay)

# Pearson residuals
pr_clay <- residuals(m_rich_clay, type = "pearson")
qqnorm(pr_clay)
qqline(pr_clay, col = "red")

# DHARMa diagnostics (works with GLMs too)
library(DHARMa)
sim_clay <- simulateResiduals(m_rich_clay, n = 1000)
plot(sim_clay)
testDispersion(sim_clay)
testZeroInflation(sim_clay)

```
#wow some interesting results for the clay model - what is your interpretation? We'd love to hear it.

##ECOLOGICAL STORY TO BE DELETED FROM WORKSHOP VERSION## (nb: ONLY FOR OUR ANSWERS VERSION)

#Key Ecological Story:
Your rangeland plant diversity is primarily driven by temperature extremes (the min/max temp effects are opposite and strongest) and soil chemistry (pH, clay). The "Goldilocks zone" appears to be: warmer minimum temperatures (avoiding cold stress), cooler maximum temperatures (avoiding heat stress), and more acidic, less clay-heavy soils.
The fact that precipitation wasn't significant is interesting - it suggests that in your dataset, temperature stress may be a stronger filter on diversity than water availability per se. This could make sense if most of your sites are already in water-limited rangelands where species are adapted to low rainfall, but temperature extremes still matter.

#Model Performance

Good model fit: The reduction from null deviance (1136.99) to residual deviance (796.15) shows your predictors explain a substantial portion of variation in species richness
Theta = 7.38: This indicates moderate overdispersion is present (if it were very high, like >50, you'd be close to Poisson; low values indicate more overdispersion). The negative binomial is definitely the right choice here.
DHARMa diagnostics look great: No dispersion issues (p=0.24) and no zero-inflation problems (p=1.0)

Significant Predictors (interpreting on the log scale, then I'll convert):
Minimum Temperature (positive effect, p < 0.001)*

Coefficient: +0.121
Interpretation: For each 1 SD increase in minimum temperature, species richness increases by ~12.9% (exp(0.121) = 1.129)
Ecological meaning: Warmer minimum temps (less cold stress) → more species

#Maximum Temperature (negative effect, p < 0.001)*

Coefficient: -0.188
Interpretation: For each 1 SD increase in maximum temperature, species richness decreases by ~17.1% (exp(-0.188) = 0.829)
Ecological meaning: Extreme heat stress → fewer species can tolerate conditions

#pH (negative effect, p < 0.001)*

Coefficient: -0.139
Interpretation: For each 1 SD increase in pH (more alkaline), species richness decreases by ~13.0% (exp(-0.139) = 0.870)
Ecological meaning: More acidic soils support higher diversity in these rangelands

#Clay percentage (negative effect, p = 0.006)

Coefficient: -0.044
Interpretation: For each 1 SD increase in clay, species richness decreases by ~4.3% (exp(-0.044) = 0.957)
Ecological meaning: Heavier clay soils → lower diversity (possibly due to drainage/aeration issues)

#Phosphorus (marginally significant, p = 0.054)

Coefficient: -0.049
Interpretation: Borderline negative effect; higher P → slightly fewer species
Ecological meaning: Could indicate competitive exclusion at high nutrient levels, but interpret cautiously

#Non-significant Predictors:

Precipitation (p = 0.68): Surprisingly not significant - possibly because temperature already captures moisture stress, or your sites don't span a wide enough precipitation gradient
Available Water Capacity (p = 0.11): Marginal but not quite significant
Nitrogen (p = 0.61): No effect
Sand/Silt (not in model - excluded due to collinearity with clay)


#let's check species richness against the same glm but with sand substiuted for clay

```{r}

# ============================================================================
# SPECIES RICHNESS ~ SAND MODEL
# ============================================================================

library(MASS)  # for glm.nb()
library(DHARMa)  # for diagnostics

######################run this extra bit if sand df has dissapeared
# Quick fix: just add the sand column to mod_df
mod_df$sand_percentage_z <- plot_level_data$sand_percentage_z[
  match(mod_df$siteName, plot_level_data$siteName)
]

# Verify it worked
cat("sand_percentage_z added to mod_df\n")
cat("NAs:", sum(is.na(mod_df$sand_percentage_z)), "\n")
cat("Range:", range(mod_df$sand_percentage_z, na.rm = TRUE), "\n")


# --- Fit Negative Binomial GLM with SAND as predictor ---
m_rich_sand <- glm.nb(
  species_richness ~
    minimumTemperature_degreeCelsius_z +
    maximumTemperature_degreeCelsius_z +
    precipitationAnnualMean_millimetre_z +
    availableWaterCapacity_percentage_z +
    sand_percentage_z +  # SAND instead of clay
    nitrogen_percentage_z +
    pH_z +
    phosphorus_percentage_z,
  data = mod_df
)

# --- Model summary ---
summary(m_rich_sand)

# --- Check dispersion parameter ---
cat("\nDispersion parameter (theta):", m_rich_sand$theta, "\n\n")

# --- Visual diagnostics ---
par(mfrow = c(2, 2))
plot(m_rich_sand)

# --- Pearson residuals QQ plot ---
pr_sand <- residuals(m_rich_sand, type = "pearson")
qqnorm(pr_sand, main = "QQ Plot: Species Richness ~ Sand Model")
qqline(pr_sand, col = "red")

# --- DHARMa diagnostics ---
sim_sand <- simulateResiduals(m_rich_sand, n = 1000)
plot(sim_sand, main = "DHARMa: Species Richness ~ Sand Model")
testDispersion(sim_sand)
testZeroInflation(sim_sand)


```
#Key insight: Sand and clay are telling the SAME story from opposite directions (they're negatively correlated, remember?). Sandier soils = less clay = better drainage = higher diversity. Both models are essentially equivalent in explanatory power.

#Ecological Story
Your rangeland diversity is maximized in:

Sandier soils (better drainage, aeration)
Warmer minimum temps (less cold stress)
Cooler maximum temps (less heat stress)
More acidic pH (lower pH values)
Lower phosphorus (possibly reduced competitive dominance)

Recommendation: Since sand and clay are just mirror images of the same gradient, I'd suggest picking one model (probably sand, since it has slightly lower AIC and phosphorus becomes significant). You could also mention in your write-up that "clay showed the opposite pattern (β = -0.044, p = 0.006), as expected given the negative correlation between sand and clay content."

#Pick the best model / model selection for species richness

```{r}

# Choose the sand model (slightly better AIC, phosphorus significant)
# Mention clay in text as supporting evidence

cat("=== MODEL SELECTION ===\n")
cat("Clay model AIC:", AIC(m_rich_clay), "\n")
cat("Sand model AIC:", AIC(m_rich_sand), "\n")
cat("Difference:", abs(AIC(m_rich_clay) - AIC(m_rich_sand)), "\n")
cat("\nAIC difference < 2 indicates models are essentially equivalent.\n")
cat("Choosing SAND model (marginally lower AIC, clearer P effect).\n")
cat("Clay shows inverse pattern as expected (r = -0.XX between sand/clay).\n")

```
#summary - delete later

```{r}
# ============================================================================
# FINAL MODEL SELECTION: SPECIES RICHNESS
# ============================================================================

cat("=== SPECIES RICHNESS MODEL SELECTION ===\n\n")

cat("Clay vs Sand comparison:\n")
cat("  Clay model AIC:", round(AIC(m_rich_clay), 2), "\n")
cat("  Sand model AIC:", round(AIC(m_rich_sand), 2), "\n")
cat("  ΔAIC =", round(abs(AIC(m_rich_clay) - AIC(m_rich_sand)), 2), "\n\n")

cat("Decision: SAND model selected (ΔAIC < 2 = models equivalent)\n")
cat("Sand and clay are negatively correlated and represent\n")
cat("opposite ends of the same soil texture gradient.\n\n")

# Calculate correlation between sand and clay
cor_sand_clay <- cor(mod_df$sand_percentage_z, 
                      plot_level_data$clay_percentage_z[
                        match(mod_df$siteName, plot_level_data$siteName)
                      ], 
                      use = "complete.obs")

cat("Correlation between sand and clay:", round(cor_sand_clay, 3), "\n\n")

cat("Final species richness model uses SAND percentage.\n")
cat("Key findings:\n")
cat("  • Sand (+): β = +0.047, p = 0.005 → sandier soils = higher richness\n")
cat("  • Min temp (+): β = +0.118, p < 0.001 → warmer minimums = higher richness\n")  
cat("  • Max temp (−): β = -0.185, p < 0.001 → cooler maximums = higher richness\n")
cat("  • pH (−): β = -0.138, p < 0.001 → more acidic soils = higher richness\n")
cat("  • Phosphorus (−): β = -0.050, p = 0.048 → lower P = higher richness\n\n")




```
#now the SDI (simpson's diversity index) models

```{r}

#1) SDI and clay

# ============================================================================
# SDI ~ CLAY MODEL
# ============================================================================

library(betareg)
library(statmod)

m_SDI_clay <- betareg(
  SDI ~
    minimumTemperature_degreeCelsius_z +
    maximumTemperature_degreeCelsius_z +
    precipitationAnnualMean_millimetre_z +
    availableWaterCapacity_percentage_z +
    clay_percentage_z +
    nitrogen_percentage_z +
    pH_z +
    phosphorus_percentage_z,
  data = plot_level_data
)

cat("=== SDI ~ CLAY MODEL ===\n")
summary(m_SDI_clay)

# Diagnostics
par(mfrow = c(2, 2))
plot(m_SDI_clay, which = 1:4)

# Pearson residuals QQ plot
par(mfrow = c(1, 1))
res_clay <- residuals(m_SDI_clay, type = "pearson")
qqnorm(res_clay, main = "Pearson Residuals: SDI ~ Clay")
qqline(res_clay, col = "red")

```
#SDI ~ clay predictors model interpretation:

SDI ~ Clay Model Interpretation
Model Fit Notes:

The warning about "extended-support beta regression" and diagnostic plots is normal - betareg automatically uses extended support when SDI values are near 0 or 1
"So-so" residuals are common with beta regression on ecological diversity indices - they're rarely perfect
As long as there's no major pattern in residuals vs fitted, you're okay

Significant Predictors:
Maximum Temperature (negative, p < 0.001)*

Coefficient: -0.227
Strong negative effect: higher max temps → lower diversity (evenness)
Same pattern as species richness!

Minimum Temperature (positive, p = 0.039)

Coefficient: +0.089
Warmer minimum temps → higher diversity
Weaker effect than on species richness, but same direction

Non-significant (but Clay is borderline):

Clay: p = 0.165 (not significant for SDI, unlike richness where it was p = 0.006)
pH: p = 0.261 (not significant for SDI, unlike richness where it was p < 0.001)
Phosphorus: p = 0.204 (not significant)
AWC: p = 0.073 (marginally)

Key Finding:
Temperature matters more for SDI than soil does! For species richness, both temperature AND soil (clay, pH) were important. For diversity/evenness (SDI), it's mostly about temperature extremes.


#let's move onto the model for SDI ~ sand + predictors

```{r}
# ============================================================================
# SDI ~ SAND MODEL
# ============================================================================

m_SDI_sand <- betareg(
  SDI ~
    minimumTemperature_degreeCelsius_z +
    maximumTemperature_degreeCelsius_z +
    precipitationAnnualMean_millimetre_z +
    availableWaterCapacity_percentage_z +
    sand_percentage_z +
    nitrogen_percentage_z +
    pH_z +
    phosphorus_percentage_z,
  data = plot_level_data
)

cat("\n=== SDI ~ SAND MODEL ===\n")
summary(m_SDI_sand)

# Diagnostics
par(mfrow = c(2, 2))
plot(m_SDI_sand, which = c(1, 4, 5))  # Skip unavailable plots

# Pearson residuals QQ plot
par(mfrow = c(1, 1))
res_sand <- residuals(m_SDI_sand, type = "pearson")
qqnorm(res_sand, main = "Pearson Residuals: SDI ~ Sand")
qqline(res_sand, col = "red")

```
#the Ecological Story:
For Species Richness: You need the right soil (sandy, acidic, low P) to support a diverse flora. Temperature matters, but soil is the primary filter.
For Diversity (SDI): Once species are present, temperature extremes determine whether communities are dominated by a few stress-tolerant species (low SDI) or have more even abundance distributions (high SDI). Soil matters less for competitive dynamics.

The AWC result is intriguing: Higher water-holding capacity might favor competitive dominants that can monopolize resources, reducing evenness.

#let's create a summary table of all 4 models
##Elisa I'm still playing around with this and the AI (LOL - not sure if we should keep or delete heehee let's see)

```{r}
# ============================================================================
# COMPREHENSIVE MODEL COMPARISON TABLE
# ============================================================================

# Extract coefficients and p-values from all models
library(dplyr)

# Function to extract model results
extract_results <- function(model, model_name) {
  if (inherits(model, "negbin")) {
    # For glm.nb models
    coef_summary <- summary(model)$coefficients
    results <- data.frame(
      Model = model_name,
      Predictor = rownames(coef_summary),
      Estimate = coef_summary[, "Estimate"],
      SE = coef_summary[, "Std. Error"],
      z_value = coef_summary[, "z value"],
      p_value = coef_summary[, "Pr(>|z|)"],
      sig = ifelse(coef_summary[, "Pr(>|z|)"] < 0.001, "***",
            ifelse(coef_summary[, "Pr(>|z|)"] < 0.01, "**",
            ifelse(coef_summary[, "Pr(>|z|)"] < 0.05, "*",
            ifelse(coef_summary[, "Pr(>|z|)"] < 0.1, ".", "")))),
      stringsAsFactors = FALSE
    )
  } else if (inherits(model, "betareg")) {
    # For betareg models (only mu coefficients)
    coef_summary <- summary(model)$coefficients$mean
    results <- data.frame(
      Model = model_name,
      Predictor = rownames(coef_summary),
      Estimate = coef_summary[, "Estimate"],
      SE = coef_summary[, "Std. Error"],
      z_value = coef_summary[, "z value"],
      p_value = coef_summary[, "Pr(>|z|)"],
      sig = ifelse(coef_summary[, "Pr(>|z|)"] < 0.001, "***",
            ifelse(coef_summary[, "Pr(>|z|)"] < 0.01, "**",
            ifelse(coef_summary[, "Pr(>|z|)"] < 0.05, "*",
            ifelse(coef_summary[, "Pr(>|z|)"] < 0.1, ".", "")))),
      stringsAsFactors = FALSE
    )
  }
  return(results)
}

# Extract results from all models
rich_clay_results <- extract_results(m_rich_clay, "Richness_Clay")
rich_sand_results <- extract_results(m_rich_sand, "Richness_Sand")
sdi_clay_results <- extract_results(m_SDI_clay, "SDI_Clay")
sdi_sand_results <- extract_results(m_SDI_sand, "SDI_Sand")

# Combine all results
all_results <- rbind(rich_clay_results, rich_sand_results, 
                     sdi_clay_results, sdi_sand_results)

# Remove intercept rows for cleaner comparison
all_results <- all_results[all_results$Predictor != "(Intercept)", ]

# Create wide format table for easy comparison
library(tidyr)

comparison_table <- all_results %>%
  select(Model, Predictor, Estimate, p_value, sig) %>%
  mutate(
    coef_sig = sprintf("%.3f %s", Estimate, sig),
    p_display = sprintf("%.4f", p_value)
  ) %>%
  select(Model, Predictor, coef_sig, p_display) %>%
  pivot_wider(
    names_from = Model,
    values_from = c(coef_sig, p_display),
    names_glue = "{Model}_{.value}"
  )

# Clean up predictor names for display
comparison_table$Predictor <- gsub("_z$", "", comparison_table$Predictor)
comparison_table$Predictor <- gsub("_", " ", comparison_table$Predictor)

# Print the table
cat("\n=== COMPREHENSIVE MODEL COMPARISON ===\n\n")
print(comparison_table, row.names = FALSE)

# Model fit statistics
cat("\n\n=== MODEL FIT STATISTICS ===\n")
fit_stats <- data.frame(
  Model = c("Richness ~ Clay", "Richness ~ Sand", "SDI ~ Clay", "SDI ~ Sand"),
  AIC = c(AIC(m_rich_clay), AIC(m_rich_sand), AIC(m_SDI_clay), AIC(m_SDI_sand)),
  N_obs = c(nobs(m_rich_clay), nobs(m_rich_sand), nobs(m_SDI_clay), nobs(m_SDI_sand))
)

if (inherits(m_rich_clay, "negbin")) {
  fit_stats$Theta <- c(m_rich_clay$theta, m_rich_sand$theta, NA, NA)
  fit_stats$LogLik <- c(logLik(m_rich_clay)[1], logLik(m_rich_sand)[1], 
                         logLik(m_SDI_clay)[1], logLik(m_SDI_sand)[1])
}

print(fit_stats, row.names = FALSE)


# ============================================================================
# SUMMARY TABLE: WHICH PREDICTORS ARE SIGNIFICANT?
# ============================================================================

cat("\n\n=== SIGNIFICANCE SUMMARY ===\n")
sig_summary <- all_results %>%
  mutate(
    is_sig = ifelse(p_value < 0.05, "✓", ""),
    is_marginal = ifelse(p_value >= 0.05 & p_value < 0.1, "(.)", ""),
    direction = ifelse(Estimate > 0, "+", "−")
  ) %>%
  select(Model, Predictor, direction, is_sig, is_marginal, p_value) %>%
  pivot_wider(
    names_from = Model,
    values_from = c(direction, is_sig, is_marginal),
    names_glue = "{Model}_{.value}"
  )

sig_summary$Predictor <- gsub("_z$", "", sig_summary$Predictor)
sig_summary$Predictor <- gsub("_", " ", sig_summary$Predictor)

# Create clean display
sig_display <- sig_summary %>%
  mutate(
    Richness_Clay = paste0(Richness_Clay_direction, " ", Richness_Clay_is_sig, Richness_Clay_is_marginal),
    Richness_Sand = paste0(Richness_Sand_direction, " ", Richness_Sand_is_sig, Richness_Sand_is_marginal),
    SDI_Clay = paste0(SDI_Clay_direction, " ", SDI_Clay_is_sig, SDI_Clay_is_marginal),
    SDI_Sand = paste0(SDI_Sand_direction, " ", SDI_Sand_is_sig, SDI_Sand_is_marginal)
  ) %>%
  select(Predictor, Richness_Clay, Richness_Sand, SDI_Clay, SDI_Sand)

print(sig_display, row.names = FALSE)
cat("\nLegend: ✓ = significant (p<0.05), (.) = marginal (p<0.1), + = positive, − = negative\n")


# ============================================================================
# KEY INSIGHTS
# ============================================================================

cat("\n\n=== KEY INSIGHTS ===\n\n")

cat("SPECIES RICHNESS (count of species):\n")
cat("  • Temperature effects: Min temp (+) ***, Max temp (−) ***\n")
cat("  • Soil texture: Clay (−) **, Sand (+) ** [opposite, as expected]\n")
cat("  • Soil chemistry: pH (−) ***\n")
cat("  • Nutrients: Phosphorus (−) * to **\n")
cat("  • AIC: Clay=", round(AIC(m_rich_clay), 1), ", Sand=", round(AIC(m_rich_sand), 1), 
    " (essentially equivalent)\n\n")

cat("SIMPSON DIVERSITY INDEX (evenness):\n")
cat("  • Temperature effects: Min temp (+) *, Max temp (−) ***\n")
cat("  • Soil properties: Mostly NON-significant\n")
cat("  • Water: AWC (−) * in sand model only\n")
cat("  • AIC: Clay=", round(AIC(m_SDI_clay), 1), ", Sand=", round(AIC(m_SDI_sand), 1), 
    " (essentially equivalent)\n\n")

cat("ECOLOGICAL INTERPRETATION:\n")
cat("  1. Soil properties (texture, pH, P) are strong filters for RICHNESS\n")
cat("     → determine which species can establish and persist\n\n")
cat("  2. Temperature extremes drive both RICHNESS and EVENNESS (SDI)\n")
cat("     → Min temp (+): less cold stress → more/better species\n")
cat("     → Max temp (−): less heat stress → more/better species\n\n")
cat("  3. For EVENNESS specifically, soil matters less than for richness\n")
cat("     → Temperature stress determines competitive dominance patterns\n")
cat("     → Communities under extreme heat have a few dominant species\n\n")

cat("MODEL SELECTION:\n")
cat("  • For Richness: Use SAND model (slightly lower AIC, P significant)\n")
cat("  • For SDI: Use SAND model (slightly lower AIC, AWC becomes significant)\n")
cat("  • Clay models tell same story (opposite sign for clay vs sand)\n\n")
```



