---
title: "TERN_workshop_ESA_2025"
author: "Elisa Girola and Eleanor Velasquez"
date: "2025-09-22"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading the packages
```{r, include=FALSE}
# Loading the packages ----------------------------------------------------
rm(list=ls()) # cleans the workspace

```

# FOLDER PATH
For each dataset we are providing both the code to access the files from a locally saved copy and from a remote copy. Remote access is typically slower, but ensures that you are always accessing the latest version of the data. Due to time constraints, during the workshop we will access locally saved data. The code that accesses the data remotely has been commented out. 

The following code specifies the folder path to access data saved locally. You can simply copy and paste your file path here, without the quotation marks. 
```{r}
# Select data folder
data_folder <- rstudioapi::selectDirectory()

```



# VEGETATION

<!-- ## Import the data from Ecoplots -->
<!-- ```{r} -->


<!-- ``` -->

## Import the data from a locally saved file
```{r}
# File name
file_name <- "Plant_occurrence.csv"

# Combine folder + file name
file_path <- file.path(data_folder, file_name)

# Read the CSV
vegetation <- read.csv(file_path, header = TRUE, fill = TRUE)

```

## For sites that have multiple site visits keep only the latest one
```{r}
library(dplyr)
vegetation <- vegetation %>%
  group_by(siteName) %>% # group rows for each site
  filter(siteVisitName == max(siteVisitName)) %>% # keep rows only for the latest visit name within that site
  ungroup() # remove grouping after filtering

```

# Create a table of plot level data
```{r}
# Keep one row per site
plot_level_data <- vegetation %>%
  group_by(siteName) %>%
  slice(1) %>%   # pick the first row per site (they all share the same plot details)
  ungroup()

```

## Create a map to show the locations of each plot
```{r}
library(dplyr)
library(ggplot2)
library(maps)

# Load Australia map
australia_map <- map_data("world", region = "Australia")

# Plot
ggplot() +
  geom_polygon(data = australia_map, aes(x = long, y = lat, group = group),
               fill = "gray90", color = "gray50") +
  geom_point(data = plot_level_data,
             aes(x = longitude_Degree, y = latitude_Degree),
             color = "darkgreen", size = 2) +
  coord_quickmap(xlim = c(110, 155), ylim = c(-45, -10)) +  # focus on Australia
  theme_minimal() +
  labs(title = "Rangeland vegetation plots",
       x = "Longitude", y = "Latitude")

rm(australia_map) # remove temporary tables

```

## Calculate species richness, species diversity and fractional cover
## NOTE: FOR ROWS THAT HAVE A SUBSPECIES, WE MAY BE ABLE TO EXTRACT THE SPECIES NAME FROM TAXON (TO DO). 
```{r}
# Calculate species richness 
library(dplyr)

richness_table <- vegetation %>%
  group_by(siteName) %>%
  summarise(
    species_richness = n_distinct(scientificName),
    .groups = "drop"
  )

# Calculate species diversity (Simpson Diversity Index)
library(dplyr)
library(vegan)

simpson_table <- vegetation %>%
  group_by(siteName) %>%
  summarise(
    SDI = diversity(table(scientificName), index = "simpson"),
    .groups = "drop"
  )

# Calculate "fractional cover", i.e. proportion of points in each plot that had vegetation

# Join data into main dataframe
plot_level_data <- plot_level_data %>%
  left_join(richness_table, by = "siteName") %>%
  left_join(simpson_table, by = "siteName")



# Remove temporary data
rm(richness_table, simpson_table) 
```


# TEMPERATURE
<!-- ## Import the data from Ecoplots -->
<!-- ```{r} -->


<!-- ``` -->

## Import the data from a locally saved file
```{r}
# File name
file_name <- "Temperature.csv"

# Combine folder + file name
file_path <- file.path(data_folder, file_name)

# Read the CSV
temperature <- read.csv(file_path, header = TRUE, fill = TRUE)

```

## Select data that match plot ID and site visit for vegetation
```{r}
library(dplyr)

# Join data into main dataframe
plot_level_data <- plot_level_data %>%
  left_join(
    temperature %>%
      select(siteName, siteVisitName, 
             maximumTemperature_degreeCelsius, 
             minimumTemperature_degreeCelsius, 
             temperatureAnnualMean_degreeCelsius),
    by = c("siteName", "siteVisitName")
  )


# Remove temporary data
rm(temperature) 
```

## Plot the vegetation and temperature data
```{r}
library(ggplot2)

# Species richness vs temperature
ggplot(plot_level_data, aes(x = temperatureAnnualMean_degreeCelsius, y = species_richness)) +
  geom_point(color = "blue", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "darkblue") +
  labs(
    x = "Annual Mean Temperature (°C)",
    y = "Species Richness") +
  theme_minimal()

# Simpson Diversity Index (SDI) vs temperature
ggplot(plot_level_data, aes(x = temperatureAnnualMean_degreeCelsius, y = SDI)) +
  geom_point(color = "green", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "darkgreen") +
  labs(
    x = "Annual Mean Temperature (°C)",
    y = "Simpson Diversity Index (SDI)") +
  theme_minimal()

```

# PRECIPITATION

# NOTE: CHECK THE DEFINITION OF RANGELAND. DOES IT INCLUDE PRECIPITATION???

<!-- ## Import the data from Ecoplots -->
<!-- ```{r} -->


<!-- ``` -->

## Import the data from a locally saved file
```{r}
# File name
file_name <- "Precipitation.csv"

# Combine folder + file name
file_path <- file.path(data_folder, file_name)

# Read the CSV
precipitation <- read.csv(file_path, header = TRUE, fill = TRUE)

```

## Select data that match plot ID and sitevisit for vegetation
```{r}
library(dplyr)

# Join data into main dataframe
plot_level_data <- plot_level_data %>%
  left_join(
    precipitation %>%
      select(siteName, siteVisitName, 
             precipitationAnnualMean_millimetre),
    by = c("siteName", "siteVisitName")
  )

# Remove temporary data
rm(precipitation) 

```

## Plot the vegetation and precipitation data
```{r}
library(ggplot2)

# Species richness vs precipitation
ggplot(plot_level_data, aes(x = precipitationAnnualMean_millimetre, y = species_richness)) +
  geom_point(color = "blue", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "darkblue") +
  labs(
    x = "Annual Mean Precipitation (mm)",
    y = "Species Richness") +
  theme_minimal()

# Simpson Diversity Index (SDI) vs precipitation
ggplot(plot_level_data, aes(x = precipitationAnnualMean_millimetre, y = SDI)) +
  geom_point(color = "green", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "darkgreen") +
  labs(
    x = "Annual Mean Precipitation (mm)",
    y = "Simpson Diversity Index (SDI)") +
  theme_minimal()

```

# SOIL - LANDSCAPE
## Access data directly from WCS
This is an example to access just one file directly. More layers can be added. Please note each layer requires several minutes to run. 
```{r}
# library(terra)
# 
# # WCS endpoint and layer
# base_url <- "https://geoserver.tern.org.au/geoserver/slga/CLY_v2/wcs"
# layer <- "slga__cly_ev_n_p_au_trn_n_20210902"
# 
# # Small buffer around each point (0.01 degrees ~1 km)
# buffer <- 0.01
# 
# # Function to get clay value at a single point
# get_clay_at_point <- function(lon, lat) {
#   xmin <- lon - buffer
#   xmax <- lon + buffer
#   ymin <- lat - buffer
#   ymax <- lat + buffer
#   
#   getcov <- paste0(
#     base_url,
#     "?service=WCS&version=2.0.1&request=GetCoverage",
#     "&coverageId=", layer,
#     "&format=image/geotiff",
#     "&subset=Long(", xmin, ",", xmax, ")",
#     "&subset=Lat(", ymin, ",", ymax, ")"
#   )
#   
#   # Download tiny raster
#   r <- rast(getcov)
#   
#   # Extract value at the exact point
#   extract(r, vect(data.frame(lon=lon, lat=lat), geom=c("lon","lat"), crs="EPSG:4326"))[,2]
# }
# 
# # Apply to all points
# plot_level_data$CLY_0_5cm <- mapply(get_clay_at_point,
#                                lon = plot_level_data$longitude_Degree,
#                                lat = plot_level_data$latitude_Degree)
# 
# head(plot_level_data)

```

## From locally saved files
File naming convention is as follows. Please note we are considering only the 0-5cm layer in this example. Models exists for deeper layers.
SOC = Organic Carbon, i.e.	mass fraction of carbon by weight in the < 2 mm soil material as determined by dry combustion at 900° C 
CLY =	Clay, i.e.< 2 um mass fraction of the < 2 mm soil material determined using the pipette method 	
SLT = Silt, i.e.	2-20 um mass fraction of the < 2 mm soil material determined using the pipette method 	
SND = Sand, i.e.	20 um - 2 mm mass fraction of the < 2 mm soil material determined using the pipette method 
PHC = pH - CaCl2, i.e.	pH of 1:5 soil/0.01M calcium chloride extract 
AWC = Available Water Capacity
NTO = Total Nitrogen 
PTO = Total Phosphorus
```{r}
library(terra)
library(dplyr)

# List all .tif files in the folder
tif_files <- list.files(data_folder, pattern = "\\.tif$", full.names = TRUE)

# Extract unique parameters from file names (first 3 letters)
parameters <- unique(substr(basename(tif_files), 1, 3))

# Create a SpatVector of plot locations
plot_points <- vect(plot_level_data, geom = c("longitude_Degree", "latitude_Degree"), crs = "EPSG:4326")

# Loop through each parameter
for (param in parameters) {
  
  # Select all files for this parameter
  param_files <- tif_files[substr(basename(tif_files), 1, 3) == param]
  
  # Load rasters
  rasters <- rast(param_files)
  
  # Extract values at plot locations
  values <- terra::extract(rasters, plot_points)[,-1]  # remove ID column
  
  # Add as new column to plot_level_data
  plot_level_data[[param]] <- values
}

library(dplyr)

# Rename columns
plot_level_data <- plot_level_data %>%
  rename(
    availableWaterCapacity_percentage = AWC,
    clay_percentage = CLY,
    nitrogen_percentage = NTO,
    pH = PHC,
    phosphorus_percentage = PTO,
    silt_percentage = SLT,
    sand_percentage = SND,
    organicCarbon_percentage = SOC
  )

# Remove temporary data
rm(rasters, plot_points) 
```

# create graphs for each variable
```{r}
library(ggplot2)
library(tidyr)
library(dplyr)

# Variables of interest
vars <- c(
  "availableWaterCapacity_percentage",
  "clay_percentage",
  "nitrogen_percentage",
  "pH",
  "phosphorus_percentage",
  "silt_percentage",
  "sand_percentage",
  "organicCarbon_percentage"
)

# Reshape your data from wide to long format
# You want to plot the same relationship (e.g., species richness vs each soil variable) in a single call to ggplot(), with multiple facets — one facet for each variable. To do that, ggplot2 needs a single column to hold all soil variable values, and another column to indicate which variable they belong to.
plot_long <- plot_level_data %>%
  pivot_longer(
    cols = all_of(vars),
    names_to = "variable",
    values_to = "value"
  )

# Create the plot titles
plot_long$variable <- factor(plot_long$variable,
  levels = vars,
  labels = c(
    "Available Water Capacity (%)",
    "Clay (%)",
    "Nitrogen (%)",
    "pH",
    "Phosphorus (%)",
    "Silt (%)",
    "Sand (%)",
    "Organic Carbon (%)"
  )
)

# Plot species_richness vs each variable
ggplot(plot_long, aes(x = value, y = species_richness)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "darkred", linewidth = 0.8) +
  facet_wrap(~ variable, scales = "free_x") +
  theme_bw() +
  labs(
    x = "Soil property value",
    y = "Species richness"
  )

# Plot SDI vs each variable
ggplot(plot_long, aes(x = value, y = SDI)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "darkred", linewidth = 0.8) +
  facet_wrap(~ variable, scales = "free_x") +
  theme_bw() +
  labs(
    x = "Soil property value",
    y = "Simpson Diversity Index"
  )
```


# SOIL - SURVEILLANCE
```{r}
# Grade (soil texture)

# Soil pH 

# match locations with vegetation data > is any plot missing? 

# compare values with Landscape > choose which one is better

# create histograms for each variable

```


# MODEL
```{r}

# correlation matrix
# 
# Plant richness ~ soil + Precip + AirTemp 
# 
# Plant diversity ~ soil + Precip + AirTemp 
# 
# Plant density ~ soil + Precip + AirTemp 

```
```
