---
title: "TERN_workshop_ESA_2025"
author: "Elisa Girola and Eleanor Velasquez"
date: "2025-09-22"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading the packages
```{r, include=FALSE}
# Loading the packages ----------------------------------------------------

rm(list=ls()) # cleans the workspace
```

# FOLDER PATH
For each dataset we are providing both the code to access the files from a locally saved copy and from a remote copy. Remote access is typically slower, but ensures that you are always accessing the latest version of the data. Due to time constraints, during the workshop we will access locally saved data. The code that accesses the data remotely has been commented out. 

The following code specifies the folder path to access data saved locally. You can simply copy and paste your file path here, without the quotation marks. 
```{r}
# Ask user for folder path at runtime
data_folder <- choose.dir()  # lets you select the folder interactively

```



# VEGETATION

<!-- ## Import the data from Ecoplots -->
<!-- ```{r} -->


<!-- ``` -->

## Import the data from a locally saved file
```{r}
# File name
file_name <- "Plant_occurrence.csv"

# Combine folder + file name
file_path <- file.path(data_folder, file_name)

# Read the CSV
vegetation <- read.csv(file_path, header = TRUE, fill = TRUE)

```

## For sites that have multiple site visits keep only the latest one
```{r}
library(dplyr)
vegetation <- vegetation %>%
  group_by(siteName) %>% # group rows for each site
  filter(siteVisitName == max(siteVisitName)) %>% # keep rows only for the latest visit name within that site
  ungroup() # remove grouping after filtering

```

## Create a map to show the locations of each plot
```{r}
library(dplyr)
library(ggplot2)
library(maps)

# Keep one row per site
vegetation_sites <- vegetation %>%
  group_by(siteName) %>%
  slice(1) %>%   # pick the first row per site (they all share same coords)
  ungroup()

# Load Australia map
australia_map <- map_data("world", region = "Australia")

# Plot
ggplot() +
  geom_polygon(data = australia_map, aes(x = long, y = lat, group = group),
               fill = "gray90", color = "gray50") +
  geom_point(data = vegetation_sites,
             aes(x = longitude_Degree, y = latitude_Degree),
             color = "darkgreen", size = 2) +
  coord_quickmap(xlim = c(110, 155), ylim = c(-45, -10)) +  # focus on Australia
  theme_minimal() +
  labs(title = "Vegetation Plot Locations in Australia",
       x = "Longitude", y = "Latitude")

```

## Calculate species richness, species diversity and fractional cover
## NOTE: WE MAY WANT TO REDOWNLOAD THE DATA INCLUDING THE PARAMETER TAXON (DONE NOW). FOR ROWS THAT HAVE A SUBSPECIES, WE MAY BE ABLE TO EXTRACT THE SPECIES NAME FROM TAXON (TO DO). 
```{r}

# Calculate species richness 
library(dplyr)

plot_table <- vegetation %>%
  group_by(siteName, latitude_Degree, longitude_Degree, 
           plotArea_SquareMetre, plotWidth_Metre, plotLength_Metre,
           plotShape, siteVisitName, siteVisitStartDate, siteVisitEndDate) %>%
  summarise(species_richness = n_distinct(scientificName), .groups = "drop") %>%
  select(siteName, latitude_Degree, longitude_Degree, 
         plotArea_SquareMetre, plotWidth_Metre, plotLength_Metre,
         plotShape, siteVisitName, siteVisitStartDate, siteVisitEndDate,
         species_richness)

# Calculate species diversity (Simpson Diversity Index)
library(dplyr)
library(vegan)

simpson_table <- vegetation %>%
  group_by(siteName) %>%
  summarise(
    SDI = diversity(table(scientificName), index = "simpson"),
    .groups = "drop"
  )

plot_table <- plot_table %>%
  left_join(simpson_table, by = "siteName") # saves to the plot_table

# Calculate "fractional cover", i.e. proportion of points in each plot that had vegetation




```


# TEMPERATURE
<!-- ## Import the data from Ecoplots -->
<!-- ```{r} -->


<!-- ``` -->

## Import the data from a locally saved file
```{r}
# File name
file_name <- "Temperature.csv"

# Combine folder + file name
file_path <- file.path(data_folder, file_name)

# Read the CSV
temperature <- read.csv(file_path, header = TRUE, fill = TRUE)

```

## Select data that match plot ID and sitevisit for vegetation
```{r}
library(dplyr)

# Join temperature data into plot_table
plot_table <- plot_table %>%
  left_join(
    temperature %>%
      select(siteName, siteVisitName, 
             maximumTemperature_degreeCelsius, 
             minimumTemperature_degreeCelsius, 
             temperatureAnnualMean_degreeCelsius),
    by = c("siteName", "siteVisitName")
  )

```

## Plot the vagetation and temperature data
```{r}
library(ggplot2)

# Species richness vs temperature
ggplot(plot_table, aes(x = temperatureAnnualMean_degreeCelsius, y = species_richness)) +
  geom_point(color = "blue", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "darkblue") +
  labs(
    x = "Annual Mean Temperature (°C)",
    y = "Species Richness") +
  theme_minimal()

# Simpson Diversity Index (SDI) vs temperature
ggplot(plot_table, aes(x = temperatureAnnualMean_degreeCelsius, y = SDI)) +
  geom_point(color = "green", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "darkgreen") +
  labs(
    x = "Annual Mean Temperature (°C)",
    y = "Simpson Diversity Index (SDI)") +
  theme_minimal()

```

# PRECIPITATION

# NOTE: CHECK THE DEFINITION OF RANGELAND. DOES IT INCLUDE PRECIPITATION???

<!-- ## Import the data from Ecoplots -->
<!-- ```{r} -->


<!-- ``` -->

## Import the data from a locally saved file
```{r}
# File name
file_name <- "Precipitation.csv"

# Combine folder + file name
file_path <- file.path(data_folder, file_name)

# Read the CSV
precipitation <- read.csv(file_path, header = TRUE, fill = TRUE)

```

## Select data that match plot ID and sitevisit for vegetation
```{r}
library(dplyr)

# Join temperature data into plot_table
plot_table <- plot_table %>%
  left_join(
    precipitation %>%
      select(siteName, siteVisitName, 
             precipitationAnnualMean_millimetre),
    by = c("siteName", "siteVisitName")
  )

```

## Plot the vegetation and precipitation data
```{r}
library(ggplot2)

# Species richness vs precipitation
ggplot(plot_table, aes(x = precipitationAnnualMean_millimetre, y = species_richness)) +
  geom_point(color = "blue", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "darkblue") +
  labs(
    x = "Annual Mean Precipitation (mm)",
    y = "Species Richness") +
  theme_minimal()

# Simpson Diversity Index (SDI) vs precipitation
ggplot(plot_table, aes(x = precipitationAnnualMean_millimetre, y = SDI)) +
  geom_point(color = "green", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "darkgreen") +
  labs(
    x = "Annual Mean Precipitation (mm)",
    y = "Simpson Diversity Index (SDI)") +
  theme_minimal()

```

# SOIL - LANDSCAPE
## Access data directly from WCS
This is an example to access just one file directly. More layers can be added. Please note each layer requires several minutes to run. 
```{r}
# library(terra)
# 
# # WCS endpoint and layer
# base_url <- "https://geoserver.tern.org.au/geoserver/slga/CLY_v2/wcs"
# layer <- "slga__cly_ev_n_p_au_trn_n_20210902"
# 
# # Small buffer around each point (0.01 degrees ~1 km)
# buffer <- 0.01
# 
# # Function to get clay value at a single point
# get_clay_at_point <- function(lon, lat) {
#   xmin <- lon - buffer
#   xmax <- lon + buffer
#   ymin <- lat - buffer
#   ymax <- lat + buffer
#   
#   getcov <- paste0(
#     base_url,
#     "?service=WCS&version=2.0.1&request=GetCoverage",
#     "&coverageId=", layer,
#     "&format=image/geotiff",
#     "&subset=Long(", xmin, ",", xmax, ")",
#     "&subset=Lat(", ymin, ",", ymax, ")"
#   )
#   
#   # Download tiny raster
#   r <- rast(getcov)
#   
#   # Extract value at the exact point
#   extract(r, vect(data.frame(lon=lon, lat=lat), geom=c("lon","lat"), crs="EPSG:4326"))[,2]
# }
# 
# # Apply to all points
# plot_table$CLY_0_5cm <- mapply(get_clay_at_point,
#                                lon = plot_table$longitude_Degree,
#                                lat = plot_table$latitude_Degree)
# 
# head(plot_table)

```

## From locally saved files

## NOTE: FOR https://portal.tern.org.au/metadata/TERN/482301c2-b9a1-4345-b142-815f9b37890a I.E. WATER CONTEN, WE NEED TO CLARIFY IF IT'S MM OR PERCENT
## FOR NITROGEN AND PHOSPHORUS WE NEED THE UNIT. BETTER CHECK THE UNITS FOR ALL VARIABLES AND STANDARDISE VARIABLE NAMES. 
```{r}
library(terra)
library(dplyr)

# List all .tif files in the folder
tif_files <- list.files(data_folder, pattern = "\\.tif$", full.names = TRUE)

# Extract unique parameters from file names (first 3 letters)
parameters <- unique(substr(basename(tif_files), 1, 3))

# Create a SpatVector of plot locations
plot_points <- vect(plot_table, geom = c("longitude_Degree", "latitude_Degree"), crs = "EPSG:4326")

# Loop through each parameter
for (param in parameters) {
  
  # Select all files for this parameter
  param_files <- tif_files[substr(basename(tif_files), 1, 3) == param]
  
  # Load rasters
  rasters <- rast(param_files)
  
  # Extract values at plot locations for all depths (or single raster if only one file)
  values <- extract(rasters, plot_points)[,-1]  # remove ID column
  
  # Add as new column to plot_table
  plot_table[[param]] <- values
}

library(dplyr)

plot_table <- plot_table %>%
  rename(
    availableWaterCapacity_millimetre = AWC,
    clay_percentage = CLY,
    nitrogen = NTO,
    pH = PHC,
    phosphorus = PTO,
    silt_percentage = SLT,
    sand_percentage = SND,
    organicCarbon = SOC
  )
```

# create graphs for each variable



# SOIL - SURVEILLANCE
```{r}
# Grade (soil texture)

# Soil pH 

# match locations with vegetation data > is any plot missing? 

# compare values with Landscape > choose which one is better

# create histograms for each variable

```


# MODEL
```{r}

# correlation matrix
# 
# Plant richness ~ soil + Precip + AirTemp 
# 
# Plant diversity ~ soil + Precip + AirTemp 
# 
# Plant density ~ soil + Precip + AirTemp 

##test commit

```
```
