---
title: "TERN_workshop_ESA_2025"
author: "Elisa"
date: "2025-09-22"
output: word_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading the packages
```{r, include=FALSE}
# Loading the packages ----------------------------------------------------

rm(list=ls()) # cleans the workspace
```



# For each dataset, we'll write code to access from a locally saved file and from a url (e.g. using the R generating code)

# VEGETATION

<!-- ## Import the data from Ecoplots -->
<!-- ```{r} -->


<!-- ``` -->

## Import the data from a locally saved file
```{r}
vegetation<-read.csv ("C:\\Users\\uqegirol\\Desktop\\Documents\\Conferences\\ESA\\ESA 2025\\Rangeland_PointIntAndVouchering_1a15b96498f911f089bb26ea16246ad1\\Rangeland_PointIntAndVouchering_1a15b96498f911f089bb26ea16246ad1\\data\\csv\\plant-occurrence_data_26ec5224d682b8d0eab19bd53afb0454.csv", header=TRUE, fill = TRUE) #  imports the file with the data

```

## For sites that have multiple site visits keep only the latest one
```{r}
library(dplyr)
vegetation <- vegetation %>%
  group_by(siteName) %>% # group rows for each site
  filter(siteVisitName == max(siteVisitName)) %>% # keep rows only for the latest visit name within that site
  ungroup() # remove grouping after filtering

```

## Create a map to show the locations of each plot
```{r}
library(dplyr)
library(ggplot2)
library(maps)

# Keep one row per site
vegetation_sites <- vegetation %>%
  group_by(siteName) %>%
  slice(1) %>%   # pick the first row per site (they all share same coords)
  ungroup()

# Load Australia map
australia_map <- map_data("world", region = "Australia")

# Plot
ggplot() +
  geom_polygon(data = australia_map, aes(x = long, y = lat, group = group),
               fill = "gray90", color = "gray50") +
  geom_point(data = vegetation_sites,
             aes(x = longitude_Degree, y = latitude_Degree),
             color = "darkgreen", size = 2) +
  coord_quickmap(xlim = c(110, 155), ylim = c(-45, -10)) +  # focus on Australia
  theme_minimal() +
  labs(title = "Vegetation Plot Locations in Australia",
       x = "Longitude", y = "Latitude")

```


## Calculate species richness, species diversity and fractional cover
## NOTE: WE MAY WANT TO REDOWNLOAD THE DATA INCLUDING THE PARAMETER TAXON. FOR ROWS THAT HAVE A SUBSPECIES, WE MAY BE ABLE TO EXTRACT THE SPECIES NAME FROM TAXON. 
```{r}

# Calculate species richness 
library(dplyr)

plot_table <- vegetation %>%
  group_by(siteName, latitude_Degree, longitude_Degree, 
           plotArea_SquareMetre, plotWidth_Metre, plotLength_Metre,
           plotShape, siteVisitName, siteVisitStartDate, siteVisitEndDate) %>%
  summarise(species_richness = n_distinct(scientificName), .groups = "drop") %>%
  select(siteName, latitude_Degree, longitude_Degree, 
         plotArea_SquareMetre, plotWidth_Metre, plotLength_Metre,
         plotShape, siteVisitName, siteVisitStartDate, siteVisitEndDate,
         species_richness)

# Calculate species diversity (Simpson Diversity Index)
library(dplyr)
library(vegan)

simpson_table <- vegetation %>%
  group_by(siteName) %>%
  summarise(
    SDI = diversity(table(scientificName), index = "simpson"),
    .groups = "drop"
  )

plot_table <- plot_table %>%
  left_join(simpson_table, by = "siteName") # saves to the plot_table

# Calculate "fractional cover", i.e. proportion of points in each plot that had vegetation




```


# TEMPERATURE
<!-- ## Import the data from Ecoplots -->
<!-- ```{r} -->


<!-- ``` -->

## Import the data from a locally saved file
```{r}
temperature<-read.csv ("C:\\Users\\uqegirol\\Desktop\\Documents\\Conferences\\ESA\\ESA 2025\\Temperature_15649682977911f09e3f26ea16246ad1\\Temperature_15649682977911f09e3f26ea16246ad1\\data\\csv\\weather_data_091e6bfe7d81bff49fc08251b743e2cc.csv", header=TRUE, fill = TRUE) #  imports the file with the data

```

## Select data that match plot ID and sitevisit for vegetation
```{r}
library(dplyr)

# Join temperature data into plot_table
plot_table <- plot_table %>%
  left_join(
    temperature %>%
      select(siteName, siteVisitName, 
             maximumTemperature_degreeCelsius, 
             minimumTemperature_degreeCelsius, 
             temperatureAnnualMean_degreeCelsius),
    by = c("siteName", "siteVisitName")
  )

```

## Plot the vagetation and temperature data
```{r}
library(ggplot2)

# Species richness vs temperature
ggplot(plot_table, aes(x = temperatureAnnualMean_degreeCelsius, y = species_richness)) +
  geom_point(color = "blue", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "darkblue") +
  labs(
    x = "Annual Mean Temperature (°C)",
    y = "Species Richness") +
  theme_minimal()

# Simpson Diversity Index (SDI) vs temperature
ggplot(plot_table, aes(x = temperatureAnnualMean_degreeCelsius, y = SDI)) +
  geom_point(color = "green", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "darkgreen") +
  labs(
    x = "Annual Mean Temperature (°C)",
    y = "Simpson Diversity Index (SDI)") +
  theme_minimal()

```

# PRECIPITATION

# NOTE: CHECK THE DEFINITION OF RANGELAND. DOES IT INCLUDE PRECIPITATION???

<!-- ## Import the data from Ecoplots -->
<!-- ```{r} -->


<!-- ``` -->

## Import the data from a locally saved file
```{r}
precipitation<-read.csv ("C:\\Users\\uqegirol\\Desktop\\Documents\\Conferences\\ESA\\ESA 2025\\Precipitation_dabcf7bfa7f911f09d2526ea16246ad1\\data\\csv\\weather_data_e1f82ca1c440bdc1a664bb0242f085c4.csv", header=TRUE, fill = TRUE) #  imports the file with the data

```

## Select data that match plot ID and sitevisit for vegetation
```{r}
library(dplyr)

# Join temperature data into plot_table
plot_table <- plot_table %>%
  left_join(
    precipitation %>%
      select(siteName, siteVisitName, 
             precipitationAnnualMean_millimetre),
    by = c("siteName", "siteVisitName")
  )

```

## Plot the vagetation and precipitation data
```{r}
library(ggplot2)

# Species richness vs precipitation
ggplot(plot_table, aes(x = precipitationAnnualMean_millimetre, y = species_richness)) +
  geom_point(color = "blue", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "darkblue") +
  labs(
    x = "Annual Mean Precipitation (mm)",
    y = "Species Richness") +
  theme_minimal()

# Simpson Diversity Index (SDI) vs precipitation
ggplot(plot_table, aes(x = precipitationAnnualMean_millimetre, y = SDI)) +
  geom_point(color = "green", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "darkgreen") +
  labs(
    x = "Annual Mean Precipitation (mm)",
    y = "Simpson Diversity Index (SDI)") +
  theme_minimal()


```

# SOIL - LANDSCAPE
## From locally saved files

```{r}
library(terra)
library(dplyr)

# Folder with soil .tif files
soil_folder <- "C:/Users/uqegirol/Desktop/Documents/Conferences/ESA/ESA 2025/Soil"

# List all .tif files
tif_files <- list.files(soil_folder, pattern = "\\.tif$", full.names = TRUE)

# Extract unique parameters from file names (first 3 letters)
parameters <- unique(substr(basename(tif_files), 1, 3))

# Create a SpatVector of plot locations
plot_points <- vect(plot_table, geom = c("longitude_Degree", "latitude_Degree"), crs = "EPSG:4326")

# Loop through each parameter
for (param in parameters) {
  
  # Select all files for this parameter
  param_files <- tif_files[substr(basename(tif_files), 1, 3) == param]
  
  # Load rasters
  rasters <- rast(param_files)
  
  # Extract values at plot locations for all depths
  values_all_depths <- extract(rasters, plot_points)[,-1]  # remove ID column
  
  # Calculate mean across depths
  values_mean <- rowMeans(values_all_depths, na.rm = TRUE)
  
  # Add as new column to plot_table
  plot_table[[param]] <- values_mean
}


# Percentage silt 

# Percentage clay 

# Percentage sand 
# 
# Nitrogen 
# 
# Phosphorus 
# 
# Soil pH 
# 
# Organic matter content 
# 
# Soil moisture 
```

# create graphs for each variable

```


# SOIL - SURVEILLANCE
```{r}
# Grade (soil texture)

# Soil pH 

# match locations with vegetation data > is any plot missing? 

# compare values with Landscape > choose which one is better

# create histograms for each variable

```


# MODEL
```{r}

# correlation matrix
# 
# Plant richness ~ soil + Precip + AirTemp 
# 
# Plant diversity ~ soil + Precip + AirTemp 
# 
# Plant density ~ soil + Precip + AirTemp 


```
```
